---
title: "CSF_Spectronaut_sample_swap data visualization"
subtitle: "Explicit rebuilt sample-swap data compared with the unswapped CSF reference"
format:
  html:
    toc: true
    code-fold: true
    fig-width: 9
    fig-height: 5
    embed-resources: true
  pdf:
    toc: true
    fig-width: 9
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

This vignette applies the same diagnostic structure as
`CSF_Spectronaut_data_vis.qmd`, but uses the explicit rebuilt
`CSF_Spectronaut_sample_swap` data as the after-swap state. The purpose is to
check whether this materialized sample-swap file preserves the same H0/H1
variance behaviour as the authors' in-script MSstats benchmark swap.

All quantitative panels in this document read the **Spectronaut report files
directly**. Intensities come from `FG.Quantity`, are transformed with `log2`,
deduplicated at `(run, protein, precursor)`, and then averaged to one
protein-level value per run before fold changes and SDs are computed. No
MSstats RDA object is used in this vignette.

```{r setup}
library(data.table)
library(dplyr)
library(ggplot2)
library(stringr)
library(tidyr)
library(patchwork)

theme_set(theme_minimal(base_size = 11))

QUANT_ROOT <- normalizePath(file.path(dirname(knitr::current_input(dir = TRUE)), ".."))
q <- function(...) file.path(QUANT_ROOT, ...)

annotation_before <- fread(q("CSF_Spectronaut/good_data/annotation.csv"))
annotation_after <- fread(q("CSF_Spectronaut_sample_swap/good_data/annotation.csv"))
swap_list <- fread(q("CSF_Spectronaut_sample_swap/CSF_protein_swap_list.csv"))
embedded_truth <- fread(q("CSF_Spectronaut_sample_swap/Report_sample_swap_ground_truth.tsv"))

make_pair_annotation <- function(annotation) {
  ann <- copy(annotation)
  setorder(ann, Order)
  ann[, Pair := NA_integer_]
  cond1_idx <- which(ann$Condition == "Condition1")
  cond2_idx <- which(ann$Condition == "Condition2")
  flip1 <- cond1_idx[seq(2, length(cond1_idx), by = 2)]
  flip2 <- cond2_idx[seq(2, length(cond2_idx), by = 2)]
  n_pairs <- min(length(flip1), length(flip2))
  ann[c(flip1[seq_len(n_pairs)], flip2[seq_len(n_pairs)]),
      Pair := rep(seq_len(n_pairs), 2)]
  ann
}

annotation_after <- make_pair_annotation(annotation_after)

load_report_protein_run <- function(report_path, annotation) {
  report <- fread(
    report_path,
    sep = "\t",
    select = c("R.FileName", "PG.ProteinGroups", "EG.PrecursorId", "FG.Quantity")
  )
  report <- report[
    nzchar(PG.ProteinGroups) & is.finite(FG.Quantity) & FG.Quantity > 0
  ]
  report <- unique(
    report,
    by = c("R.FileName", "PG.ProteinGroups", "EG.PrecursorId")
  )
  out <- report[
    ,
    .(LogIntensities = mean(log2(FG.Quantity), na.rm = TRUE)),
    by = .(Protein = PG.ProteinGroups, originalRUN = R.FileName)
  ]
  ann <- annotation[, .(originalRUN = R.FileName, GROUP = Condition, Order)]
  out <- merge(out, ann, by = "originalRUN", all.x = TRUE)
  out[, Protein := as.character(Protein)]
  out
}

prot_before <- load_report_protein_run(
  q("CSF_Spectronaut/good_data/Report.tsv"),
  annotation_before
)
prot_after <- load_report_protein_run(
  q("CSF_Spectronaut_sample_swap/good_data/Report.tsv"),
  annotation_after
)

true_pos <- swap_list[Label == "Positive", Protein]
true_neg <- swap_list[Label == "Negative", Protein]

prot_before[, state := "before swap reference"]
prot_after[, state := "after explicit sample-swap file"]

prot_all <- rbindlist(list(prot_before, prot_after), use.names = TRUE, fill = TRUE)
prot_all[, class := fifelse(Protein %in% true_pos, "Positive",
                            fifelse(Protein %in% true_neg, "Negative", NA_character_))]
prot_all <- prot_all[!is.na(class)]
prot_all[, state := factor(state, levels = c("before swap reference", "after explicit sample-swap file"))]
prot_all[, class := factor(class, levels = c("Negative", "Positive"))]
```

# Truth-file consistency

The explicit `Report.tsv` was generated from a ground-truth file. This check
compares that embedded generation truth with the current canonical
`CSF_protein_swap_list.csv` in the folder.

The table must show identical row counts and zero label mismatches. If it does
not, the explicit `Report.tsv` and the current truth file are out of sync and
the Positive/Negative figures below are not interpretable.

```{r truth_consistency}
truth_check <- merge(
  swap_list[, .(Protein, current_label = Label)],
  embedded_truth[, .(Protein, embedded_label = Label)],
  by = "Protein",
  all = TRUE
)

truth_summary <- data.table(
  check = c(
    "current canonical truth rows",
    "embedded generation truth rows",
    "shared proteins",
    "label mismatches on shared proteins",
    "current Positives labelled Negative in embedded truth",
    "current Negatives labelled Positive in embedded truth"
  ),
  value = c(
    nrow(swap_list),
    nrow(embedded_truth),
    sum(!is.na(truth_check$current_label) & !is.na(truth_check$embedded_label)),
    sum(truth_check$current_label != truth_check$embedded_label, na.rm = TRUE),
    sum(truth_check$current_label == "Positive" & truth_check$embedded_label == "Negative", na.rm = TRUE),
    sum(truth_check$current_label == "Negative" & truth_check$embedded_label == "Positive", na.rm = TRUE)
  )
)

knitr::kable(truth_summary, caption = "Consistency check between the current canonical truth and the truth used to generate the explicit sample-swap file.")
```

# Acquisition schedule

This figure shows the Good-run subset used in the explicit sample-swap file.
The points are runs from `CSF_Spectronaut_sample_swap/good_data/annotation.csv`.
Curves show the same alternating run pairs used to rewrite Negative-protein
rows in the generated Spectronaut report. The data plotted later are read from
`CSF_Spectronaut_sample_swap/good_data/Report.tsv`.

```{r schedule, fig.height=5, fig.width=12}
ann_plot <- copy(annotation_after)
ann_plot[, Sample := str_extract(R.FileName, "(Neat|1to[0-9]+)")]
ann_plot[, y := fifelse(Condition == "Condition1", 2, 1)]
pair_segs <- merge(
  ann_plot[!is.na(Pair) & Condition == "Condition1", .(Pair, x1 = Order, y1 = y)],
  ann_plot[!is.na(Pair) & Condition == "Condition2", .(Pair, x2 = Order, y2 = y)],
  by = "Pair"
)

ggplot(ann_plot, aes(x = Order, y = y)) +
  geom_curve(
    data = pair_segs,
    aes(x = x1, y = y1, xend = x2, yend = y2),
    curvature = 0.25,
    colour = "grey55",
    linewidth = 0.4,
    inherit.aes = FALSE
  ) +
  geom_point(aes(colour = Condition, shape = Sample), size = 4, stroke = 1.1) +
  geom_text(
    data = pair_segs,
    aes(x = (x1 + x2) / 2, y = (y1 + y2) / 2 + 0.06, label = Pair),
    size = 2.8,
    colour = "grey35",
    inherit.aes = FALSE
  ) +
  geom_text(aes(label = Sample, y = y - 0.18), size = 2.6, colour = "grey30", angle = 35, hjust = 1) +
  scale_colour_manual(values = c(Condition1 = "#1f77b4", Condition2 = "#d62728")) +
  scale_shape_manual(values = c(Neat = 16, "1to2" = 15)) +
  scale_x_continuous(breaks = seq(1, max(ann_plot$Order), by = 1), minor_breaks = NULL, expand = expansion(add = 1)) +
  scale_y_continuous(breaks = c(1, 2), labels = c("Condition2", "Condition1"), limits = c(0.4, 2.6)) +
  labs(title = "Run schedule and paired runs in CSF_Spectronaut_sample_swap/good_data", x = "Injection order", y = NULL, colour = NULL, shape = "Dilution") +
  theme(panel.grid.major.y = element_blank(), panel.grid.minor.x = element_blank(), legend.position = "top")
```

# Heatmap of the explicit sample-swap file

This heatmap uses direct report-derived protein/run intensities. For each run,
precursor-level `FG.Quantity` values are log2-transformed and averaged to one
value per protein. Rows are proteins from the current sample-swap truth list;
columns are Good runs. Values are row-z-scored within protein. Positive
proteins should keep the Condition1/Condition2 abundance pattern, while
Negative proteins should lose the mean contrast after the explicit sample
swap.

```{r heatmap, fig.height=12, fig.width=12}
good_runs <- annotation_after$R.FileName
good_order <- annotation_after[order(Condition, Order), .(Run = R.FileName, Condition, Order)]

prot_g <- prot_after[Protein %in% c(true_pos, true_neg) & originalRUN %in% good_runs]
prot_g[, class := fifelse(Protein %in% true_pos, "Positive",
                          fifelse(Protein %in% true_neg, "Negative", NA_character_))]
prot_g <- prot_g[!is.na(class)]
prot_g[, z := scale(LogIntensities)[, 1], by = Protein]
prot_g[, z := pmax(pmin(z, 2), -2)]

make_row_order <- function(dat) {
  m <- dcast(dat, Protein ~ originalRUN, value.var = "z")
  pids <- m$Protein
  mat <- as.matrix(m[, -1, with = FALSE])
  mat[is.na(mat)] <- 0
  pids[if (nrow(mat) > 1) hclust(dist(mat))$order else 1]
}

pos_ord <- make_row_order(prot_g[class == "Positive"])
neg_ord <- make_row_order(prot_g[class == "Negative"])
prot_g[, Protein := factor(Protein, levels = c(pos_ord, neg_ord))]
prot_g[, originalRUN := factor(originalRUN, levels = good_order$Run)]
prot_g[, class := factor(class, levels = c("Positive", "Negative"))]

col_strip <- good_order[, .(originalRUN = factor(Run, levels = good_order$Run), Condition)]
p_strip <- ggplot(col_strip, aes(x = originalRUN, y = 1, fill = Condition)) +
  geom_tile() +
  scale_fill_manual(values = c(Condition1 = "#1f77b4", Condition2 = "#d62728"), name = "Run condition") +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = NULL, y = NULL) +
  theme_void()

p_heat <- ggplot(prot_g, aes(x = originalRUN, y = Protein, fill = z)) +
  geom_tile() +
  facet_grid(class ~ ., scales = "free_y", space = "free_y", switch = "y") +
  scale_fill_gradient2(low = "#2ca02c", mid = "grey90", high = "#d62728", midpoint = 0, limits = c(-2, 2), name = "row z-score") +
  scale_x_discrete(expand = c(0, 0)) +
  labs(title = "Per-protein intensity in the explicit sample-swap file (Good runs)", x = NULL, y = NULL) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    strip.text.y.left = element_text(angle = 0, face = "bold"),
    legend.position = "right"
  )

p_strip / p_heat + plot_layout(heights = c(0.04, 1), guides = "collect")
```

# Per-protein effect size and within-condition variance

The following panels summarize the same direct report-derived protein/run
matrix. The before panel uses the unswapped `CSF_Spectronaut/good_data/Report.tsv`
as a reference. The after panel uses the regenerated explicit
`CSF_Spectronaut_sample_swap/good_data/Report.tsv`. The effect size is mean
log2 intensity in Condition1 minus Condition2. The pooled SD is computed from
within-condition SDs across runs.

```{r per_protein_stats}
stats <- prot_all[
  ,
  .(
    mean_log = mean(LogIntensities, na.rm = TRUE),
    sd_log = sd(LogIntensities, na.rm = TRUE),
    n = .N
  ),
  by = .(Protein, GROUP, state, class)
]

wide <- dcast(stats, Protein + state + class ~ GROUP, value.var = c("mean_log", "sd_log", "n"))
wide[, mean_diff := mean_log_Condition1 - mean_log_Condition2]
wide[, sd_pool := sqrt(((n_Condition1 - 1) * sd_log_Condition1^2 +
                          (n_Condition2 - 1) * sd_log_Condition2^2) /
                         pmax(n_Condition1 + n_Condition2 - 2, 1))]
wide[, mean_intensity := 0.5 * (mean_log_Condition1 + mean_log_Condition2)]

sd_summary <- wide[
  is.finite(sd_pool),
  .(
    proteins = .N,
    median_sd = round(median(sd_pool), 3),
    mean_sd = round(mean(sd_pool), 3),
    q25_sd = round(quantile(sd_pool, 0.25), 3),
    q75_sd = round(quantile(sd_pool, 0.75), 3)
  ),
  by = .(state, class)
]
```

```{r sd_summary}
knitr::kable(sd_summary, caption = "Pooled within-condition SD summary for the rebuilt explicit sample-swap data.")
```

## Effect size

This density plot checks the main sample-swap objective. Before swapping, both
classes carry the CSF dilution effect. After the explicit sample swap,
Negative proteins should be centered near zero, while Positive proteins should
remain near the original one-log2 dilution effect.

```{r effect_density, fig.height=4}
ggplot(wide[is.finite(mean_diff)], aes(x = mean_diff, fill = class)) +
  geom_density(alpha = 0.55) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey40") +
  facet_wrap(~ state) +
  scale_fill_manual(values = c(Positive = "#2ca02c", Negative = "#7f7f7f")) +
  labs(title = "Per-protein effect size", x = "log2 fold-change (Condition1 - Condition2)", y = "density", fill = NULL) +
  theme(legend.position = "top")
```

## Within-condition standard deviation

This density plot checks the variance consequence of the sample swap. After
the explicit swap, Negative proteins should have larger pooled SD than
Positive proteins, because each condition contains a mixture of Neat and 1:2
measurements for Negative proteins. Positive proteins are not swapped.

```{r sd_density, fig.height=4}
ggplot(wide[is.finite(sd_pool)], aes(x = sd_pool, fill = class)) +
  geom_density(alpha = 0.55, colour = "black", linewidth = 0.35) +
  facet_wrap(~ state) +
  scale_fill_manual(values = c(Positive = "#2ca02c", Negative = "#7f7f7f")) +
  labs(title = "Within-condition spread (pooled SD)", x = "sd(log2 intensity) inside each condition", y = "density", fill = NULL) +
  theme(legend.position = "top")
```

## Mean-variance relationship in the explicit sample-swap file

This plot shows the after-swap direct report-derived protein/run matrix. Each
point is one protein. The x-axis is mean log2 intensity and the y-axis is
pooled within-condition SD. Separation of Positive and Negative trends means
the explicit file reproduces the label-dependent variance structure.

```{r mean_var_trend, fig.height=4}
mv <- wide[is.finite(sd_pool) & state == "after explicit sample-swap file"]

ggplot(mv, aes(x = mean_intensity, y = sd_pool, colour = class)) +
  geom_point(alpha = 0.4, size = 0.7) +
  geom_smooth(se = FALSE, method = "loess") +
  scale_colour_manual(values = c(Positive = "#2ca02c", Negative = "#7f7f7f")) +
  labs(
    title = "Mean-variance relationship in CSF_Spectronaut_sample_swap",
    subtitle = "This checks whether the explicit rebuilt file preserves label-linked variance inflation",
    x = "mean log2 intensity",
    y = "pooled SD",
    colour = NULL
  ) +
  theme(legend.position = "top")
```
