---
title: "CSF_Spectronaut data visualization"
subtitle: "Authors' MSstats sample-swap benchmark implementation"
format:
  html:
    toc: true
    code-fold: true
    fig-width: 9
    fig-height: 5
    embed-resources: true
  pdf:
    toc: true
    fig-width: 9
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

This vignette reproduces the CSF Spectronaut sample-swap diagnostic from the
authors' benchmark implementation. The after-swap state is created in memory
with the same alternating run-label swap used by the MSstats benchmark code.

All quantitative panels in this document use the **MSstats protein-level
summary table** stored in
`CSF_Spectronaut/good_data/log2/swap/MSstats/MSstats_summarized.rda`. The
plotted intensity variable is `LogIntensities`, already on the log2 scale. This
is intentionally different from the explicit rebuilt sample-swap and
protein-swap vignettes, which read `Report.tsv` directly.

```{r setup}
library(data.table)
library(dplyr)
library(ggplot2)
library(stringr)
library(tidyr)
library(patchwork)

theme_set(theme_minimal(base_size = 11))

QUANT_ROOT <- normalizePath(file.path(dirname(knitr::current_input(dir = TRUE)), ".."))
q <- function(...) file.path(QUANT_ROOT, ...)

source(q("benchmark_experiments_functions.R"))

annotation <- fread(q("CSF_Spectronaut/annotation.csv"))
swap_list <- fread(q("CSF_Spectronaut/CSF_protein_swap_list.csv"))

e <- new.env()
load(q("CSF_Spectronaut/good_data/log2/swap/MSstats/MSstats_summarized.rda"), envir = e)
base_msstats_summarized <- e$base_msstats_summarized
prot <- as.data.table(base_msstats_summarized$ProteinLevelData)
prot[, Order := as.integer(str_split_i(originalRUN, "Seq", 2))]

make_pair_annotation <- function(annotation) {
  ann <- copy(annotation)
  setorder(ann, Order)
  ann[, Pair := NA_integer_]
  cond1_idx <- which(ann$Condition == "Condition1")
  cond2_idx <- which(ann$Condition == "Condition2")
  flip1 <- cond1_idx[seq(2, length(cond1_idx), by = 2)]
  flip2 <- cond2_idx[seq(2, length(cond2_idx), by = 2)]
  n_pairs <- min(length(flip1), length(flip2))
  ann[c(flip1[seq_len(n_pairs)], flip2[seq_len(n_pairs)]),
      Pair := rep(seq_len(n_pairs), 2)]
  ann
}

annotation <- make_pair_annotation(annotation)
```

# Acquisition schedule and swap runs

This figure shows the acquisition order and the run pairs used by the authors'
in-script sample swap. The points are MS runs from `CSF_Spectronaut/annotation.csv`.
The labels show the dilution sample in each run. Curved lines connect the
Condition1 and Condition2 runs whose condition labels are exchanged for
Negative proteins. Positive proteins are not swapped.

```{r schedule, fig.height=5, fig.width=12}
ann_plot <- annotation[Condition != "Blank"]
ann_plot[, Sample := str_extract(R.FileName, "(Neat|1to[0-9]+)")]
ann_plot[, y := fifelse(Condition == "Condition1", 2, 1)]
pair_segs <- merge(
  ann_plot[!is.na(Pair) & Condition == "Condition1", .(Pair, x1 = Order, y1 = y)],
  ann_plot[!is.na(Pair) & Condition == "Condition2", .(Pair, x2 = Order, y2 = y)],
  by = "Pair"
)

region <- ann_plot[
  ,
  .(xmin = min(Order) - 0.5, xmax = max(Order) + 0.5),
  by = Label
]

ggplot(ann_plot, aes(x = Order, y = y)) +
  geom_rect(
    data = region,
    inherit.aes = FALSE,
    aes(xmin = xmin, xmax = xmax, ymin = 0.4, ymax = 2.6),
    fill = "grey80",
    alpha = 0.10
  ) +
  geom_curve(
    data = pair_segs,
    aes(x = x1, y = y1, xend = x2, yend = y2),
    curvature = 0.25,
    colour = "grey55",
    linewidth = 0.4,
    inherit.aes = FALSE
  ) +
  geom_point(aes(colour = Condition, shape = Sample), size = 4, stroke = 1.1) +
  geom_text(
    data = pair_segs,
    aes(x = (x1 + x2) / 2, y = (y1 + y2) / 2 + 0.06, label = Pair),
    size = 2.8,
    colour = "grey35",
    inherit.aes = FALSE
  ) +
  geom_text(aes(label = Sample, y = y - 0.18), size = 2.6, colour = "grey30", angle = 35, hjust = 1) +
  scale_colour_manual(values = c(Condition1 = "#1f77b4", Condition2 = "#d62728")) +
  scale_shape_manual(values = c(Neat = 16, "1to2" = 15, "1to4" = 18, "1to8" = 17, "1to16" = 25, "1to32" = 19, "1to64" = 22)) +
  scale_x_continuous(breaks = seq(1, max(ann_plot$Order), by = 1), minor_breaks = NULL, expand = expansion(add = 1)) +
  scale_y_continuous(breaks = c(1, 2), labels = c("Condition2", "Condition1"), limits = c(0.4, 2.6)) +
  labs(title = "Run schedule and paired runs", x = "Injection order", y = NULL, colour = NULL, shape = "Dilution") +
  theme(panel.grid.major.y = element_blank(), panel.grid.minor.x = element_blank(), legend.position = "top")
```

# Apply the benchmark swap

The swap is applied to the MSstats protein-level table, not to a rewritten
Spectronaut report file. For the "before swap" panel, the protein-level
`LogIntensities` are used as loaded. For the "after swap" panel,
`swap_condition_labels()` relabels every second run within each condition for
Negative proteins while leaving Positive proteins unchanged.

```{r apply_swap}
true_pos <- swap_list[Label == "Positive", Protein]
true_neg <- swap_list[Label == "Negative", Protein]

prot_before <- copy(prot)
prot_before[, state := "before swap"]

prot_after <- swap_condition_labels(prot, true_pos, true_neg) |> as.data.table()
prot_after[, state := "after swap"]

prot_all <- rbindlist(list(prot_before, prot_after), use.names = TRUE, fill = TRUE)
prot_all[, Protein := as.character(Protein)]
prot_all[, class := fifelse(Protein %in% true_pos, "Positive",
                            fifelse(Protein %in% true_neg, "Negative", NA_character_))]
prot_all <- prot_all[!is.na(class)]
prot_all[, state := factor(state, levels = c("before swap", "after swap"))]
prot_all[, class := factor(class, levels = c("Negative", "Positive"))]
```

# Heatmap after the swap

This heatmap uses the same MSstats protein-level `LogIntensities`. Rows are
proteins from the benchmark truth list and columns are Good runs only. Values
are row-z-scored within protein, so the colors show relative high/low
abundance across runs rather than absolute intensity. The top facet contains
Positive proteins; the bottom facet contains Negative proteins after the
sample-swap operation.

```{r heatmap, fig.height=12, fig.width=12}
good_runs <- annotation[Label == "Good", R.FileName]
good_order <- annotation[Label == "Good"][order(Condition, Order), .(Run = R.FileName, Condition, Order)]

run_table <- unique(prot[, .(originalRUN, GROUP, Order)])
setorder(run_table, Order)
cond1 <- run_table[GROUP == "Condition1", originalRUN]
cond2 <- run_table[GROUP == "Condition2", originalRUN]
flip1 <- cond1[seq(2, length(cond1), by = 2)]
flip2 <- cond2[seq(2, length(cond2), by = 2)]
partner_map <- c(setNames(flip2, flip1), setNames(flip1, flip2))

prot_runswap <- copy(prot)
prot_runswap[, originalRUN := as.character(originalRUN)]
prot_runswap[, Protein := as.character(Protein)]
prot_runswap[Protein %in% true_neg & originalRUN %in% names(partner_map),
             originalRUN := unname(partner_map[originalRUN])]
prot_runswap[, class := fifelse(Protein %in% true_pos, "Positive",
                                fifelse(Protein %in% true_neg, "Negative", NA_character_))]

prot_g <- prot_runswap[!is.na(class) & originalRUN %in% good_runs]
prot_g[, z := scale(LogIntensities)[, 1], by = Protein]
prot_g[, z := pmax(pmin(z, 2), -2)]

make_row_order <- function(dat) {
  m <- dcast(dat, Protein ~ originalRUN, value.var = "z")
  pids <- m$Protein
  mat <- as.matrix(m[, -1, with = FALSE])
  mat[is.na(mat)] <- 0
  pids[if (nrow(mat) > 1) hclust(dist(mat))$order else 1]
}

pos_ord <- make_row_order(prot_g[class == "Positive"])
neg_ord <- make_row_order(prot_g[class == "Negative"])
prot_g[, Protein := factor(Protein, levels = c(pos_ord, neg_ord))]
prot_g[, originalRUN := factor(originalRUN, levels = good_order$Run)]
prot_g[, class := factor(class, levels = c("Positive", "Negative"))]

col_strip <- good_order[, .(originalRUN = factor(Run, levels = good_order$Run), Condition)]
p_strip <- ggplot(col_strip, aes(x = originalRUN, y = 1, fill = Condition)) +
  geom_tile() +
  scale_fill_manual(values = c(Condition1 = "#1f77b4", Condition2 = "#d62728"), name = "Run condition") +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = NULL, y = NULL) +
  theme_void()

p_heat <- ggplot(prot_g, aes(x = originalRUN, y = Protein, fill = z)) +
  geom_tile() +
  facet_grid(class ~ ., scales = "free_y", space = "free_y", switch = "y") +
  scale_fill_gradient2(low = "#2ca02c", mid = "grey90", high = "#d62728", midpoint = 0, limits = c(-2, 2), name = "row z-score") +
  scale_x_discrete(expand = c(0, 0)) +
  labs(title = "Per-protein intensity after the benchmark swap (Good runs only)", x = NULL, y = NULL) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    strip.text.y.left = element_text(angle = 0, face = "bold"),
    legend.position = "right"
  )

p_strip / p_heat + plot_layout(heights = c(0.04, 1), guides = "collect")
```

# Per-protein effect size and within-condition variance

The following panels summarize the MSstats protein-level data by protein,
condition, class, and swap state. For each protein, the effect size is the mean
`LogIntensities` in Condition1 minus the mean in Condition2. The pooled SD is
computed from the within-condition SDs of `LogIntensities`.

```{r per_protein_stats}
prot_good <- prot_all[originalRUN %in% good_runs]
stats <- prot_good[
  ,
  .(
    mean_log = mean(LogIntensities, na.rm = TRUE),
    sd_log = sd(LogIntensities, na.rm = TRUE),
    n = .N
  ),
  by = .(Protein, GROUP, state, class)
]

wide <- dcast(stats, Protein + state + class ~ GROUP, value.var = c("mean_log", "sd_log", "n"))
wide[, mean_diff := mean_log_Condition1 - mean_log_Condition2]
wide[, sd_pool := sqrt(((n_Condition1 - 1) * sd_log_Condition1^2 +
                          (n_Condition2 - 1) * sd_log_Condition2^2) /
                         pmax(n_Condition1 + n_Condition2 - 2, 1))]
wide[, mean_intensity := 0.5 * (mean_log_Condition1 + mean_log_Condition2)]
```

## Effect size

This density plot shows the per-protein log2 fold change. Before the swap,
both Positive and Negative proteins carry the CSF dilution effect. After the
authors' sample swap, Negative proteins are centered near zero while Positive
proteins retain the dilution effect.

```{r effect_density, fig.height=4}
ggplot(wide[is.finite(mean_diff)], aes(x = mean_diff, fill = class)) +
  geom_density(alpha = 0.55) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey40") +
  facet_wrap(~ state) +
  scale_fill_manual(values = c(Positive = "#2ca02c", Negative = "#7f7f7f")) +
  labs(title = "Per-protein effect size", x = "log2 fold-change (Condition1 - Condition2)", y = "density", fill = NULL) +
  theme(legend.position = "top")
```

## Within-condition standard deviation

This density plot shows the pooled within-condition SD of the same MSstats
protein-level `LogIntensities`. After the sample swap, Negative proteins have
higher within-condition spread because each condition contains a mixture of
Neat and 1:2 measurements for those proteins. Positive proteins are not
swapped, so their spread remains lower.

```{r sd_density, fig.height=4}
ggplot(wide[is.finite(sd_pool)], aes(x = sd_pool, fill = class)) +
  geom_density(alpha = 0.55, colour = "black", linewidth = 0.35) +
  facet_wrap(~ state) +
  scale_fill_manual(values = c(Positive = "#2ca02c", Negative = "#7f7f7f")) +
  labs(title = "Within-condition spread (pooled SD)", x = "sd(log2 intensity) inside each condition", y = "density", fill = NULL) +
  theme(legend.position = "top")
```

## Mean-variance relationship after swap

This scatter plot shows each protein after the swap. The x-axis is mean log2
intensity across the two conditions and the y-axis is pooled within-condition
SD. The separate trends for Positive and Negative proteins are the variance
moderation problem: methods using one shared mean-variance trend are asked to
moderate across two label-dependent variance regimes.

```{r mean_var_trend, fig.height=4}
mv <- wide[is.finite(sd_pool) & state == "after swap"]

ggplot(mv, aes(x = mean_intensity, y = sd_pool, colour = class)) +
  geom_point(alpha = 0.4, size = 0.7) +
  geom_smooth(se = FALSE, method = "loess") +
  scale_colour_manual(values = c(Positive = "#2ca02c", Negative = "#7f7f7f")) +
  labs(
    title = "Mean-variance relationship after swap",
    subtitle = "The benchmark creates separate H1 and H0 variance behaviour",
    x = "mean log2 intensity",
    y = "pooled SD",
    colour = NULL
  ) +
  theme(legend.position = "top")
```
