---
title: "CSF_Spectronaut_protein_swap data visualization"
subtitle: "Explicit protein-swap benchmark data"
format:
  html:
    toc: true
    code-fold: true
    fig-width: 9
    fig-height: 5
    embed-resources: true
  pdf:
    toc: true
    fig-width: 9
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

This vignette uses the same structure as the CSF sample-swap data-visualization
vignettes, but it describes the explicit protein-swap benchmark. In this
dataset, the Positive class is created by swapping matched proteins rather than
by swapping sample labels. The `good_data` subset uses neat samples only, so
the Positive/Negative labels come from the protein-pair construction and not
from dilution-level sample swapping.

All quantitative panels in this document read
`CSF_Spectronaut_protein_swap/good_data/Report.tsv` directly. Intensities come
from `FG.Quantity`, are transformed with `log2`, deduplicated at
`(run, protein, precursor)`, and then averaged to one protein-level value per
run before fold changes and SDs are computed. No MSstats RDA object is used in
this vignette.

```{r setup}
library(data.table)
library(dplyr)
library(ggplot2)
library(stringr)
library(tidyr)
library(patchwork)

theme_set(theme_minimal(base_size = 11))

QUANT_ROOT <- normalizePath(file.path(dirname(knitr::current_input(dir = TRUE)), ".."))
q <- function(...) file.path(QUANT_ROOT, ...)

annotation <- fread(q("CSF_Spectronaut_protein_swap/good_data/annotation.csv"))
swap_list <- fread(q("CSF_Spectronaut_protein_swap/CSF_protein_swap_list.csv"))
pair_truth <- fread(q("CSF_Spectronaut_protein_swap/Report_swap_ground_truth.tsv"))

load_report_protein_run <- function(report_path, annotation) {
  report <- fread(
    report_path,
    sep = "\t",
    select = c("R.FileName", "PG.ProteinGroups", "EG.PrecursorId", "FG.Quantity")
  )
  report <- report[
    nzchar(PG.ProteinGroups) & is.finite(FG.Quantity) & FG.Quantity > 0
  ]
  report <- unique(
    report,
    by = c("R.FileName", "PG.ProteinGroups", "EG.PrecursorId")
  )
  out <- report[
    ,
    .(LogIntensities = mean(log2(FG.Quantity), na.rm = TRUE)),
    by = .(Protein = PG.ProteinGroups, originalRUN = R.FileName)
  ]
  ann <- annotation[, .(originalRUN = R.FileName, GROUP = Condition, Order)]
  out <- merge(out, ann, by = "originalRUN", all.x = TRUE)
  out[, Protein := as.character(Protein)]
  out
}

prot <- load_report_protein_run(
  q("CSF_Spectronaut_protein_swap/good_data/Report.tsv"),
  annotation
)

true_pos <- swap_list[Label == "Positive", Protein]
true_neg <- swap_list[Label == "Negative", Protein]

prot[, class := fifelse(Protein %in% true_pos, "Positive",
                        fifelse(Protein %in% true_neg, "Negative", NA_character_))]
prot <- prot[!is.na(class)]
prot[, class := factor(class, levels = c("Negative", "Positive"))]
```

# Truth-file consistency

The canonical protein-swap truth list is `CSF_protein_swap_list.csv`. The
pair-construction table is `Report_swap_ground_truth.tsv`; it contains one row
per matched protein pair and records the expected pair-level log2 fold change.
The table below checks that the canonical Positive proteins are exactly the
proteins represented by the pair table.

```{r truth_consistency}
pair_positive <- unique(c(pair_truth$PG_high, pair_truth$PG_low))
truth_summary <- data.table(
  check = c(
    "canonical truth rows",
    "canonical Positive proteins",
    "canonical Negative proteins",
    "protein-pair rows",
    "unique pair-table Positive proteins",
    "canonical Positives missing from pair table",
    "pair-table proteins missing from canonical Positives"
  ),
  value = c(
    nrow(swap_list),
    sum(swap_list$Label == "Positive"),
    sum(swap_list$Label == "Negative"),
    nrow(pair_truth),
    length(pair_positive),
    length(setdiff(true_pos, pair_positive)),
    length(setdiff(pair_positive, true_pos))
  )
)

knitr::kable(truth_summary, caption = "Consistency check between the canonical protein-swap truth list and the protein-pair table used to generate the explicit report.")
```

# Acquisition schedule

This figure shows the Good-run subset used for the protein-swap benchmark.
The points are runs from
`CSF_Spectronaut_protein_swap/good_data/annotation.csv`. In this dataset the
data modification is at the protein level, not at the run-label level, so no
sample-pair curves are shown. The plotted data later are read from
`CSF_Spectronaut_protein_swap/good_data/Report.tsv`.

```{r schedule, fig.height=5, fig.width=12}
ann_plot <- copy(annotation)
ann_plot[, Sample := str_extract(R.FileName, "(Neat|1to[0-9]+)")]
ann_plot[, y := fifelse(Condition == "Condition1", 2, 1)]

ggplot(ann_plot, aes(x = Order, y = y)) +
  geom_point(aes(colour = Condition, shape = Sample), size = 4, stroke = 1.1) +
  geom_text(aes(label = Sample, y = y - 0.18), size = 2.6, colour = "grey30", angle = 35, hjust = 1) +
  scale_colour_manual(values = c(Condition1 = "#1f77b4", Condition2 = "#d62728")) +
  scale_shape_manual(values = c(Neat = 16, "1to2" = 15)) +
  scale_x_continuous(breaks = sort(unique(ann_plot$Order)), minor_breaks = NULL, expand = expansion(add = 1)) +
  scale_y_continuous(breaks = c(1, 2), labels = c("Condition2", "Condition1"), limits = c(0.4, 2.6)) +
  labs(title = "Run schedule in CSF_Spectronaut_protein_swap/good_data", x = "Injection order", y = NULL, colour = NULL, shape = "Dilution") +
  theme(panel.grid.major.y = element_blank(), panel.grid.minor.x = element_blank(), legend.position = "top")
```

# Heatmap of the explicit protein-swap file

This heatmap uses direct report-derived protein/run intensities. For each run,
precursor-level `FG.Quantity` values are log2-transformed and averaged to one
value per protein. Rows are proteins from the current protein-swap truth list;
columns are Good runs. Values are row-z-scored within protein. Positive
proteins are the protein-pair members and should carry the constructed signal;
Negative proteins should remain near the null.

```{r heatmap, fig.height=12, fig.width=12}
good_runs <- annotation$R.FileName
good_order <- annotation[order(Condition, Order), .(Run = R.FileName, Condition, Order)]

prot_g <- prot[Protein %in% c(true_pos, true_neg) & originalRUN %in% good_runs]
prot_g[, z := scale(LogIntensities)[, 1], by = Protein]
prot_g[, z := pmax(pmin(z, 2), -2)]

make_row_order <- function(dat) {
  m <- dcast(dat, Protein ~ originalRUN, value.var = "z")
  pids <- m$Protein
  mat <- as.matrix(m[, -1, with = FALSE])
  mat[is.na(mat)] <- 0
  pids[if (nrow(mat) > 1) hclust(dist(mat))$order else 1]
}

pos_ord <- make_row_order(prot_g[class == "Positive"])
neg_ord <- make_row_order(prot_g[class == "Negative"])
prot_g[, Protein := factor(Protein, levels = c(pos_ord, neg_ord))]
prot_g[, originalRUN := factor(originalRUN, levels = good_order$Run)]
prot_g[, class := factor(class, levels = c("Positive", "Negative"))]

col_strip <- good_order[, .(originalRUN = factor(Run, levels = good_order$Run), Condition)]
p_strip <- ggplot(col_strip, aes(x = originalRUN, y = 1, fill = Condition)) +
  geom_tile() +
  scale_fill_manual(values = c(Condition1 = "#1f77b4", Condition2 = "#d62728"), name = "Run condition") +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = NULL, y = NULL) +
  theme_void()

p_heat <- ggplot(prot_g, aes(x = originalRUN, y = Protein, fill = z)) +
  geom_tile() +
  facet_grid(class ~ ., scales = "free_y", space = "free_y", switch = "y") +
  scale_fill_gradient2(low = "#2ca02c", mid = "grey90", high = "#d62728", midpoint = 0, limits = c(-2, 2), name = "row z-score") +
  scale_x_discrete(expand = c(0, 0)) +
  labs(title = "Per-protein intensity in the explicit protein-swap file (Good runs)", x = NULL, y = NULL) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    strip.text.y.left = element_text(angle = 0, face = "bold"),
    legend.position = "right"
  )

p_strip / p_heat + plot_layout(heights = c(0.04, 1), guides = "collect")
```

# Per-protein effect size and within-condition variance

The following panels summarize the same direct report-derived protein/run
matrix. The effect size is mean log2 intensity in Condition1 minus Condition2.
The pooled SD is computed from within-condition SDs across runs. Unlike the
sample-swap benchmark, this design should not make the Negative class a
within-condition mixture of Neat and 1:2 samples; `good_data` contains neat
samples only.

```{r per_protein_stats}
stats <- prot[
  ,
  .(
    mean_log = mean(LogIntensities, na.rm = TRUE),
    sd_log = sd(LogIntensities, na.rm = TRUE),
    n = .N
  ),
  by = .(Protein, GROUP, class)
]

wide <- dcast(stats, Protein + class ~ GROUP, value.var = c("mean_log", "sd_log", "n"))
wide[, mean_diff := mean_log_Condition1 - mean_log_Condition2]
wide[, sd_pool := sqrt(((n_Condition1 - 1) * sd_log_Condition1^2 +
                          (n_Condition2 - 1) * sd_log_Condition2^2) /
                         pmax(n_Condition1 + n_Condition2 - 2, 1))]
wide[, mean_intensity := 0.5 * (mean_log_Condition1 + mean_log_Condition2)]

sd_summary <- wide[
  is.finite(sd_pool),
  .(
    proteins = .N,
    median_fc = round(median(mean_diff), 3),
    mean_fc = round(mean(mean_diff), 3),
    median_sd = round(median(sd_pool), 3),
    mean_sd = round(mean(sd_pool), 3),
    q25_sd = round(quantile(sd_pool, 0.25), 3),
    q75_sd = round(quantile(sd_pool, 0.75), 3)
  ),
  by = class
]
```

```{r sd_summary}
knitr::kable(sd_summary, caption = "Effect-size and pooled within-condition SD summary for the explicit protein-swap good_data report.")
```

## Effect size

This density plot checks the protein-swap signal. Negative proteins should be
centered closer to zero. Positive proteins are matched protein-pair members, so
their effect-size distribution reflects the constructed protein-pair signal
rather than a sample dilution effect.

```{r effect_density, fig.height=4}
ggplot(wide[is.finite(mean_diff)], aes(x = mean_diff, fill = class)) +
  geom_density(alpha = 0.55) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey40") +
  scale_fill_manual(values = c(Positive = "#2ca02c", Negative = "#7f7f7f")) +
  labs(title = "Per-protein effect size", x = "log2 fold-change (Condition1 - Condition2)", y = "density", fill = NULL) +
  theme(legend.position = "top")
```

## Within-condition standard deviation

This density plot checks whether the protein-swap benchmark avoids the
sample-swap variance artifact. Because `good_data` contains neat samples only,
Negative proteins should not acquire the same within-condition dilution-mixture
variance that appears in the sample-swap benchmark.

The pattern shown here is the expected direction for regulated proteins in this
protein-swap construction. The Positive class contains matched protein-pair
members and therefore combines the constructed abundance difference with
protein-specific measurement heterogeneity. The same direction is also
biologically plausible: in real treatment-control experiments, regulated
proteins are the proteins that respond to treatment, and that response can add
between-sample heterogeneity on top of technical variation. It is therefore
not suspicious that the Positive density is broader and shifted to higher
pooled SD than the Negative density. Regulated proteins are not expected to be
systematically less variable than all null proteins.

This is the opposite of the problematic sample-swap pattern. In the authors'
sample-swap benchmark, the null/Negative class becomes a within-condition
mixture of Neat and 1:2 measurements, so the Negative density is artificially
shifted to higher SD than the Positive density. That label-dependent variance
inflation is a property of the benchmark construction, not a property one
would generally expect from biological regulation.

```{r sd_density, fig.height=4}
ggplot(wide[is.finite(sd_pool)], aes(x = sd_pool, fill = class)) +
  geom_density(alpha = 0.55, colour = "black", linewidth = 0.35) +
  scale_fill_manual(values = c(Positive = "#2ca02c", Negative = "#7f7f7f")) +
  labs(title = "Within-condition spread (pooled SD)", x = "sd(log2 intensity) inside each condition", y = "density", fill = NULL) +
  theme(legend.position = "top")
```

## Mean-variance relationship in the explicit protein-swap file

This plot shows the direct report-derived protein/run matrix. Each point is
one protein. The x-axis is mean log2 intensity and the y-axis is pooled
within-condition SD. It is the protein-swap analogue of the sample-swap
mean-variance diagnostic.

Here the Positive trend being above the Negative trend is compatible with the
protein-swap design: positives are the constructed regulated proteins. The
important distinction is that the null class is not forced into a higher
variance regime by mixing dilution levels inside each condition. This is why
the protein-swap diagnostic is a better control for variance-moderated methods
than the sample-swap diagnostic where H0 proteins are made artificially more
variable than H1 proteins.

```{r mean_var_trend, fig.height=4}
mv <- wide[is.finite(sd_pool)]

ggplot(mv, aes(x = mean_intensity, y = sd_pool, colour = class)) +
  geom_point(alpha = 0.4, size = 0.7) +
  geom_smooth(se = FALSE, method = "loess") +
  scale_colour_manual(values = c(Positive = "#2ca02c", Negative = "#7f7f7f")) +
  labs(
    title = "Mean-variance relationship in CSF_Spectronaut_protein_swap",
    subtitle = "Direct report-derived protein/run summaries from good_data",
    x = "mean log2 intensity",
    y = "pooled SD",
    colour = NULL
  ) +
  theme(legend.position = "top")
```
