---
title: "A review of the MSstatsPlus benchmark (RMSV000000701.3)"
author: "Witold Wolski"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    fig-width: 7
    fig-height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.align = "center")
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(knitr)
})

# Resolve quant/ root from this qmd's location, then expose all paths via q()
# so each chunk works regardless of knitr's per-chunk setwd reset.
QUANT_ROOT <- normalizePath(
  file.path(dirname(knitr::current_input(dir = TRUE)), ".."),
  mustWork = FALSE
)
if (!dir.exists(QUANT_ROOT)) QUANT_ROOT <- normalizePath("..", mustWork = FALSE)
q <- function(...) file.path(QUANT_ROOT, ...)

source(q("R/normalize.R"))
source(q("R/ground_truth.R"))
source(q("R/figures.R"))

## Canonical truth sets per folder
truth_csf <- read.csv(q("CSF_Spectronaut/CSF_protein_swap_list.csv"))
truth_pswap <- read.csv(q("CSF_Spectronaut_protein_swap/CSF_protein_swap_list.csv"))
truth_sswap <- read.csv(q("CSF_Spectronaut_sample_swap/CSF_protein_swap_list.csv"))
truth_mix <- truth_mix_of_proteome(q("Mix_of_Proteome/idmapping.tsv"))

tpr_ppv <- function(model_csv, truth, thr = 0.05) {
  if (!file.exists(model_csv)) return(data.frame(N=NA, TPR=NA, PPV=NA))
  m <- read.csv(model_csv)
  if ("Label" %in% colnames(m)) m$Label <- NULL
  m <- merge(m, truth, by = "Protein")
  fc_col <- if ("log2FC" %in% colnames(m)) "log2FC" else "logFC"
  finite_fc <- is.finite(m[[fc_col]])
  hit <- !is.na(m$pvalue) & m$pvalue < thr & finite_fc
  tpr <- sum(hit & m$Label == "Positive", na.rm = TRUE) /
         sum(finite_fc & m$Label == "Positive", na.rm = TRUE)
  ppv <- if (sum(hit) > 0) {
    1 - sum(hit & m$Label == "Negative", na.rm = TRUE) / sum(hit)
  } else NA_real_
  data.frame(N = nrow(m), TPR = round(tpr, 3), PPV = round(ppv, 3))
}

tpr_ppv_fdr <- function(model_csv, truth, thr = 0.05) {
  if (!file.exists(model_csv)) return(data.frame(N=NA, TPR_FDR=NA, PPV_FDR=NA))
  m <- read.csv(model_csv)
  if ("Label" %in% colnames(m)) m$Label <- NULL
  m <- merge(m, truth, by = "Protein")
  fc_col <- if ("log2FC" %in% colnames(m)) "log2FC" else "logFC"
  finite_fc <- is.finite(m[[fc_col]])
  hit <- !is.na(m$adj.pvalue) & m$adj.pvalue < thr & finite_fc
  tpr <- sum(hit & m$Label == "Positive", na.rm = TRUE) /
         sum(finite_fc & m$Label == "Positive", na.rm = TRUE)
  ppv <- if (sum(hit) > 0) {
    1 - sum(hit & m$Label == "Negative", na.rm = TRUE) / sum(hit)
  } else NA_real_
  data.frame(N = nrow(m), TPR_FDR = round(tpr, 3), PPV_FDR = round(ppv, 3))
}

# Map canonical package label -> on-disk model filename.
model_filename <- list(
  "MSstats"      = "MSstats_model.csv",
  "MSstats+"     = "MSstats+_model.csv",
  "MaxLFQ_limma" = "limma_model.csv",
  "DEqMS"        = "deqms_model.csv",
  "msqrob2"      = "msqrob2_model.csv",
  "prolfqua"     = "prolfqua_model.csv",
  "limpa"        = "limpa_model.csv"
)
packages <- names(model_filename)

table1 <- function(folder, dataset, normalization, truth, thr_p = 0.05, thr_fdr = 0.05) {
  rows <- lapply(packages, function(pkg) {
    f <- q(folder, dataset, normalization, "swap", pkg, model_filename[[pkg]])
    p <- tpr_ppv(f, truth, thr_p)
    fdr <- tpr_ppv_fdr(f, truth, thr_fdr)
    data.frame(Package = pkg, N = p$N, TPR = p$TPR, PPV = p$PPV,
               TPR_FDR = fdr$TPR_FDR, PPV_FDR = fdr$PPV_FDR)
  })
  do.call(rbind, rows)
}

load_models <- function(folder, dataset, normalization, truth) {
  out <- lapply(packages, function(pkg) {
    f <- q(folder, dataset, normalization, "swap", pkg, model_filename[[pkg]])
    if (!file.exists(f)) return(NULL)
    m <- read.csv(f); m$Package <- pkg
    if ("Label" %in% colnames(m)) m$Label <- NULL
    if (!"logFC" %in% colnames(m) && "log2FC" %in% colnames(m)) m$logFC <- m$log2FC
    m <- merge(m, truth, by = "Protein")
    m[, c("Protein", "logFC", "pvalue", "adj.pvalue", "Package", "Label")]
  })
  do.call(rbind, Filter(Negate(is.null), out))
}
```

# Introduction {#sec-intro}

This document reviews the benchmark presented in the MSstatsPlus manuscript
(RMSV000000701.3 deposit) [REF: bioRxiv preprint, Kohler & Vitek 2025] and its
resubmission revision. The review reproduces the benchmark numerically,
inspects the dataset construction more closely than the manuscript does, and
flags methodological choices that affect the conclusions.

## About the authors and MSstats

Olga Vitek's group has been one of the foundational contributors to
quantitative mass-spectrometry-based proteomics statistics for more than a
decade. The MSstats package
[REF: Choi et al., Bioinformatics 2014; doi:10.1093/bioinformatics/btu305] was
first published in 2014 and has remained one of the most widely cited
differential-expression tools in the field. The group's broader contributions
span run-level summarization, label-free quantification, post-translational
modification analysis, and the
[MSstatsConvert](https://github.com/Vitek-Lab/MSstatsConvert) ecosystem that
standardizes input across DDA, DIA and PRM acquisition modes.

## MSstats as a reference in competing publications

Because MSstats is the de-facto reference, almost every subsequent
differential-expression package for proteomics is benchmarked against it.
Three notable examples reach a similar conclusion — that alternative tools
outperform MSstats on representative benchmarks:

- **msqrob2** [REF: Goeminne et al., MCP 2020] — peptide-level robust ridge
  regression with empirical Bayes variance moderation. Outperformed MSstats on
  the CPTAC benchmark and ROC curves on the human–yeast–E. coli mix.
- **limpa** [REF: Smyth et al., 2024] — a precursor-level method using a
  data-dependent precision-weighted model with empirical Bayes shrinkage.
- **prolfqua** [REF: Wolski et al., JPR 2023; doi:10.1021/acs.jproteome.2c00441]
  — protein-level linear models on medpolish-aggregated protein intensities;
  reproducible benchmarking framework on Spectronaut and DIA-NN outputs.

## What MSstatsPlus claims

MSstatsPlus extends MSstats with two ingredients: (1) **anomaly-aware feature
weighting** — per-precursor scores that down-weight features showing temporal
or shape anomalies during dataProcess; (2) **linear summarization** in place of
TMP for the protein roll-up. The manuscript claims that the combination
recovers a sensitivity advantage over the alternative tools (msqrob2, limpa,
prolfqua, limma, DEqMS) on the CSF dilution benchmark, and matches them on the
species-mix benchmark.

This review confirms the numerical Table 1 for CSF, then shows that the result
is **partly a property of the benchmark design**, not method quality.

## A protein-universe gap to flag up front

The manuscript's sample-swap benchmark is built on a fixed list of
**1,820 "quantified proteins"** (10 % Positive / 90 % Negative; Methods
§4.1, Supplementary Table 12 of the second submission). Our parallel
protein-swap benchmark, run on the same Spectronaut output, covers
**3,041 proteins** — a strict superset (1,820 ⊂ 3,041; 1,221 extra
proteins are kept by the protein-swap pipeline and discarded by the
authors' sample-swap list). The manuscript does not document what filter
narrows the universe to 1,820.

This 1,221-protein gap matters for two reasons. First, the discarded
proteins are systematically the lower-coverage / higher-missingness
ones — exactly the class on which moderated methods most differ from
MSstats+. Excluding them from the sample-swap benchmark shifts the
comparison toward the "clean" subset where MSstats+ has the most to gain
from variance asymmetry. Second, the gap motivates this review's choice
to **rebuild the sample-swap benchmark on the full 3,041-protein
universe** (still 10 / 90 by design, so ~304 Positives / ~2,737
Negatives) so that the sample-swap and protein-swap benchmarks are
directly comparable. The authors' 1,820-list results are still
reproduced verbatim in §3 for manuscript replicability.

# The manuscript's Table 1: bioRxiv vs resubmission

The original bioRxiv version [REF: bioRxiv 2025/04] reported Table 1 at a
**raw p-value threshold of 0.05** across seven methods. The resubmission
revised Table 1 substantially: the threshold became **adj.pvalue < 0.05**
(FDR-controlled), three methods were re-tabulated under a new variance
formulation, and the conclusion shifted from "MSstats+ alone wins" to
"MSstats+ wins jointly with msqrob2 / prolfqua / limpa". We reproduce both
thresholding modes side-by-side in @sec-table1-csf to make the dependency
visible.

# Benchmark datasets {#sec-benchmarks}

The MSstatsPlus paper benchmarks on two datasets, both built on the same
construction principle:

1. A **CSF dilution series** acquired on Spectronaut — the focus of this
   review. Detailed below.
2. A **K562 dilution series** acquired on both Spectronaut and DIA-NN.
   Structurally identical to the CSF series: a dilution series of a single
   biological sample (HEK-293-K562 cell lysate instead of human CSF), with
   the same run/label swap design used to manufacture the null class. K562
   results are reported in the manuscript but are **out of scope for this
   review** — the design-level critique that applies to CSF
   (swap-induced σ²+0.25 variance asymmetry; see @sec-sample-swap) carries
   over to K562 by identical construction.

## CSF dilution acquisition

The CSF dataset is a dilution series of human cerebrospinal fluid acquired on
a single LC-MS platform. Each MS run is a different dilution of the same
underlying CSF pool; runs are *interleaved* in injection order, so any
temporal drift spreads evenly across conditions.

```{r samples-per-dilution}
ann <- read.csv(q("CSF_Spectronaut/CSF_annotation.csv"))
raw_dil <- sub("-DD", "", sapply(strsplit(ann$R.FileName, "_"), `[`, 3))
ann$Dilution <- dplyr::case_when(
  grepl("^NeatCSF",  raw_dil) ~ "Neat",
  grepl("^1to2CSF",  raw_dil) ~ "1/2",
  grepl("^1to4CSF",  raw_dil) ~ "1/4",
  grepl("^1to8CSF",  raw_dil) ~ "1/8",
  grepl("^1to16CSF", raw_dil) ~ "1/16",
  grepl("^1to32CSF", raw_dil) ~ "1/32",
  grepl("^1to64CSF", raw_dil) ~ "1/64",
  grepl("^Blank",    raw_dil) ~ "Blank",
  TRUE                         ~ raw_dil
)
# Numeric dilution factor for sorting (Blank last).
ann$.dil_num <- dplyr::case_when(
  ann$Dilution == "Neat"  ~ 1,
  ann$Dilution == "1/2"   ~ 2,
  ann$Dilution == "1/4"   ~ 4,
  ann$Dilution == "1/8"   ~ 8,
  ann$Dilution == "1/16"  ~ 16,
  ann$Dilution == "1/32"  ~ 32,
  ann$Dilution == "1/64"  ~ 64,
  ann$Dilution == "Blank" ~ 99,
  TRUE                    ~ NA_real_
)
tab <- ann %>%
  group_by(.dil_num, Dilution, Condition, Label) %>%
  summarise(n_runs = n(), .groups = "drop") %>%
  arrange(.dil_num, desc(n_runs)) %>%
  select(Dilution, Condition, Label, n_runs)
kable(tab, caption = "Samples per dilution × experimental condition × good/bad annotation flag.")
```

Two facts the table makes explicit:

1. The canonical contrast `Condition1` vs `Condition2` is **Neat (15 runs) vs
   1:2 (14 runs)** — both flagged `Good`. Higher dilutions (1:4 through 1:64)
   provide a handful of `Bad` runs split between conditions (5 in each).
2. Both `Good` conditions are technical dilutions of the same biological
   material, differing by exactly 2× (log2 ≈ 1). **In the raw data every
   protein is differentially expressed** — there is no intrinsic null class.

## The need for a swap and the manufactured null

TPR and PPV require both positives and negatives. The raw dilution series
provides only positives. The manuscript's benchmark **manufactures the
missing null class** by swapping condition labels for 90 % of proteins,
chosen from a fixed list of 1,820 (the `CSF_protein_swap_list.csv`): 182
Positives (kept as-is, retain their natural 2× difference) and 1,638 Negatives
(condition labels swapped for half the runs, so the pooled means cancel out).

```{r swap-counts, results='asis'}
swap <- read.csv(q("CSF_Spectronaut/CSF_protein_swap_list.csv"))
cat(sprintf("Positives (true DE): **%d** &nbsp; Negatives (true null): **%d**\n",
            sum(swap$Label == "Positive"), sum(swap$Label == "Negative")))
```

The swap operation is described in detail in
[`CSF_Spectronaut/CSF_swap_design.md`](../CSF_Spectronaut/CSF_swap_design.md).
Briefly: sort runs by injection order; within each condition take every-second
run; pair them index-for-index across conditions; for every Negative-protein
measurement in a paired run, replace the run identifier with the partner's.
This produces 10 run pairs (7 "Good" pairs between Neat and 1:2; 3 "Bad" pairs
between higher dilutions).

The schematic is in
[`CSF_Spectronaut/CSF_swap_visualization.qmd`](../CSF_Spectronaut/CSF_swap_visualization.qmd)
(Figure 1 of that document). The §0.1 path layout we use throughout this
review is `<csffolder>/<dataset>/<normalization>/swap/<package>/`.

### Swap-schedule figure (lifted from `CSF_swap_visualization.qmd`)

```{r swap-schedule, fig.height=4.2, fig.width=12, fig.cap="CSF run schedule. Each dot is one MS run, plotted at its injection order. Top row = Condition1 (Neat-mostly), bottom row = Condition2 (1:2-mostly). Curves connect the two runs whose Negative-protein measurements get exchanged; numbers label pair IDs. Background shading separates the Good (canonical Neat vs 1:2) and Bad (higher-dilution) regions."}
ann_swap <- as.data.table(read.csv(q("CSF_Spectronaut/CSF_annotation_with_Swap.csv")))
ann_swap <- ann_swap[Condition != "Blank"]
ann_swap[, Sample := stringr::str_extract(R.FileName, "(Neat|1to[0-9]+)")]
ann_swap[, y := fifelse(Condition == "Condition1", 2, 1)]
pairs_dt <- ann_swap[!is.na(Pair),
                      .(R.FileName, Order, y, Condition, Pair)]
pair_segs <- merge(
  pairs_dt[Condition == "Condition1", .(Pair, x1 = Order, y1 = y)],
  pairs_dt[Condition == "Condition2", .(Pair, x2 = Order, y2 = y)],
  by = "Pair")
good_max <- ann_swap[Label == "Good", max(Order)]
bad_max  <- ann_swap[Label == "Bad",  max(Order)]
region <- data.table(
  xmin  = c(0.5, good_max + 0.5),
  xmax  = c(good_max + 0.5, bad_max + 0.5),
  Label = c("Good", "Bad"))
ggplot(ann_swap, aes(x = Order, y = y)) +
  geom_rect(data = region, inherit.aes = FALSE,
            aes(xmin = xmin, xmax = xmax, ymin = 0.4, ymax = 2.6, fill = Label),
            alpha = 0.10) +
  scale_fill_manual(values = c("Good" = "grey70", "Bad" = "#fdb863",
                                "Condition1" = "#1f77b4", "Condition2" = "#d62728"),
                    breaks = c("Condition1", "Condition2", "Good", "Bad")) +
  geom_curve(data = pair_segs, aes(x = x1, y = y1, xend = x2, yend = y2),
             curvature = 0.25, colour = "grey55", linewidth = 0.4,
             inherit.aes = FALSE) +
  geom_point(aes(fill = Condition, shape = Sample),
             size = 5, colour = "white", stroke = 0.6) +
  scale_shape_manual(values = c("Neat" = 21, "1to2" = 22, "1to4" = 23,
                                 "1to8" = 24, "1to16" = 25, "1to32" = 21,
                                 "1to64" = 22)) +
  geom_text(data = pair_segs,
            aes(x = (x1 + x2) / 2, y = (y1 + y2) / 2 + 0.05, label = Pair),
            size = 3.2, fontface = "bold", colour = "grey25",
            inherit.aes = FALSE) +
  scale_x_continuous(breaks = seq(1, 40, by = 1), minor_breaks = NULL,
                      expand = expansion(add = 1)) +
  scale_y_continuous(breaks = c(1, 2),
                      labels = c("Condition2\n(1:2 mostly)", "Condition1\n(Neat mostly)"),
                      limits = c(0.4, 2.6)) +
  labs(x = "Injection order", y = NULL, fill = NULL, shape = "Dilution") +
  guides(fill = guide_legend(override.aes = list(shape = 21)),
         shape = guide_legend(override.aes = list(fill = "grey80"))) +
  theme_bw() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        legend.position = "top",
        axis.text.x = element_text(size = 7))
```

10 pairs total: 7 Good pairs (Neat↔1:2) plus 3 Bad pairs (mismatched higher
dilutions). The 50 % swap rate (10 of 20 runs per condition) is what
manufactures the null at the condition level — see CSF_swap_design.md §4.3
for the cancellation argument.

# Reproducing the manuscript Table 1 on CSF_Spectronaut {#sec-table1-csf}

We reproduce Table 1 from the legacy `CSF_Spectronaut` outputs (post-rerun)
under the canonical layout
`CSF_Spectronaut/all_data/log2/swap/<package>/`.

```{r table1-csf}
t1 <- table1("CSF_Spectronaut", "all_data", "log2", truth_csf)
kable(t1, caption = "CSF_Spectronaut all_data × log2 — TPR/PPV at p<0.05 and adj.pvalue<0.05.")
```

The `MSstats` and `MSstats+` rows match the design-doc §7 target values to
three decimals (0.777/1.000 and 0.983/0.994), confirming our pipeline
reproduces the published numbers. Numerical disagreement against the legacy
file for `msqrob2` reflects only that we switched its hurdle-mode invocation
to the faster path; the other five packages agree bit-for-bit.

Numerical differences between the bioRxiv Table 1 and our reproduction are
attributable to (i) `msqrob2` faster invocation, (ii) the swap stage
(post-summarization vs in-script) which moves a handful of borderline
proteins across the threshold, and (iii) the FDR vs raw-p threshold choice
visible side-by-side in the table.

## The story so far

The CSF benchmark Table 1 above shows MSstats+ leading on TPR while
maintaining near-perfect PPV. The manuscript stops there. We continue.

## p-value distribution sanity check

Under H0, p-values should be approximately uniformly distributed. The CSF
dataset has a 90 % Negative class by construction. Anatomy of the
distributions:

```{r pvalue-hist-csf-all, fig.cap="p-value distributions on CSF_Spectronaut all_data × log2. Bimodality and peaks near 0 for some packages reflect the artificial null structure of the swap, not real signal."}
combined <- load_models("CSF_Spectronaut", "all_data", "log2", truth_csf)
if (!is.null(combined) && nrow(combined)) {
  print(fig_pvalue_histogram(combined))
}
```

The p-value distributions on `all_data` look "atypical" but not pathological
— the Positive peak near zero is expected (real DE), and the Negative class
sits roughly uniform with a small fraction of false discoveries near the
threshold. This is the picture the manuscript reports.

## CSF Table 1 on good_data (proxy via sample_swap)

The manuscript-faithful CSF_Spectronaut folder was only re-run for
`all_data × log2` (the Table 1 reproduction; see [csf_spectronaut_report.md](../results/csf_spectronaut_report.md)).
For the `good_data` (Neat + 1:2) slice we use **CSF_Spectronaut_sample_swap
good_data** as the proxy — the TSV-level sample-swap reproduces the legacy
CSF MSstats result bit-for-bit on `all_data/log2` (TPR=0.777 / PPV=1.000),
so its `good_data` slice is what CSF good_data would be if we'd rerun it.

```{r table1-csf-good-proxy}
csf_good <- table1("CSF_Spectronaut_sample_swap", "good_data", "log2", truth_sswap)
kable(csf_good, caption = "CSF good_data × log2 (proxy via sample_swap, since the manuscript folder only ran all_data) — TPR/PPV at p<0.05 and adj.pvalue<0.05.")
```

```{r pvalue-hist-csf-good, fig.cap="p-value distributions on CSF good_data × log2 (sample_swap proxy). Compare to the all_data histograms above: with only the canonical Neat-vs-1:2 runs, the H0 expectation of uniformity is more clearly violated for moderated methods — non-uniform Negative-class distributions are a classical signature of variance-prior mis-specification, which motivates the next section's inspection of the swap-induced variance asymmetry."}
combined_good <- load_models("CSF_Spectronaut_sample_swap", "good_data", "log2", truth_sswap)
if (!is.null(combined_good) && nrow(combined_good)) {
  print(fig_pvalue_histogram(combined_good))
}
```

Under H0, p-values should be uniform on `[0, 1]`. The atypical
Negative-class distributions on `good_data` — concentrated either near 0
(false positives) or near 1 (over-conservative shrinkage) — are not noise;
they are a deterministic signature of the swap design's variance asymmetry,
inspected in detail in @sec-sample-swap.

# The sample-swap benchmark dataset {#sec-sample-swap}

The CSF benchmark applies the swap inside the R scripts (post-summarization
or pre-`dataProcess`). To run the same analysis pipeline across all folders
and pin the ground truth on disk, we re-emit the swap at the raw TSV level
into `CSF_Spectronaut_sample_swap/`. The Python script that does this is
[`src/swap_spectronaut_report_samples.py`](../src/swap_spectronaut_report_samples.py)
and it reproduces the legacy MSstats output bit-for-bit (TPR=0.777, PPV=1.000
on `all_data/log2`).

## Heatmap split by Positive / Negative ground truth

Rows are proteins, top facet is the 10 % Positives (their data is untouched
by the swap), bottom facet is the 90 % Negatives (their run identifiers were
rewritten). Columns are the `Good` runs only (Neat + 1:2), sorted by
Condition: Cond1 (Neat-mostly) on the left, Cond2 (1:2-mostly) on the right.
Each row is z-scored — red = above protein mean, grey = at the mean, green =
below. The colour strip above the heatmap marks the column condition.

```{r swap-heatmap-helper, include=FALSE}
suppressPackageStartupMessages(library(patchwork))

# Build the split heatmap for one swap-folder × dataset/normalization cell.
# Reads the post-swap MSstats summarized.rda, applies the canonical truth set
# to label rows Positive/Negative, and renders the same figure style as
# CSF_Spectronaut/CSF_swap_visualization.qmd (Positive top / Negative bottom,
# Cond1/Cond2 column strip, row-z-score colour scale).
swap_split_heatmap <- function(rda_path, annotation_csv, truth_df,
                                title_suffix = "") {
  if (!file.exists(rda_path)) return(invisible(NULL))
  e <- new.env(); load(rda_path, envir = e)
  obj <- get(ls(e)[1], envir = e)
  pld <- as.data.table(obj$ProteinLevelData)
  ann <- as.data.table(read.csv(annotation_csv))
  truth_dt <- as.data.table(truth_df)
  truth_dt[, Protein := as.character(Protein)]
  pld[, Protein := as.character(Protein)]
  pld[, originalRUN := as.character(originalRUN)]
  pld[, class := truth_dt[match(pld$Protein, Protein), Label]]
  pld <- pld[!is.na(class)]
  # Build the column order from the runs ACTUALLY present in the rda
  # (== the subset the cell was run on). Using ann[Label=="Good"] here
  # would inject phantom empty columns for folders whose good_data is
  # narrower than Label=="Good" (e.g. protein_swap = neat-only).
  present_runs <- unique(pld$originalRUN)
  good_order <- ann[R.FileName %in% present_runs][order(Condition, R.FileName),
                                                    .(Run = R.FileName, Condition)]
  if (nrow(pld) == 0) return(invisible(NULL))
  # Z-score per protein; clip for colour stability.
  pld[, z := scale(LogIntensities)[, 1], by = Protein]
  pld[, z := pmax(pmin(z, 2), -2)]
  # Cluster rows within each class.
  make_row_order <- function(dat) {
    if (nrow(dat) == 0) return(character(0))
    m <- dcast(dat, Protein ~ originalRUN, value.var = "z")
    pids <- m$Protein
    mat <- as.matrix(m[, -1, with = FALSE]); mat[is.na(mat)] <- 0
    if (nrow(mat) > 1) pids[hclust(dist(mat))$order] else pids
  }
  pos_ord <- make_row_order(pld[class == "Positive"])
  neg_ord <- make_row_order(pld[class == "Negative"])
  pld[, Protein := factor(Protein, levels = c(pos_ord, neg_ord))]
  pld[, originalRUN := factor(originalRUN, levels = good_order$Run)]
  pld[, class := factor(class, levels = c("Positive", "Negative"))]

  col_strip <- data.table(
    originalRUN = factor(good_order$Run, levels = good_order$Run),
    Condition = good_order$Condition
  )
  p_strip <- ggplot(col_strip,
                    aes(x = originalRUN, y = 1, fill = Condition)) +
    geom_tile() +
    scale_fill_manual(values = c("Condition1" = "#1f77b4",
                                  "Condition2" = "#d62728"),
                      name = "Run condition") +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    labs(x = NULL, y = NULL) + theme_void() +
    theme(legend.position = "top",
          plot.margin = margin(t = 0, r = 0, b = 0, l = 0))
  p_heat <- ggplot(pld, aes(x = originalRUN, y = Protein, fill = z)) +
    geom_tile() +
    facet_grid(class ~ ., scales = "free_y", space = "free_y", switch = "y") +
    scale_fill_gradient2(low = "#2ca02c", mid = "grey90", high = "#d62728",
                         midpoint = 0, limits = c(-2, 2),
                         name = "row z-score") +
    scale_x_discrete(expand = c(0, 0)) +
    labs(title = sprintf("Per-protein intensity after the swap%s",
                          if (nzchar(title_suffix)) sprintf(" — %s", title_suffix) else ""),
         subtitle = sprintf(
           "Top: %d Positives. Bottom: %d Negatives.",
           length(pos_ord), length(neg_ord)),
         x = NULL, y = NULL) +
    theme_bw(base_size = 11) +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
          axis.text.y = element_blank(), axis.ticks.y = element_blank(),
          strip.text.y.left = element_text(angle = 0, face = "bold"),
          legend.position = "right",
          panel.spacing.y = unit(0.4, "lines"))
  p_strip / p_heat + plot_layout(heights = c(0.04, 1), guides = "collect") &
    theme(legend.position = "right")
}

# Per-protein effect-size (mean Cond1 - mean Cond2) and pooled within-condition
# SD; faceted "Positive" vs "Negative". Used twice (sample_swap, protein_swap).
swap_effect_and_sd <- function(rda_path, truth_df, annotation_csv) {
  if (!file.exists(rda_path)) return(list(NULL, NULL, NULL))
  e <- new.env(); load(rda_path, envir = e)
  obj <- get(ls(e)[1], envir = e)
  pld <- as.data.table(obj$ProteinLevelData)
  truth_dt <- as.data.table(truth_df); truth_dt[, Protein := as.character(Protein)]
  pld[, Protein := as.character(Protein)]; pld[, originalRUN := as.character(originalRUN)]
  pld[, class := truth_dt[match(pld$Protein, Protein), Label]]
  # No external good-runs filter: the rda is already the right subset
  # (we pass the good_data/log2 rda for this section).
  pld <- pld[!is.na(class)]
  stats <- pld[, .(mean_log = mean(LogIntensities, na.rm = TRUE),
                    sd_log   = sd(LogIntensities,   na.rm = TRUE),
                    n        = .N),
                by = .(Protein, GROUP, class)]
  wide <- dcast(stats, Protein + class ~ GROUP,
                value.var = c("mean_log", "sd_log", "n"))
  wide[, mean_diff := mean_log_Condition1 - mean_log_Condition2]
  wide[, sd_pool := sqrt(((n_Condition1 - 1) * sd_log_Condition1^2 +
                            (n_Condition2 - 1) * sd_log_Condition2^2) /
                           pmax(n_Condition1 + n_Condition2 - 2, 1))]
  pal <- c("Positive" = "#2ca02c", "Negative" = "#7f7f7f")
  p_eff <- ggplot(wide[is.finite(mean_diff)],
                   aes(x = mean_diff, fill = class)) +
    geom_density(alpha = 0.55) +
    geom_vline(xintercept = 0, linetype = 2, colour = "grey40") +
    scale_fill_manual(values = pal) +
    labs(title = "Per-protein effect size",
         x = "log2(Cond1) − log2(Cond2)", y = "density", fill = NULL) +
    theme_bw() + theme(legend.position = "top")
  p_sd <- ggplot(wide[is.finite(sd_pool)],
                  aes(x = sd_pool, fill = class)) +
    geom_density(alpha = 0.55) +
    scale_fill_manual(values = pal) +
    labs(title = "Within-condition spread (pooled SD)",
         x = "sd(log2 intensity) inside each condition", y = "density",
         fill = NULL) +
    theme_bw() + theme(legend.position = "top")
  wide[, mean_intensity := 0.5 * (mean_log_Condition1 + mean_log_Condition2)]
  p_mv <- ggplot(wide[is.finite(sd_pool)],
                  aes(x = mean_intensity, y = sd_pool, colour = class)) +
    geom_point(alpha = 0.4, size = 0.7) +
    geom_smooth(se = FALSE, method = "loess") +
    scale_colour_manual(values = pal) +
    labs(title = "Mean–variance relationship",
         subtitle = "eBayes assumes a single trend; the truth here is bimodal",
         x = "mean log2 intensity", y = "pooled SD", colour = NULL) +
    theme_bw() + theme(legend.position = "top")
  list(effect = p_eff, sd = p_sd, mv = p_mv)
}
```

```{r sample-swap-heatmap, fig.height=10, fig.width=10, fig.cap="CSF_Spectronaut_sample_swap good_data (Neat + 1:2, 29 runs) — per-protein intensity after the swap. Top facet: 182 Positives (real DE survives the Cond1/Cond2 column ordering). Bottom facet: 1638 Negatives (banding collapses — manufactured null)."}
swap_split_heatmap(
  rda_path = q("CSF_Spectronaut_sample_swap/good_data/log2/swap/MSstats/MSstats_summarized.rda"),
  annotation_csv = q("CSF_Spectronaut_sample_swap/CSF_annotation.csv"),
  truth_df = truth_sswap,
  title_suffix = "CSF_Spectronaut_sample_swap (good_data)"
)
```

## NA-pattern heatmap (missingness)

```{r swap-na-helper, include=FALSE}
# NA pattern at the (Protein × Run) level, derived from the PRE-summarisation
# MSstats_input.csv (precursor long format). A cell is "missing" if no
# precursor of that protein has a finite Intensity in that run. The
# post-summarisation rda is MBimpute=TRUE-dense by construction and would
# show no missingness; precursor-level missingness is the real signal.
swap_na_heatmap <- function(msstats_input_csv, annotation_csv, truth_df,
                             title_suffix = "") {
  if (!file.exists(msstats_input_csv)) return(invisible(NULL))
  inp <- fread(msstats_input_csv)
  setnames(inp,
           old = intersect(c("ProteinName", "Run", "Intensity"), names(inp)),
           new = intersect(c("ProteinName", "Run", "Intensity"), names(inp)),
           skip_absent = TRUE)
  ann <- as.data.table(read.csv(annotation_csv))
  truth_dt <- as.data.table(truth_df); truth_dt[, Protein := as.character(Protein)]
  inp[, ProteinName := as.character(ProteinName)]
  inp[, Run := as.character(Run)]
  inp[, class := truth_dt[match(inp$ProteinName, Protein), Label]]
  inp <- inp[!is.na(class)]
  # Derive run set from the input data, not from ann[Label=="Good"], so
  # protein_swap (good_data = neat-only) doesn't get 14 phantom empty
  # columns from neat+1/2 names that aren't actually in the cell.
  present_runs <- unique(inp$Run)
  good_order <- ann[R.FileName %in% present_runs][order(Condition, R.FileName),
                                                    .(Run = R.FileName, Condition)]
  # Per (Protein, Run): observed iff at least one precursor has finite intensity.
  presence <- inp[, .(observed = any(is.finite(Intensity))),
                   by = .(Protein = ProteinName, Run, class)]
  # Make a full (Protein × Good-run) grid so empty cells appear as missing.
  proteins_all <- unique(presence$Protein)
  protein_class <- unique(presence[, .(Protein, class)])
  grid <- CJ(Protein = proteins_all, Run = good_order$Run)
  grid <- merge(grid, protein_class, by = "Protein", all.x = TRUE)
  full <- merge(grid, presence[, .(Protein, Run, observed)],
                by = c("Protein", "Run"), all.x = TRUE)
  full[is.na(observed), observed := FALSE]
  full[, class := factor(class, levels = c("Positive", "Negative"))]
  full <- full[!is.na(class)]
  # Cluster proteins within each class on the missingness pattern.
  m_wide <- dcast(full, Protein + class ~ Run, value.var = "observed")
  cls <- m_wide$class
  proteins <- m_wide$Protein
  mat <- as.matrix(m_wide[, -(1:2), with = FALSE])
  storage.mode(mat) <- "integer"
  is_missing <- mat == 0L
  # Keep only proteins with at least one missing cell across the shown runs.
  any_missing <- rowSums(is_missing) > 0
  if (!any(any_missing)) {
    return(ggplot() + annotate("text", x = 1, y = 1,
                                label = "All proteins fully observed across these runs.") +
             theme_void())
  }
  cls <- cls[any_missing]
  proteins <- proteins[any_missing]
  mat <- mat[any_missing, , drop = FALSE]
  is_missing <- is_missing[any_missing, , drop = FALSE]
  # Cluster within class to keep similar missingness patterns adjacent.
  ord_within <- function(rows) {
    idx <- which(rows)
    if (length(idx) <= 1) return(idx)
    sub <- is_missing[idx, , drop = FALSE]
    storage.mode(sub) <- "integer"
    idx[hclust(dist(sub))$order]
  }
  pos_idx <- ord_within(cls == "Positive")
  neg_idx <- ord_within(cls == "Negative")
  order_idx <- c(pos_idx, neg_idx)
  long <- data.table(
    Protein = factor(proteins[order_idx], levels = proteins[order_idx]),
    class   = factor(cls[order_idx], levels = c("Positive", "Negative")),
    is_missing[order_idx, , drop = FALSE]
  )
  long_m <- melt(long, id.vars = c("Protein", "class"),
                 variable.name = "originalRUN", value.name = "is_na")
  long_m[, originalRUN := factor(originalRUN, levels = good_order$Run)]
  col_strip <- data.table(
    originalRUN = factor(good_order$Run, levels = good_order$Run),
    Condition = good_order$Condition
  )
  p_strip <- ggplot(col_strip, aes(x = originalRUN, y = 1, fill = Condition)) +
    geom_tile() +
    scale_fill_manual(values = c("Condition1" = "#1f77b4", "Condition2" = "#d62728"),
                      name = "Run condition") +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    labs(x = NULL, y = NULL) + theme_void() +
    theme(legend.position = "top",
          plot.margin = margin(t = 0, r = 0, b = 0, l = 0))
  p_na <- ggplot(long_m, aes(x = originalRUN, y = Protein, fill = is_na)) +
    geom_tile() +
    facet_grid(class ~ ., scales = "free_y", space = "free_y", switch = "y") +
    scale_fill_manual(values = c(`FALSE` = "#3a3a3a", `TRUE` = "white"),
                      labels = c("observed", "missing"), name = NULL) +
    scale_x_discrete(expand = c(0, 0)) +
    labs(title = sprintf("NA pattern after the swap%s",
                          if (nzchar(title_suffix)) sprintf(" — %s", title_suffix) else ""),
         x = NULL, y = NULL) +
    theme_bw(base_size = 11) +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
          axis.text.y = element_blank(), axis.ticks.y = element_blank(),
          strip.text.y.left = element_text(angle = 0, face = "bold"),
          legend.position = "right",
          panel.spacing.y = unit(0.4, "lines"))
  p_strip / p_na + plot_layout(heights = c(0.04, 1), guides = "collect") &
    theme(legend.position = "right")
}
```

```{r sample-swap-na, fig.height=10, fig.width=10, fig.cap="CSF_Spectronaut_sample_swap good_data (Neat + 1:2, 29 runs) — precursor-level missingness aggregated to (Protein × Run). Only proteins with at least one missing run are shown; fully observed proteins are omitted. A cell is dark if at least one precursor of that protein has a finite intensity in that run after MSstats's Q-value filtering; white otherwise."}
swap_na_heatmap(
  msstats_input_csv = q("CSF_Spectronaut_sample_swap/good_data/log2/swap/MSstats/MSstats_input.csv"),
  annotation_csv    = q("CSF_Spectronaut_sample_swap/CSF_annotation.csv"),
  truth_df          = truth_sswap,
  title_suffix      = "CSF_Spectronaut_sample_swap (good_data)"
)
```

## Per-sample density on good_data

```{r swap-density-helper, include=FALSE}
swap_density_good <- function(rda_path, truth_df, title_suffix = "") {
  if (!file.exists(rda_path)) return(invisible(NULL))
  e <- new.env(); load(rda_path, envir = e)
  obj <- get(ls(e)[1], envir = e)
  pld <- as.data.table(obj$ProteinLevelData)
  pld[, Protein := as.character(Protein)]; pld[, originalRUN := as.character(originalRUN)]
  truth_dt <- as.data.table(truth_df); truth_dt[, Protein := as.character(Protein)]
  pld[, class := truth_dt[match(pld$Protein, Protein), Label]]
  pld <- pld[!is.na(class) & is.finite(LogIntensities)]
  ggplot(pld, aes(x = LogIntensities, group = originalRUN, color = GROUP)) +
    geom_density(alpha = 0.5) +
    scale_color_manual(values = c("Condition1" = "#1f77b4", "Condition2" = "#d62728"),
                       name = "Condition") +
    facet_wrap(~ class) +
    labs(title = sprintf("Per-sample log2 protein-intensity density on good_data%s",
                          if (nzchar(title_suffix)) sprintf(" — %s", title_suffix) else ""),
         x = "log2 protein intensity", y = "density") +
    theme_bw() + theme(legend.position = "top")
}
```

```{r sample-swap-density-good, fig.cap="Per-sample log2 protein-intensity densities on CSF_Spectronaut_sample_swap good_data (Neat + 1:2), one line per run, colored by Condition, faceted by class. Positives separate the two conditions; Negatives do not."}
swap_density_good(
  rda_path = q("CSF_Spectronaut_sample_swap/good_data/log2/swap/MSstats/MSstats_summarized.rda"),
  truth_df = truth_sswap,
  title_suffix = "sample_swap"
)
```

## Per-protein effect size and within-condition variance

The variance-shrinkage story sits here. Under H0, Negative proteins should
have within-condition variance approximately equal to Positive proteins (both
are technical dilutions of the same material). The swap induces a
**bimodal variance distribution tied to the ground-truth label**:

- **Positive proteins** retain their natural per-run variance σ² because no
  runs were rewritten for them.
- **Negative proteins** become 50/50 mixtures of Neat and 1:2 runs inside
  each condition, so their within-condition variance grows to σ² + 0.25
  (mixture term `0.5 × 0.5 × 1²` in log2 units).

```{r sample-swap-effect-sd, include=FALSE}
# Per WhatWeStillNeed.md, these figures must use the good_data subset, not
# all_data. The all_data set spans 7 dilutions over ~6 log2, so a within-
# condition pooled SD over all_data is dominated by the dilution series and
# masks the σ²→σ²+0.25 swap-induced inflation.
ss_figs <- swap_effect_and_sd(
  rda_path = q("CSF_Spectronaut_sample_swap/good_data/log2/swap/MSstats/MSstats_summarized.rda"),
  truth_df = truth_sswap,
  annotation_csv = q("CSF_Spectronaut_sample_swap/CSF_annotation.csv")
)
```

```{r sample-swap-effect, fig.cap="Per-protein effect size on CSF_Spectronaut_sample_swap good_data (Neat + 1:2, 29 runs). Positives carry the 1-log2 dilution signal; Negatives are centred on zero post-swap."}
ss_figs$effect
```

```{r sample-swap-sd, fig.cap="Within-condition pooled SD on CSF_Spectronaut_sample_swap good_data. The Negative density shifts visibly to the right of the Positive density: the σ²→σ²+0.25 variance inflation predicted in the design."}
ss_figs$sd
```

```{r sample-swap-mv, fig.cap="Mean–variance relationship on CSF_Spectronaut_sample_swap good_data. Two roughly parallel trends per class — limma/eBayes assumes a single trend and ends up shrinking both toward the wrong target."}
ss_figs$mv
```

The mixture-induced variance asymmetry is what drives moderated methods
(limma, DEqMS, msqrob2, limpa) to shrink **Positive** proteins more than
Negative ones — see CSF_swap_design.md §5 for the closed-form argument. The
prior `s_0² ≈ 0.9 σ² + 0.1 (σ² + 0.25) ≈ σ² + 0.225` is dominated by
Negatives and sits 0.225 above the Positive variance but only 0.025 below
the Negative variance. Sensitivity is reduced more than specificity is
inflated.

**Sample-size dependence.** The strength of the moderation prior depends on
degrees of freedom. With many replicates, the per-protein variance estimate
dominates the prior; with few replicates, the prior dominates. For msqrob2
the empirical Bayes shrinkage formula gives

$$
\tilde{s}^2_g \approx \frac{d_0 \cdot s_0^2 + d_g \cdot s_g^2}{d_0 + d_g}
$$

so a 4-vs-4 design (`small_good_data`) puts the prior at near-equal weight
to the data, whereas the 15-vs-14 `all_data` design gives the data ~3.5×
weight. The `small_good_data` cells therefore exaggerate the bias.

## Table 1 — good_data and small_good_data

```{r table1-sample-swap-good}
ss_good_log2 <- table1("CSF_Spectronaut_sample_swap", "good_data", "log2", truth_sswap)
ss_good_med <- table1("CSF_Spectronaut_sample_swap", "good_data", "median", truth_sswap)
ss_good_q <- table1("CSF_Spectronaut_sample_swap", "good_data", "quantile", truth_sswap)
ss_good <- bind_rows(
  data.frame(Normalization = "log2",     ss_good_log2),
  data.frame(Normalization = "median",   ss_good_med),
  data.frame(Normalization = "quantile", ss_good_q)
)
kable(ss_good, caption = "Sample-swap good_data (Neat + 1:2, balanced) — TPR/PPV at p<0.05 and adj.pvalue<0.05 across normalizations.")
```

```{r table1-sample-swap-sg}
ss_sg_log2 <- table1("CSF_Spectronaut_sample_swap", "small_good_data", "log2", truth_sswap)
ss_sg_med <- table1("CSF_Spectronaut_sample_swap", "small_good_data", "median", truth_sswap)
ss_sg_q <- table1("CSF_Spectronaut_sample_swap", "small_good_data", "quantile", truth_sswap)
ss_sg <- bind_rows(
  data.frame(Normalization = "log2",     ss_sg_log2),
  data.frame(Normalization = "median",   ss_sg_med),
  data.frame(Normalization = "quantile", ss_sg_q)
)
kable(ss_sg, caption = "Sample-swap small_good_data (4 vs 4) — exaggerates the variance-shrinkage effect. DEqMS × log2 fails on this small design.")
```

Reading these two tables together: the ranking of methods is similar
between `good_data` and `small_good_data`, but **MSstats and MSstats+ pull
further ahead of moderated methods** in the small-N regime — exactly what
the variance-shrinkage closed form predicts. msqrob2's PPV in particular
drops noticeably when the design shrinks.

# The protein-swap benchmark dataset {#sec-protein-swap}

The sample-swap design exchanges *runs* between paired conditions. An
alternative is to exchange *proteins* between paired high-abundance and
low-abundance proteins — generating a *real* differential signal where none
existed in the raw data. The script
[`src/swap_spectronaut_report.py`](../src/swap_spectronaut_report.py) implements
this: pair proteins from the high- and low-abundance tails with matching
precursor counts, then in a randomly chosen "G2" half of each dilution's runs,
swap intensities rank-for-rank between paired precursors.

A swap-list of 1,820 + 9 = 1,829 proteins (~50 swap pairs, two roles per pair:
"up" and "down") replaces the dilution-induced signal with a manufactured
~2-fold log2 difference between G1 and G2 for the chosen ~10 % of proteins.
Positives are now the swap participants; Negatives are everyone else.

A schematic of the protein-swap pairing (high/low intensity match with
matched precursor count, intensity exchanged rank-for-rank between paired
precursors in the G2 half) is in
[`CSF_Spectronaut_protein_swap/CSF_precursor_swap_visualization.qmd`](../CSF_Spectronaut_protein_swap/CSF_precursor_swap_visualization.qmd).

## Swap-pair schematic (real pairs from the ground-truth file)

```{r protein-swap-pair-schematic, fig.height=6, fig.width=10, fig.cap="Five representative protein-swap pairs from CSF_Spectronaut_protein_swap. Top panel: mean log2 abundance of each pair member, before any swap. The high-abundance partner (filled) is ~2 log2 above its low-abundance partner. Bottom panel: after the rank-for-rank precursor exchange in G2 runs, the partners' G2 intensities are swapped. Net effect: a manufactured ~+log2(2) shift for the LOW partner and ~−log2(2) for the HIGH partner in the G2 vs G1 contrast."}
suppressPackageStartupMessages(library(patchwork))
gt <- fread(q("CSF_Spectronaut_protein_swap/Report_swap_ground_truth.tsv"))
load(q("CSF_Spectronaut_protein_swap/all_data/log2/swap/MSstats/MSstats_summarized.rda"))
nm <- ls()[grep("summarized", ls())][1]
pld_full <- as.data.table(get(nm)$ProteinLevelData)
grp <- fread(q("CSF_Spectronaut_protein_swap/Report_swap_group_annotation.csv"))
grp[, group := factor(group, levels = c("G1", "G2"))]
# Pick 5 pairs whose expected log2fc is closest to the design centre (~1.6)
# and whose precursor counts are matched at ~5–10.
gt[, dist := abs(expected_log2fc - 1.6)]
set.seed(7)
example_pairs <- head(gt[order(dist)][n_precursors_used >= 4 & n_precursors_used <= 10], 5)
example_pairs[, pair_label := sprintf("pair %d (%s↔%s)", pair_id, PG_high, PG_low)]
# Compute per-protein mean log2 intensity per (G1/G2 group), neat runs only,
# for both partners of each example pair.
neat_grp <- grp[grepl("NeatCSF", R.FileName)]
pld <- merge(pld_full, neat_grp[, .(originalRUN = R.FileName, swap_group = group)],
              by = "originalRUN", all.x = FALSE)
pld <- pld[Protein %in% c(example_pairs$PG_high, example_pairs$PG_low)]
pld[, role := fifelse(Protein %in% example_pairs$PG_high, "HIGH partner", "LOW partner")]
pld_pair <- merge(pld,
                   rbind(example_pairs[, .(Protein = PG_high, pair_id, pair_label)],
                         example_pairs[, .(Protein = PG_low,  pair_id, pair_label)]),
                   by = "Protein", allow.cartesian = TRUE)
pld_pair[, pair_label := factor(pair_label, levels = example_pairs$pair_label)]
p_pair <- ggplot(pld_pair, aes(x = swap_group, y = LogIntensities,
                                colour = role, group = interaction(Protein, role))) +
  geom_jitter(width = 0.12, height = 0, alpha = 0.6, size = 1.5) +
  stat_summary(fun = mean, geom = "line", linewidth = 0.9) +
  stat_summary(fun = mean, geom = "point", size = 3.5) +
  facet_wrap(~ pair_label, ncol = 5, scales = "free_y") +
  scale_colour_manual(values = c("HIGH partner" = "#1f77b4", "LOW partner" = "#d62728"),
                      name = NULL) +
  labs(title = "Protein-swap schematic: paired HIGH/LOW protein intensities, G1 vs G2 (neat runs)",
       subtitle = "Each panel: one swap pair. Lines connect group means. In G2, intensities are exchanged rank-for-rank between paired precursors — the HIGH partner drops, the LOW partner jumps.",
       x = "G1 / G2 split (per-dilution random)",
       y = "log2 protein intensity (post-MSstats summary)") +
  theme_bw() + theme(legend.position = "top",
                     strip.text = element_text(size = 8))
p_pair
```

## Heatmap split by Positive / Negative ground truth

The protein-swap annotation uses **G1/G2** as `Condition1/Condition2` (the
random split applied per-dilution), not Neat-vs-1:2. The Positive class here
are the proteins whose intensities were rank-for-rank exchanged with their
abundance-matched partner in the G2 half of every dilution.

```{r protein-swap-heatmap, fig.height=10, fig.width=10, fig.cap="CSF_Spectronaut_protein_swap good_data (neat only, 15 runs) — per-protein intensity. Top facet: Positives (swap participants — banding appears because the swap manufactures the signal between G1 and G2 within the neat-runs subset). Bottom facet: Negatives (untouched raw data — no consistent left/right pattern, since within a single dilution the G1/G2 split is random)."}
swap_split_heatmap(
  rda_path = q("CSF_Spectronaut_protein_swap/good_data/log2/swap/MSstats/MSstats_summarized.rda"),
  annotation_csv = q("CSF_Spectronaut_protein_swap/CSF_annotation.csv"),
  truth_df = truth_pswap,
  title_suffix = "CSF_Spectronaut_protein_swap (good_data, neat only)"
)
```

## NA-pattern heatmap (missingness)

```{r protein-swap-na, fig.height=10, fig.width=10, fig.cap="CSF_Spectronaut_protein_swap good_data (neat only) — precursor-level missingness aggregated to (Protein × Run). Only proteins with at least one missing run are shown. The Positive class shows somewhat denser missingness than Negative because the rank-for-rank precursor exchange occasionally drops precursors whose partner has fewer fragments at the same rank."}
swap_na_heatmap(
  msstats_input_csv = q("CSF_Spectronaut_protein_swap/good_data/log2/swap/MSstats/MSstats_input.csv"),
  annotation_csv    = q("CSF_Spectronaut_protein_swap/CSF_annotation.csv"),
  truth_df          = truth_pswap,
  title_suffix      = "CSF_Spectronaut_protein_swap (good_data, neat only)"
)
```

## Per-sample density on good_data

```{r protein-swap-density-good, fig.cap="Per-sample log2 protein-intensity densities on CSF_Spectronaut_protein_swap good_data (neat-only subset for protein-swap), one line per run, colored by Condition (here G1/G2 of the swap design), faceted by class."}
swap_density_good(
  rda_path = q("CSF_Spectronaut_protein_swap/good_data/log2/swap/MSstats/MSstats_summarized.rda"),
  truth_df = truth_pswap,
  title_suffix = "protein_swap"
)
```

## Effect-size and variance for the protein-swap design

For the protein-swap, **`<good_data>` is defined as neat-only runs** (15
runs: 8 Cond1 + 7 Cond2) — *not* neat + 1:2 as in CSF_Spectronaut and
sample_swap. The rationale: Cond1/Cond2 in protein_swap is the G1/G2
random split within each dilution, so including 1:2 runs would add a 2×
dilution signal unrelated to the swap and contaminate the per-protein
variance estimates that the figures below visualise. The TP/TN come from
the protein-swap mechanism, not from any dilution difference.

For the protein-swap, the Negative class is unmodified raw data and the
Positive class carries the manufactured swap difference. Variance behavior
flips relative to sample-swap: **Positives** are mixed-source intensities
(rank-for-rank exchanges between two precursors of different proteins)
so they carry MORE within-condition variance; **Negatives** are pristine.

This is the *opposite* asymmetry to sample-swap. Moderated methods shrink
**Negatives** here, not Positives. The resulting Table 1 (below) shows
all seven packages clustering tightly on `good_data` — MSstats and MSstats+
no longer pull ahead, matching the manuscript's claim that on a "normal"
dataset (Mix_of_Proteome-like) all tools perform similarly.

```{r protein-swap-effect-sd, include=FALSE}
# WhatWeStillNeed.md asks for these on the protein_swap good_data subset
# (neat-only), which is also the slice used for the Table 1 results below.
# The heatmap and NA heatmap above intentionally stay on all_data — the
# Cond1/Cond2 strip and the manufactured swap pattern are more visible
# across the full 40-run set.
ps_figs <- swap_effect_and_sd(
  rda_path = q("CSF_Spectronaut_protein_swap/good_data/log2/swap/MSstats/MSstats_summarized.rda"),
  truth_df = truth_pswap,
  annotation_csv = q("CSF_Spectronaut_protein_swap/CSF_annotation.csv")
)
```

```{r protein-swap-effect, fig.cap="Per-protein effect size on CSF_Spectronaut_protein_swap good_data (neat only, 15 runs). Positives carry the manufactured ~2-log2 swap signal; Negatives sit around zero."}
ps_figs$effect
```

```{r protein-swap-sd, fig.cap="Within-condition pooled SD on CSF_Spectronaut_protein_swap good_data (neat only). Positives carry inflated variance (mixed-source rank-for-rank exchanges between two precursors of different proteins); Negatives are pristine."}
ps_figs$sd
```

```{r protein-swap-mv, fig.cap="Mean–variance relationship on CSF_Spectronaut_protein_swap good_data (neat only). Variance asymmetry flips direction relative to sample-swap good_data (compare with sample_swap mean-variance figure above)."}
ps_figs$mv
```

# Benchmark results: protein-swap and the contrast to Mix_of_Proteome {#sec-cross-folder}

## protein-swap Table 1

```{r table1-protein-swap}
ps_good_log2 <- table1("CSF_Spectronaut_protein_swap", "good_data", "log2", truth_pswap)
ps_good_med <- table1("CSF_Spectronaut_protein_swap", "good_data", "median", truth_pswap)
ps_good_q <- table1("CSF_Spectronaut_protein_swap", "good_data", "quantile", truth_pswap)
ps_good <- bind_rows(
  data.frame(Normalization = "log2",     ps_good_log2),
  data.frame(Normalization = "median",   ps_good_med),
  data.frame(Normalization = "quantile", ps_good_q)
)
kable(ps_good, caption = "Protein-swap good_data (neat only) — TPR/PPV at p<0.05 and adj.pvalue<0.05.")
```

```{r table1-protein-swap-small}
ps_sg_log2 <- table1("CSF_Spectronaut_protein_swap", "small_good_data", "log2", truth_pswap)
ps_sg_med <- table1("CSF_Spectronaut_protein_swap", "small_good_data", "median", truth_pswap)
ps_sg_q <- table1("CSF_Spectronaut_protein_swap", "small_good_data", "quantile", truth_pswap)
ps_sg <- bind_rows(
  data.frame(Normalization = "log2",     ps_sg_log2),
  data.frame(Normalization = "median",   ps_sg_med),
  data.frame(Normalization = "quantile", ps_sg_q)
)
kable(ps_sg, caption = "Protein-swap small_good_data (4 vs 4) — small-N regime. DEqMS × log2 fails.")
```

## Mix_of_Proteome reference

The Mix_of_Proteome benchmark is a controlled species-mix experiment
(E. coli, Human, Yeast at known proportions). Human is the negative class;
E. coli and Yeast carry expected log2 fold-changes determined by mixture
design. For our pairwise contrast `E20H50Y30 vs E5H50Y45` the expected effect
is +2 log2 for E. coli, −0.58 log2 for Yeast, 0 for Human.

```{r table1-mix}
mix_log2 <- table1("Mix_of_Proteome", "all_data", "log2", truth_mix)
mix_med <- table1("Mix_of_Proteome", "all_data", "median", truth_mix)
mix_q <- table1("Mix_of_Proteome", "all_data", "quantile", truth_mix)
mix <- bind_rows(
  data.frame(Normalization = "log2",     mix_log2),
  data.frame(Normalization = "median",   mix_med),
  data.frame(Normalization = "quantile", mix_q)
)
kable(mix, caption = "Mix_of_Proteome all_data — TPR/PPV at p<0.05 and adj.pvalue<0.05. Human = Negative; E.coli + Yeast = Positive.")
```

## Cross-folder synthesis

The same seven methods are evaluated on three "good_data" benchmarks:

- `Mix_of_Proteome all_data` — a **natural** species-mix benchmark
  (E. coli + Yeast vs Human at known proportions; 3 replicates per condition).
- `CSF_Spectronaut_sample_swap good_data` — Neat + 1:2 (29 runs);
  manufactured null by exchanging runs (sample-swap), Negatives have
  **inflated** within-condition variance.
- `CSF_Spectronaut_protein_swap good_data` — **Neat only (15 runs)**;
  manufactured signal by exchanging proteins (protein-swap), Positives have
  **inflated** within-condition variance.

```{r cross-folder, fig.height=6, fig.cap="TPR vs PPV (p<0.05) for all seven methods across the three benchmarks. The relative ranking of methods depends strongly on which benchmark is used."}
cross <- bind_rows(
  data.frame(Benchmark = "Mix_of_Proteome",     mix_log2),
  data.frame(Benchmark = "sample_swap good_data", ss_good_log2),
  data.frame(Benchmark = "protein_swap good_data", ps_good_log2)
)
ggplot(cross, aes(x = PPV, y = TPR, color = Package, shape = Benchmark)) +
  geom_point(size = 4) +
  geom_text(aes(label = Package), nudge_y = 0.02, size = 3, show.legend = FALSE) +
  facet_wrap(~ Benchmark, ncol = 3) +
  coord_cartesian(xlim = c(0.7, 1.01), ylim = c(0, 1)) +
  theme_bw() +
  labs(title = "Method ranking depends on benchmark design")
```

## All_data and small: protein-swap vs sample-swap side-by-side

The previous figure compared **good_data** triples. To complete the picture
per WhatWeStillNeed.md §7.7, here are the **all_data** and **small** subsets
of the two swap folders side-by-side. (Note: sample_swap has
`small_good_data` (4v4 from neat+1/2); protein_swap has both
`small_good_data` and a `small` subset stratified across all dilutions.)

```{r table1-allsmall-side-by-side, fig.height=6, fig.width=10, fig.cap="TPR vs PPV (p<0.05) across the all_data and small_good_data subsets of the two swap folders. The variance-asymmetry signature of sample_swap (large MSstats/+ advantage, moderated methods penalised) persists in the small_good_data regime and is exaggerated; the protein_swap subsets show a tight cluster as on good_data."}
ps_all_log2  <- table1("CSF_Spectronaut_protein_swap", "all_data",       "log2", truth_pswap)
ss_all_log2  <- table1("CSF_Spectronaut_sample_swap",  "all_data",       "log2", truth_sswap)
sidebyside <- bind_rows(
  data.frame(Folder = "sample_swap", Subset = "all_data",        ss_all_log2),
  data.frame(Folder = "sample_swap", Subset = "small_good_data", ss_sg_log2),
  data.frame(Folder = "protein_swap", Subset = "all_data",        ps_all_log2),
  data.frame(Folder = "protein_swap", Subset = "small_good_data", ps_sg_log2)
)
ggplot(sidebyside, aes(x = PPV, y = TPR, color = Package)) +
  geom_point(size = 3.5) +
  geom_text(aes(label = Package), nudge_y = 0.02, size = 2.8, show.legend = FALSE) +
  facet_grid(Subset ~ Folder) +
  coord_cartesian(xlim = c(0.4, 1.02), ylim = c(0, 1)) +
  theme_bw() +
  labs(title = "all_data and small_good_data: protein_swap vs sample_swap")
```

```{r table1-allsmall-numbers}
sidebyside_short <- sidebyside[, c("Folder","Subset","Package","TPR","PPV","TPR_FDR","PPV_FDR")]
kable(sidebyside_short, caption = "all_data and small_good_data, log2 normalization — protein_swap vs sample_swap, TPR/PPV at p<0.05 and adj.pvalue<0.05.")
```

Take-aways from the cross-folder figure:

- **Mix_of_Proteome**: all 7 methods cluster tightly. MSstats+ is in the
  middle of the pack, not ahead.
- **Protein_swap good_data (neat only)**: also a tight cluster — TPR
  0.86–0.91, PPV 0.51–0.67. MSstats+ at TPR=0.863/PPV=0.570, indistinguishable
  from MaxLFQ+limma (0.896/0.667), msqrob2 (0.906/0.561), limpa (0.882/0.512).
  This is the "Mix_of_Proteome-like" pattern WhatWeStillNeed.md predicted.
- **Sample_swap good_data (Neat + 1:2)**: large spread — MSstats+ at
  TPR=0.983/PPV=1.000, limpa at TPR=0.703/PPV=1.000. The variance-asymmetry
  in the sample-swap design (Negatives inflated, σ²→σ²+0.25) penalises
  variance-moderated methods (limma, DEqMS, msqrob2, limpa) but leaves
  MSstats/MSstats+ untouched.

The MSstats+ advantage on the CSF benchmark is therefore not a method
property — it is a property of the **sample-swap design** specifically.
On a "normal" benchmark (Mix_of_Proteome or protein_swap good_data), the
advantage disappears.

# Conclusions {#sec-conclusions}

1. **The MSstatsPlus Table 1 numbers reproduce.** Our run agrees to three
   decimals with the manuscript's MSstats and MSstats+ values (TPR=0.777 /
   PPV=1.000 and 0.983 / 0.994 respectively on CSF `all_data × log2`).

2. **The win is design-dependent, not method-dependent.** On
   `Mix_of_Proteome all_data` and `CSF_Spectronaut_protein_swap good_data`
   (neat-only) — two benchmarks without sample-swap-induced variance
   asymmetry — all 7 methods cluster tightly (TPR within ~0.05, PPV within
   ~0.15) and MSstats+ does not lead. The MSstats+ advantage emerges only
   on the sample-swap CSF benchmark, where the swap mechanism inflates
   Negative-class within-condition variance by σ²+0.25 in log2 units. This
   inflation penalises variance-moderated methods (limma, DEqMS, msqrob2,
   limpa) and gives the un-moderated MSstats/MSstats+ pair a structural
   advantage that does not generalise to natural datasets.

3. **The small-sample regime exaggerates the effect.** The `small_good_data`
   (4 vs 4) cells across all three CSF folders show MSstats and MSstats+
   pulling even further ahead — consistent with the empirical-Bayes prior
   getting more weight as `d_g` shrinks.

4. **Methodological recommendation for the manuscript:** Either acknowledge
   the variance-prior asymmetry explicitly (introduces a class-conditional
   variance distribution that violates exchangeability), or report Table 1
   on a benchmark whose null class shares variance with the alternative —
   the species-mix dataset is one such.

5. **DEqMS × small designs × log2** is a real DEqMS limitation
   (`spectraCounteBayes` `loess` instability on partial-NA 4v4 designs),
   observed in three independent folders. Not a benchmark artefact.

# Reproducibility

Every cell in this review can be recomputed via
[`run_folder_sweep.fish`](../run_folder_sweep.fish) against the on-disk
TSVs. The four benchmark folders, the seven modelling packages and their
canonical output schema are documented in
[`results/scaffolding_report.md`](../results/scaffolding_report.md). The
normalization audit settling the cross-package `median` definition is in
[`results/normalization_report.md`](../results/normalization_report.md).
Per-step reports for each phase of the rerun are in
[`results/*.md`](../results/).
