---
title: "Supplement to the MSstats+ benchmark review"
subtitle: "Reproducible CSF Spectronaut sample-swap and protein-swap diagnostics"
author: "Witold Wolski"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    fig-width: 8
    fig-height: 5
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    fig-width: 8
    fig-height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(ggplot2)
  library(knitr)
  library(patchwork)
  library(tidyr)
})

QUANT_ROOT <- normalizePath(
  file.path(dirname(knitr::current_input(dir = TRUE)), ".."),
  mustWork = FALSE
)
if (!dir.exists(QUANT_ROOT)) {
  QUANT_ROOT <- normalizePath("..", mustWork = FALSE)
}
q <- function(...) file.path(QUANT_ROOT, ...)

source(q("R/ground_truth.R"))
source(q("R/figures.R"))
source(q("benchmark_experiments_functions.R"))

model_filename <- list(
  "MSstats"      = "MSstats_model.csv",
  "MSstats+"     = "MSstats+_model.csv",
  "MaxLFQ_limma" = "limma_model.csv",
  "DEqMS"        = "deqms_model.csv",
  "msqrob2"      = "msqrob2_model.csv",
  "prolfqua"     = "prolfqua_model.csv",
  "limpa"        = "limpa_model.csv"
)
packages <- names(model_filename)

safe_read_csv <- function(path) {
  if (!file.exists(path)) {
    return(NULL)
  }
  tryCatch(read.csv(path), error = function(e) NULL)
}

truth_sets <- list(
  csf = read.csv(q("CSF_Spectronaut/CSF_protein_swap_list.csv")),
  protein_swap = read.csv(q("CSF_Spectronaut_protein_swap/CSF_protein_swap_list.csv")),
  sample_swap = read.csv(q("CSF_Spectronaut_sample_swap/CSF_protein_swap_list.csv"))
)

expected_cells <- dplyr::bind_rows(
  tidyr::expand_grid(
    folder = "CSF_Spectronaut",
    dataset = c("all_data", "good_data"),
    normalization = "log2",
    truth_key = "csf"
  ),
  tidyr::expand_grid(
    folder = "CSF_Spectronaut_protein_swap",
    dataset = c("all_data", "good_data", "small", "small_good_data"),
    normalization = c("log2", "median"),
    truth_key = "protein_swap"
  ),
  tidyr::expand_grid(
    folder = "CSF_Spectronaut_sample_swap",
    dataset = c("all_data", "good_data", "small_good_data"),
    normalization = c("log2", "median"),
    truth_key = "sample_swap"
  ),
)

folder_label <- c(
  CSF_Spectronaut = "Authors' CSF sample-swap benchmark",
  CSF_Spectronaut_protein_swap = "Protein-swap benchmark",
  CSF_Spectronaut_sample_swap = "Rebuilt sample-swap benchmark"
)

model_path <- function(folder, dataset, normalization, package) {
  q(folder, dataset, normalization, "swap", package, model_filename[[package]])
}

cell_stamp <- function(folder, dataset, normalization) {
  q(folder, dataset, normalization, "swap", ".stamp")
}

diagnostics_path <- function(folder, dataset, normalization) {
  q(folder, dataset, normalization, "swap", "diagnostics.html")
}

extract_csf_dilution <- function(run_name) {
  dilution <- sub("^.*Tulum_", "", run_name)
  sub("CSF-DD.*$", "", dilution)
}

protein_swap_dataset_composition <- function() {
  dataset_order <- c("all_data", "good_data", "small", "small_good_data")
  dilution_order <- c("Neat", "1to2", "1to4", "1to8", "1to16", "1to32", "1to64")

  out <- data.table::rbindlist(lapply(dataset_order, function(dataset) {
    annotation_path <- q("CSF_Spectronaut_protein_swap", dataset, "annotation.csv")
    annotation <- data.table::fread(annotation_path)
    annotation[, Dataset := dataset]
    annotation[, Dilution := extract_csf_dilution(R.FileName)]
    counts <- annotation[, .N, by = .(Dataset, Label, Dilution, Condition)]
    data.table::dcast(
      counts,
      Dataset + Label + Dilution ~ Condition,
      value.var = "N",
      fill = 0
    )
  }), use.names = TRUE, fill = TRUE)

  if (!"Condition1" %in% names(out)) out[, Condition1 := 0L]
  if (!"Condition2" %in% names(out)) out[, Condition2 := 0L]

  out[, Total := Condition1 + Condition2]
  out[, Dataset := factor(Dataset, levels = dataset_order)]
  out[, Dilution := factor(Dilution, levels = dilution_order)]
  out[order(Dataset, Label, Dilution)]
}

model_inventory <- expected_cells |>
  tidyr::crossing(Package = packages) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    model_csv = model_path(folder, dataset, normalization, Package),
    model_exists = file.exists(model_csv),
    model_rows = {
      model <- safe_read_csv(model_csv)
      if (is.null(model)) NA_integer_ else nrow(model)
    }
  ) |>
  dplyr::ungroup()

cell_inventory <- model_inventory |>
  dplyr::summarise(
    models_present = sum(model_exists),
    models_expected = dplyr::n(),
    complete_models = models_present == models_expected,
    .by = c(folder, dataset, normalization, truth_key)
  ) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    stamp_exists = file.exists(cell_stamp(folder, dataset, normalization)),
    diagnostics_exists = file.exists(diagnostics_path(folder, dataset, normalization))
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    benchmark = unname(folder_label[folder]),
    status = dplyr::case_when(
      complete_models & stamp_exists ~ "complete",
      models_present > 0 ~ "partial",
      TRUE ~ "not yet available"
    )
  ) |>
  dplyr::arrange(folder, dataset, normalization)

tpr_ppv <- function(model_csv, truth, threshold_col = "pvalue", thr = 0.05) {
  model <- safe_read_csv(model_csv)
  if (is.null(model) || !threshold_col %in% colnames(model)) {
    return(data.frame(N = NA_integer_, TPR = NA_real_, PPV = NA_real_))
  }
  if ("Label" %in% colnames(model)) {
    model$Label <- NULL
  }
  model <- merge(model, truth, by = "Protein")
  fc_col <- if ("log2FC" %in% colnames(model)) "log2FC" else "logFC"
  if (!fc_col %in% colnames(model)) {
    return(data.frame(N = nrow(model), TPR = NA_real_, PPV = NA_real_))
  }
  finite_fc <- is.finite(model[[fc_col]])
  hit <- !is.na(model[[threshold_col]]) & model[[threshold_col]] < thr & finite_fc
  denom_pos <- sum(finite_fc & model$Label == "Positive", na.rm = TRUE)
  tpr <- if (denom_pos > 0) {
    sum(hit & model$Label == "Positive", na.rm = TRUE) / denom_pos
  } else {
    NA_real_
  }
  ppv <- if (sum(hit) > 0) {
    1 - sum(hit & model$Label == "Negative", na.rm = TRUE) / sum(hit)
  } else {
    NA_real_
  }
  data.frame(N = nrow(model), TPR = round(tpr, 3), PPV = round(ppv, 3))
}

cell_table <- function(folder, dataset, normalization, truth_key) {
  truth <- truth_sets[[truth_key]]
  rows <- lapply(packages, function(pkg) {
    f <- model_path(folder, dataset, normalization, pkg)
    raw <- tpr_ppv(f, truth, "pvalue", 0.05)
    fdr <- tpr_ppv(f, truth, "adj.pvalue", 0.05)
    data.frame(
      Package = pkg,
      N = raw$N,
      TPR_raw_p05 = raw$TPR,
      PPV_raw_p05 = raw$PPV,
      TPR_FDR_05 = fdr$TPR,
      PPV_FDR_05 = fdr$PPV,
      Model = ifelse(file.exists(f), "present", "missing")
    )
  })
  dplyr::bind_rows(rows)
}

all_results <- purrr::pmap_dfr(
  cell_inventory,
  function(folder, dataset, normalization, truth_key, models_present,
           models_expected, complete_models, stamp_exists, diagnostics_exists,
           benchmark, status) {
    tab <- cell_table(folder, dataset, normalization, truth_key)
    tab$folder <- folder
    tab$Benchmark <- benchmark
    tab$Dataset <- dataset
    tab$Normalization <- normalization
    tab$Cell_status <- status
    tab
  }
)

truth_summary <- data.frame(
  Truth_set = c(
    "Authors' CSF sample-swap list",
    "Protein-swap list",
    "Rebuilt sample-swap list"
  ),
  Rows = vapply(truth_sets, nrow, integer(1)),
  Positives = vapply(truth_sets, function(x) sum(x$Label == "Positive"), integer(1)),
  Negatives = vapply(truth_sets, function(x) sum(x$Label == "Negative"), integer(1))
)

compact_cell_inventory <- cell_inventory |>
  dplyr::transmute(
    Benchmark = benchmark,
    Dataset = dataset,
    Normalization = normalization,
    Models = paste0(models_present, "/", models_expected),
    Complete = complete_models,
    Stamp = stamp_exists,
    Diagnostics = diagnostics_exists,
    Status = status
  )

coverage_by_benchmark <- cell_inventory |>
  dplyr::summarise(
    Cells_complete = sum(status == "complete"),
    Cells_partial = sum(status == "partial"),
    Cells_missing = sum(status == "not yet available"),
    Models_present = sum(models_present),
    Models_expected = sum(models_expected),
    .by = benchmark
  ) |>
  dplyr::rename(Benchmark = benchmark)

show_cell <- function(folder, dataset, normalization) {
  all_results |>
    dplyr::filter(
      .data$folder == .env$folder,
      .data$Dataset == .env$dataset,
      .data$Normalization == .env$normalization
    ) |>
    dplyr::select(
      Package, N, TPR_raw_p05, PPV_raw_p05,
      TPR_FDR_05, PPV_FDR_05
    ) |>
    dplyr::rename(
      TPR_raw_p = TPR_raw_p05,
      PPV_raw_p = PPV_raw_p05,
      TPR_FDR = TPR_FDR_05,
      PPV_FDR = PPV_FDR_05
    )
}

bind_named_cells <- function(folder, dataset, normalizations) {
  dplyr::bind_rows(lapply(normalizations, function(norm) {
    show_cell(folder, dataset, norm) |>
      dplyr::mutate(Normalization = norm, .before = 1)
  }))
}

collect_model_pvalues <- function(folder, dataset, normalization, truth_key) {
  truth <- truth_sets[[truth_key]]
  rows <- lapply(packages, function(pkg) {
    f <- model_path(folder, dataset, normalization, pkg)
    model <- safe_read_csv(f)
    if (is.null(model) || !"pvalue" %in% names(model)) {
      return(NULL)
    }
    if ("Label" %in% names(model)) {
      model$Label <- NULL
    }
    model <- merge(model, truth, by = "Protein")
    model$Package <- pkg
    model[, c("Protein", "Package", "Label", "pvalue")]
  })
  dplyr::bind_rows(rows) |>
    dplyr::filter(is.finite(.data$pvalue))
}

plot_pvalue_histogram <- function(model_df, title) {
  ggplot2::ggplot(model_df, ggplot2::aes(x = pvalue)) +
    ggplot2::geom_histogram(
      binwidth = 0.025,
      boundary = 0,
      fill = "#3b6fb6",
      colour = "white"
    ) +
    ggplot2::geom_vline(xintercept = 0.05, linetype = 2, colour = "#b22222") +
    ggplot2::facet_wrap(~ Package, scales = "free_y") +
    ggplot2::coord_cartesian(xlim = c(0, 1)) +
    ggplot2::labs(
      title = title,
      x = "p-value",
      y = "count"
    ) +
    ggplot2::theme_bw(base_size = 10)
}

best_available <- all_results |>
  dplyr::filter(Model == "present", !is.na(TPR_FDR_05), !is.na(PPV_FDR_05)) |>
  dplyr::arrange(
    Benchmark, Dataset, Normalization,
    dplyr::desc(TPR_FDR_05), dplyr::desc(PPV_FDR_05)
  ) |>
  dplyr::summarise(
    Leading_method = dplyr::first(Package),
    TPR_FDR_05 = dplyr::first(TPR_FDR_05),
    PPV_FDR_05 = dplyr::first(PPV_FDR_05),
    .by = c(Benchmark, Dataset, Normalization)
  ) |>
  dplyr::arrange(Benchmark, Dataset, Normalization)

completed_cell_count <- sum(cell_inventory$status == "complete")
partial_cell_count <- sum(cell_inventory$status == "partial")
missing_cell_count <- sum(cell_inventory$status == "not yet available")
total_cell_count <- nrow(cell_inventory)
present_model_count <- sum(model_inventory$model_exists)
total_model_count <- nrow(model_inventory)

summarize_report_for_review <- function(report_path, annotation_path) {
  report <- data.table::fread(
    report_path,
    sep = "\t",
    select = c("R.FileName", "PG.ProteinGroups", "EG.PrecursorId", "FG.Quantity")
  )
  report <- report[
    nzchar(PG.ProteinGroups) & is.finite(FG.Quantity) & FG.Quantity > 0
  ]
  report <- unique(report, by = c("R.FileName", "PG.ProteinGroups", "EG.PrecursorId"))
  summary <- report[
    ,
    .(LogIntensity = mean(log2(FG.Quantity), na.rm = TRUE)),
    by = .(R.FileName, Protein = PG.ProteinGroups)
  ]
  annotation <- data.table::fread(annotation_path)
  annotation <- annotation[, .(R.FileName, Condition)]
  merge(summary, annotation, by = "R.FileName", all.x = TRUE)
}

variance_stats_for_review <- function(report_path, annotation_path, truth) {
  summary <- summarize_report_for_review(report_path, annotation_path)
  truth <- data.table::as.data.table(truth)[, .(Protein, Label)]
  summary <- merge(summary, truth, by = "Protein")
  conditions <- sort(unique(summary$Condition))
  if (length(conditions) < 2) {
    return(data.table::data.table())
  }
  summary <- summary[Condition %in% conditions[1:2]]
  stats <- summary[
    ,
    .(
      mean_log = mean(LogIntensity, na.rm = TRUE),
      sd_log = stats::sd(LogIntensity, na.rm = TRUE),
      n = .N
    ),
    by = .(Protein, Label, Condition)
  ]
  wide <- data.table::dcast(
    stats,
    Protein + Label ~ Condition,
    value.var = c("mean_log", "sd_log", "n")
  )
  c1 <- conditions[[1]]
  c2 <- conditions[[2]]
  wide[
    ,
    mean_diff := get(paste0("mean_log_", c1)) - get(paste0("mean_log_", c2))
  ]
  wide[
    ,
    mean_intensity := 0.5 * (get(paste0("mean_log_", c1)) + get(paste0("mean_log_", c2)))
  ]
  wide[
    ,
    sd_pool := sqrt(
      ((get(paste0("n_", c1)) - 1) * get(paste0("sd_log_", c1))^2 +
         (get(paste0("n_", c2)) - 1) * get(paste0("sd_log_", c2))^2) /
        pmax(get(paste0("n_", c1)) + get(paste0("n_", c2)) - 2, 1)
    )
  ]
  wide[is.finite(mean_diff) & is.finite(sd_pool)]
}

condition_sd_stats_for_review <- function(report_path, annotation_path, truth) {
  summary <- summarize_report_for_review(report_path, annotation_path)
  truth <- data.table::as.data.table(truth)[, .(Protein, Label)]
  summary <- merge(summary, truth, by = "Protein")
  conditions <- sort(unique(summary$Condition))
  if (length(conditions) < 2) {
    return(data.table::data.table())
  }
  summary <- summary[Condition %in% conditions[1:2]]
  out <- summary[
    ,
    .(
      mean_log = mean(LogIntensity, na.rm = TRUE),
      sd_log = stats::sd(LogIntensity, na.rm = TRUE),
      n = .N
    ),
    by = .(Protein, Label, Condition)
  ]
  out[is.finite(sd_log) & n >= 2]
}

sample_swap_before_after_sd <- function(before_report_path, before_annotation_path,
                                        after_report_path, after_annotation_path,
                                        truth) {
  before <- variance_stats_for_review(before_report_path, before_annotation_path, truth)
  before$state <- "before swap"
  after <- variance_stats_for_review(after_report_path, after_annotation_path, truth)
  after$state <- "after swap"
  out <- data.table::rbindlist(list(before, after), use.names = TRUE, fill = TRUE)
  out$state <- factor(out$state, levels = c("before swap", "after swap"))
  out$Label <- factor(out$Label, levels = c("Negative", "Positive"))
  out
}

plot_sample_swap_sd <- function(stats) {
  ggplot2::ggplot(stats, ggplot2::aes(x = sd_pool, fill = Label)) +
    ggplot2::geom_density(alpha = 0.65, colour = "black", linewidth = 0.45) +
    ggplot2::facet_wrap(~ state, nrow = 1) +
    ggplot2::scale_fill_manual(
      values = c(Negative = "#9f9f9f", Positive = "#74b96e"),
      breaks = c("Negative", "Positive")
    ) +
    ggplot2::labs(
      title = "Pooled within-condition SD",
      x = "pooled sd(log2 intensity)",
      y = "density",
      fill = NULL
    ) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(
      legend.position = "top",
      legend.key.size = grid::unit(0.7, "cm"),
      panel.grid.minor = ggplot2::element_line(colour = "grey92"),
      panel.grid.major = ggplot2::element_line(colour = "grey88"),
      strip.text = ggplot2::element_text(size = 11),
      plot.title = ggplot2::element_text(size = 15)
    )
}

plot_sample_swap_mean_variance <- function(stats) {
  stats <- stats[stats$state == "after swap" & is.finite(stats$mean_intensity), ]
  stats$Label <- factor(stats$Label, levels = c("Negative", "Positive"))
  ggplot2::ggplot(
    stats,
    ggplot2::aes(x = mean_intensity, y = sd_pool, colour = Label)
  ) +
    ggplot2::geom_point(alpha = 0.45, size = 0.8) +
    ggplot2::geom_smooth(se = FALSE, method = "loess", linewidth = 1.1) +
    ggplot2::scale_colour_manual(
      values = c(Negative = "#6f6f6f", Positive = "#1f9a2b"),
      breaks = c("Negative", "Positive")
    ) +
    ggplot2::labs(
      title = "Mean-variance relationship after swap",
      subtitle = "limma's eBayes assumes a single trend; the truth here is bimodal",
      x = "mean log2 intensity",
      y = "pooled within-condition SD",
      colour = NULL
    ) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(
      legend.position = "top",
      panel.grid.minor = ggplot2::element_line(colour = "grey92"),
      panel.grid.major = ggplot2::element_line(colour = "grey88"),
      plot.title = ggplot2::element_text(size = 15)
    )
}

swap_effect_and_sd <- function(rda_path, truth_df) {
  if (!file.exists(rda_path)) {
    return(list(effect = NULL, sd = NULL, mv = NULL))
  }
  e <- new.env()
  load(rda_path, envir = e)
  obj <- get(ls(e)[1], envir = e)
  pld <- data.table::as.data.table(obj$ProteinLevelData)
  truth_dt <- data.table::as.data.table(truth_df)
  truth_dt[, Protein := as.character(Protein)]
  pld[, Protein := as.character(Protein)]
  pld[, class := truth_dt[match(pld$Protein, Protein), Label]]
  pld <- pld[!is.na(class) & is.finite(LogIntensities)]

  stats <- pld[
    ,
    .(
      mean_log = mean(LogIntensities, na.rm = TRUE),
      sd_log = stats::sd(LogIntensities, na.rm = TRUE),
      n = .N
    ),
    by = .(Protein, GROUP, class)
  ]
  wide <- data.table::dcast(
    stats,
    Protein + class ~ GROUP,
    value.var = c("mean_log", "sd_log", "n")
  )
  wide[, mean_diff := mean_log_Condition1 - mean_log_Condition2]
  wide[
    ,
    sd_pool := sqrt(
      ((n_Condition1 - 1) * sd_log_Condition1^2 +
         (n_Condition2 - 1) * sd_log_Condition2^2) /
        pmax(n_Condition1 + n_Condition2 - 2, 1)
    )
  ]
  wide[, mean_intensity := 0.5 * (mean_log_Condition1 + mean_log_Condition2)]

  pal <- c(Positive = "#2ca02c", Negative = "#7f7f7f")
  p_eff <- ggplot2::ggplot(
    wide[is.finite(mean_diff)],
    ggplot2::aes(x = mean_diff, fill = class)
  ) +
    ggplot2::geom_density(alpha = 0.55) +
    ggplot2::geom_vline(xintercept = 0, linetype = 2, colour = "grey40") +
    ggplot2::scale_fill_manual(values = pal) +
    ggplot2::labs(
      title = "Per-protein effect size",
      x = "log2(Condition1) - log2(Condition2)",
      y = "density",
      fill = NULL
    ) +
    ggplot2::theme_bw() +
    ggplot2::theme(legend.position = "top")

  p_sd <- ggplot2::ggplot(
    wide[is.finite(sd_pool)],
    ggplot2::aes(x = sd_pool, fill = class)
  ) +
    ggplot2::geom_density(alpha = 0.55) +
    ggplot2::scale_fill_manual(values = pal) +
    ggplot2::labs(
      title = "Pooled within-condition SD",
      x = "pooled sd(log2 intensity)",
      y = "density",
      fill = NULL
    ) +
    ggplot2::theme_bw() +
    ggplot2::theme(legend.position = "top")

  p_mv <- ggplot2::ggplot(
    wide[is.finite(sd_pool)],
    ggplot2::aes(x = mean_intensity, y = sd_pool, colour = class)
  ) +
    ggplot2::geom_point(alpha = 0.4, size = 0.7) +
    ggplot2::geom_smooth(se = FALSE, method = "loess") +
    ggplot2::scale_colour_manual(values = pal) +
    ggplot2::labs(
      title = "Mean-variance relationship",
      subtitle = "eBayes assumes a single trend; the truth here is bimodal",
      x = "mean log2 intensity",
      y = "pooled within-condition SD",
      colour = NULL
    ) +
    ggplot2::theme_bw() +
    ggplot2::theme(legend.position = "top")

  list(effect = p_eff, sd = p_sd, mv = p_mv)
}

plot_effect_variance <- function(stats, condition_sd, title_prefix) {
  effect_plot <- ggplot2::ggplot(
    stats,
    ggplot2::aes(x = mean_diff, fill = Label)
  ) +
    ggplot2::geom_density(alpha = 0.55) +
    ggplot2::geom_vline(xintercept = 0, linetype = 2, colour = "grey40") +
    ggplot2::scale_fill_manual(values = c(Positive = "#2ca02c", Negative = "#7f7f7f")) +
    ggplot2::labs(
      title = paste(title_prefix, "effect size"),
      x = "mean log2 difference between conditions",
      y = "density",
      fill = NULL
    ) +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(legend.position = "top")

  variance_plot <- ggplot2::ggplot(
    condition_sd[is.finite(sd_log)],
    ggplot2::aes(x = sd_log, fill = Label)
  ) +
    ggplot2::geom_density(alpha = 0.55) +
    ggplot2::scale_fill_manual(values = c(Positive = "#2ca02c", Negative = "#7f7f7f")) +
    ggplot2::labs(
      title = paste(title_prefix, "within-condition SD"),
      x = "sd(log2 intensity) within one condition",
      y = "density",
      fill = NULL
    ) +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(legend.position = "top")

  list(effect = effect_plot, variance = variance_plot)
}

authors_csf_sample_swap_stats <- function() {
  annotation <- data.table::fread(q("CSF_Spectronaut/annotation.csv"))
  swap_list <- data.table::fread(q("CSF_Spectronaut/CSF_protein_swap_list.csv"))

  e <- new.env()
  load(q("CSF_Spectronaut/good_data/log2/swap/MSstats/MSstats_summarized.rda"), envir = e)
  prot <- data.table::as.data.table(e$base_msstats_summarized$ProteinLevelData)
  prot[, Protein := as.character(Protein)]
  prot[, Order := as.integer(sub(".*Seq", "", originalRUN))]

  true_pos <- swap_list[Label == "Positive", Protein]
  true_neg <- swap_list[Label == "Negative", Protein]
  good_runs <- annotation[Label == "Good", R.FileName]

  prot_before <- data.table::copy(prot)
  prot_before[, state := "before swap"]

  prot_after <- swap_condition_labels(prot, true_pos, true_neg) |>
    data.table::as.data.table()
  prot_after[, state := "after swap"]

  prot_all <- data.table::rbindlist(
    list(prot_before, prot_after),
    use.names = TRUE,
    fill = TRUE
  )
  prot_all <- prot_all[originalRUN %in% good_runs]
  prot_all[, class := data.table::fifelse(
    Protein %in% true_pos,
    "Positive",
    data.table::fifelse(Protein %in% true_neg, "Negative", NA_character_)
  )]
  prot_all <- prot_all[!is.na(class) & is.finite(LogIntensities)]
  prot_all[, state := factor(state, levels = c("before swap", "after swap"))]
  prot_all[, class := factor(class, levels = c("Negative", "Positive"))]

  stats <- prot_all[
    ,
    .(
      mean_log = mean(LogIntensities, na.rm = TRUE),
      sd_log = stats::sd(LogIntensities, na.rm = TRUE),
      n = .N
    ),
    by = .(Protein, GROUP, state, class)
  ]

  wide <- data.table::dcast(
    stats,
    Protein + state + class ~ GROUP,
    value.var = c("mean_log", "sd_log", "n")
  )
  wide[, mean_diff := mean_log_Condition1 - mean_log_Condition2]
  wide[
    ,
    sd_pool := sqrt(
      ((n_Condition1 - 1) * sd_log_Condition1^2 +
         (n_Condition2 - 1) * sd_log_Condition2^2) /
        pmax(n_Condition1 + n_Condition2 - 2, 1)
    )
  ]
  wide[, mean_intensity := 0.5 * (mean_log_Condition1 + mean_log_Condition2)]
  condition_sd <- stats[is.finite(sd_log) & n >= 2]
  list(wide = wide, condition_sd = condition_sd)
}

plot_csf_sample_swap_effect <- function(wide) {
  ggplot2::ggplot(
    wide[is.finite(mean_diff)],
    ggplot2::aes(x = mean_diff, fill = class)
  ) +
    ggplot2::geom_density(alpha = 0.55) +
    ggplot2::geom_vline(xintercept = 0, linetype = 2, colour = "grey40") +
    ggplot2::facet_wrap(~ state) +
    ggplot2::scale_fill_manual(values = c(Positive = "#2ca02c", Negative = "#7f7f7f")) +
    ggplot2::labs(
      title = "Per-protein effect size",
      x = "log2 fold-change (Condition1 - Condition2)",
      y = "density",
      fill = NULL
    ) +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(legend.position = "top")
}

plot_csf_sample_swap_sd <- function(condition_sd) {
  ggplot2::ggplot(
    condition_sd[is.finite(sd_log)],
    ggplot2::aes(x = sd_log, fill = class)
  ) +
    ggplot2::geom_density(alpha = 0.55, colour = "black", linewidth = 0.35) +
    ggplot2::facet_wrap(~ state) +
    ggplot2::scale_fill_manual(values = c(Positive = "#2ca02c", Negative = "#7f7f7f")) +
    ggplot2::labs(
      title = "Within-condition SD",
      x = "sd(log2 intensity) within one condition",
      y = "density",
      fill = NULL
    ) +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(legend.position = "top")
}

plot_csf_sample_swap_mean_variance <- function(condition_sd) {
  mv <- condition_sd[is.finite(mean_log) & is.finite(sd_log) & state == "after swap"]
  ggplot2::ggplot(
    mv,
    ggplot2::aes(x = mean_log, y = sd_log, colour = class)
  ) +
    ggplot2::geom_point(alpha = 0.4, size = 0.7) +
    ggplot2::geom_smooth(se = FALSE, method = "loess") +
    ggplot2::scale_colour_manual(values = c(Positive = "#2ca02c", Negative = "#7f7f7f")) +
    ggplot2::labs(
      title = "Mean-variance relationship after swap",
      subtitle = "The authors' benchmark creates separate H1 and H0 variance behaviour",
      x = "mean log2 intensity",
      y = "within-condition SD",
      colour = NULL
    ) +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(legend.position = "top")
}
```

# Supplement Scope

This supplement provides the technical material supporting the short main
review. It contains the reproduced CSF Spectronaut benchmark tables, p-value
distribution diagnostics, per-protein effect-size and within-condition variance
checks, rebuilt sample-swap analyses, protein-swap sensitivity analyses, and an
author-facing list of points to address.

The rebuilt sample-swap and protein-swap analyses are not proposed as
definitive replacement benchmarks. They are sensitivity checks for the main
review concern: the manuscript's K562/CSF sample-swap benchmark can couple
ground-truth class to within-condition variance, which complicates the
interpretation of method rankings in Table 1.

The central critique is not that MSstats+ is implemented incorrectly. The
central critique is that the sample-swap construction can induce a
label-dependent variance structure. The CSF benchmark starts from a dilution
experiment where the Neat versus 1:2 comparison should make almost every
quantified protein differentially abundant. There is therefore no natural null
class. The benchmark creates a null by swapping run labels for most proteins and
leaving a minority unchanged as Positives. This gives a useful synthetic
stress-test benchmark, but it also links the ground-truth label to
within-condition variance: Positives remain relatively clean dilution
comparisons, while Negatives become mixtures of dilution levels inside each
condition.

This matters for methods that use empirical-Bayes or related variance
moderation. In the sample-swap benchmark, the shared variance prior is estimated
from an imbalanced mixture dominated by Negative proteins. If the Negative
class has higher within-condition variance because of the swap construction,
moderation can pull the Positive variances upward and reduce sensitivity. This
supports the main review conclusion: Table 1 should be interpreted as evidence
for performance on this specific sample-swap construction, not as a standalone
general ranking of differential-abundance workflows.

This supplement reports two operating points, both at threshold 0.05: raw
p-value and adjusted p-value. TPR is the fraction of truth-set Positives
detected. PPV is one minus the false-positive fraction among called proteins.
Missing model outputs appear as missing values in the result tables.

The review code is available at
<https://github.com/wolski/msstatsplus-quant-review>. See the repository
[`README.md`](https://github.com/wolski/msstatsplus-quant-review/blob/main/README.md)
for the project layout, including shared R code in `R/`, helper scripts in
`src/`, and review sources in `vignettes/`.

The GitHub repository contains code and review sources, not the raw benchmark
data. The input data are from the public MSstats+ archive: ProteomeXchange
[`PXD066486`](https://proteomecentral.proteomexchange.org/cgi/GetDataset?ID=PXD066486)
and MassIVE
[`MSV000098622`](https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=fdf984642108451cbdece7303e47f2d2),
with files available from the MassIVE FTP archive
`ftp://massive-ftp.ucsd.edu/v10/MSV000098622/`.

# Benchmark Data And Truth Sets

The authors' CSF Spectronaut benchmark uses a 1,820-protein truth list: 182
Positives and 1,638 Negatives. In the re-run workflow, I also created two
checks that use the same underlying CSF Spectronaut report but different
synthetic truth definitions.

First, the rebuilt sample-swap benchmark applies the sample-swap idea in a
single generated report rather than inside individual modelling scripts.
Second, the protein-swap benchmark swaps precursor intensities between matched
protein pairs instead of swapping sample labels. Both rebuilt benchmarks now use
the same conservative universe: proteins in non-blank runs with at least two
unique precursors. The original report contains 3,041 non-blank proteins with at
least one precursor; applying `n_precursors >= 2` leaves 2,244 proteins. The
excluded 797 proteins are single-precursor proteins. Keeping this filter is a
design choice: single-precursor proteins are less stable as benchmark Negatives,
and they cannot participate in the matched protein-pair construction.

The rebuilt sample-swap truth list therefore contains 2,244 proteins: 224
Positives and 2,020 Negatives. The protein-swap truth list contains the same
2,244-protein universe, with 224 Positives arranged as 112 matched protein
pairs and 2,020 Negatives.

```{r truth-summary}
knitr::kable(
  truth_summary,
  caption = "Truth sets used by this review.",
  row.names = FALSE
)
```

# Authors' CSF Spectronaut Benchmark

The authors' CSF Spectronaut analysis has both expected log2 cells available:
`all_data` and `good_data`. These are the closest reproduction of the
manuscript-style CSF sample-swap benchmark in this report.

The `good_data` subset is especially important because it restricts the
comparison to the cleaner Neat and 1:2 runs. If the benchmark were a neutral
standard differential-abundance problem, the non-MSstats workflows should behave
similarly across clean benchmark constructions. Instead, the sample-swap
construction remains important because the Negative class has been made by
mixing dilution levels inside condition.

## CSF all_data, log2

```{r csf-all-data-log2}
knitr::kable(
  show_cell("CSF_Spectronaut", "all_data", "log2"),
  caption = "Authors' CSF sample-swap benchmark: all_data/log2."
)
```

Table 2 is the manuscript Table 1 replication check for the CSF Spectronaut
sample-swap benchmark. The manuscript-comparable columns are the raw p-value
columns (`TPR_raw_p` and `PPV_raw_p`), evaluated at threshold 0.05 because that
is the operating point used in the original benchmark summary. At this
operating point, the rerun agrees with the manuscript for the reproduced
authors' pipeline rows: MSstats, MaxLFQ+limma, DEqMS, and limpa. MSstats+ is
also very close to the manuscript value, with only a small numerical difference
in this current render.

The `msqrob2` row should not be interpreted as a failed replication. In this
rerun we deliberately use a different, faster `msqrob2` pipeline than the one
used for the manuscript/original archive. A difference in the `msqrob2` TPR is
therefore expected and is not evidence that the CSF sample-swap replication
failed. The FDR columns are additional review diagnostics and are not the
manuscript Table 1 operating point.

The p-value distributions for Table 2 are shown below. The dashed vertical line
marks the 0.05 threshold used for the raw p-value columns.

```{r pvalue-hist-csf-all, fig.cap="P-value distributions for Table 2: authors' CSF Spectronaut all_data/log2 benchmark. The dashed red line marks p = 0.05.", fig.width=9, fig.height=6}
collect_model_pvalues(
  "CSF_Spectronaut",
  "all_data",
  "log2",
  "csf"
) |>
  plot_pvalue_histogram("CSF Spectronaut all_data/log2 p-value distributions")
```

Figure 1 is an additional diagnostic for benchmark calibration. Because the
authors' sample-swap truth list is dominated by Negatives, the overall p-value
histograms are driven mainly by proteins intended to behave as null proteins.
Under a well-calibrated null, those H0 p-values should be approximately uniform
on `[0, 1]`. Instead, the CSF all_data/log2 distributions are strongly
non-uniform across methods. Most panels show substantial mass near one, while
several also show method-dependent spikes near zero.

This pattern should be interpreted cautiously. A non-uniform p-value histogram
does not by itself identify the cause, and the all_data benchmark intentionally
contains poor-quality high-dilution runs. However, the pattern is anomalous for
a clean null benchmark and is consistent with the broader concern that the
sample-swap construction changes more than the null mean. It motivates the
variance diagnostics below, where the same benchmark is shown to couple the
truth label to within-condition variance.

## CSF good_data, log2

Here `good_data` means the authors' CSF Spectronaut benchmark restricted to the
Good runs from the canonical Neat versus 1:2 comparison, with balanced
replicate counts. It excludes the higher-dilution Bad runs present in
`all_data`. The sample-swap truth definition is otherwise unchanged: Negative
proteins are still created by the authors' in-script sample-label swap, while
Positive proteins retain the original dilution comparison.

```{r csf-good-data-log2}
knitr::kable(
  show_cell("CSF_Spectronaut", "good_data", "log2"),
  caption = "Authors' CSF sample-swap benchmark: good_data/log2."
)
```

# Per-protein Effect Size And Within-condition Variance

The sample-swap operation solves one problem and creates another. It solves the
absence of a null class by exchanging Negative-protein measurements between
paired Condition1 and Condition2 runs. At the condition level, each Negative
protein then has approximately the same mean in both groups. The Positive
proteins are not swapped and keep the original dilution effect.

The variance-shrinkage story sits here. Under H0, Negative proteins should have
within-condition variance approximately equal to Positive proteins because both
classes are technical dilutions of the same material. The swap induces a
bimodal variance distribution tied to the ground-truth label:

- **Positive proteins** retain their natural per-run variance because no runs
  were rewritten for them.
- **Negative proteins** become 50/50 mixtures of Neat and 1:2 runs inside each
  condition, so their within-condition variance grows by the mixture term
  `0.5 * 0.5 * 1^2` in log2 units.

The following figures are generated from the authors' CSF Spectronaut benchmark
implementation, not from our rebuilt explicit sample-swap file. They use the
MSstats protein-level summary object
`CSF_Spectronaut/good_data/log2/swap/MSstats/MSstats_summarized.rda`; the plotted
variable is `LogIntensities`, already on the log2 scale. The after-swap state is
created in memory with the same `swap_condition_labels()` operation used by the
authors' MSstats benchmark code. This object matches the authors' benchmark
code path and is therefore the relevant object for evaluating the Table 1 CSF
result.

```{r authors-csf-sample-swap-stats, include=FALSE}
authors_csf <- authors_csf_sample_swap_stats()
authors_csf_wide <- authors_csf$wide
authors_csf_condition_sd <- authors_csf$condition_sd
```

```{r sample-swap-effect, fig.cap="Authors' CSF sample-swap benchmark, good_data/log2. Per-protein effect size is the mean MSstats protein-level LogIntensities in Condition1 minus Condition2. Before the swap, both classes carry the dilution effect; after the authors' in-memory sample swap, Negative proteins are centered near zero while Positive proteins retain the dilution effect.", fig.width=8, fig.height=4}
plot_csf_sample_swap_effect(authors_csf_wide)
```

```{r sample-swap-sd, fig.cap="Authors' CSF sample-swap benchmark, good_data/log2. For each protein and condition, SD is computed across runs from the same MSstats protein-level LogIntensities. The density shows these condition-specific SD values by class. After the swap, the Negative class shifts to higher within-condition SD because each condition contains a mixture of Neat and 1:2 measurements for Negative proteins; Positive proteins are not swapped.", fig.width=8, fig.height=4}
plot_csf_sample_swap_sd(authors_csf_condition_sd)
```

```{r sample-swap-mv, fig.cap="Authors' CSF sample-swap benchmark, good_data/log2. Each point is one protein within one condition after the in-memory sample swap. The x-axis is the condition-specific mean MSstats protein-level LogIntensity and the y-axis is the condition-specific SD across runs. No pooled SD is used. The two trends show the variance structure that can disadvantage empirical-Bayes methods using one shared moderation trend.", fig.width=8, fig.height=5}
plot_csf_sample_swap_mean_variance(authors_csf_condition_sd)
```

The mixture-induced variance asymmetry provides a mechanism by which moderated
methods (limma, DEqMS, msqrob2, limpa) can shrink **Positive** proteins more
than Negative ones. The empirical-Bayes prior can be influenced by the
Negative-dominated variance regime and can sit farther above the Positive
variance than below the Negative variance. The expected consequence is reduced
sensitivity for the Positive class, rather than inflated specificity.

**Sample-size dependence.** The strength of the moderation prior depends on
degrees of freedom. With many replicates, the per-protein variance estimate
dominates the prior; with few replicates, the prior dominates. A 4-vs-4 design
therefore puts the prior at much higher relative weight than a larger balanced
design, so `small_good_data` accentuates this construction-dependent effect.

# Rebuilt Sample-Swap Results

The rebuilt sample-swap benchmark uses the same general sample-swap idea as the
authors' CSF benchmark, but makes the swapped report explicit on disk and
applies the same `n_precursors >= 2` universe used by protein-swap. This gives
224 Positives and 2,020 Negatives.

Here I show only the log2 `good_data` and `small_good_data` cells. The point of
this comparison is sample size.

## Sample-swap good_data

```{r sample-swap-good-data}
show_cell(
  "CSF_Spectronaut_sample_swap",
  "good_data",
  "log2"
) |>
  knitr::kable(caption = "Rebuilt sample-swap benchmark: good_data.")
```

## Sample-swap small_good_data

```{r sample-swap-small-good-data}
show_cell(
  "CSF_Spectronaut_sample_swap",
  "small_good_data",
  "log2"
) |>
  knitr::kable(caption = "Rebuilt sample-swap benchmark: small_good_data.")
```

The expected sample-size effect is visible in Tables 4 and 5. With fewer
samples, per-protein variance estimates carry less information and the
model/prior assumptions have greater influence. Under the sample-swap
construction, where the null and positive classes have different variance
regimes, this primarily reduces sensitivity for Positive proteins rather than
creating many extra false positives.

The clearest comparison is the adjusted-p-value operating point. From
`good_data` to `small_good_data`, FDR TPR drops from 0.899 to 0.550 for
MSstats, from 0.899 to 0.500 for MSstats+, from 0.890 to 0.009 for
MaxLFQ+limma, from 0.844 to 0.482 for prolfqua, and from 0.904 to 0.005 for
limpa. PPV remains high for most of these rows. Thus the smaller design mainly
makes the workflows more conservative for the Positive class, and the effect is
particularly severe for variance-moderated workflows such as MaxLFQ+limma and
limpa.

# Protein-Swap Benchmark And Variance Check

The protein-swap benchmark tests a different construction: rather than swapping
sample labels, it swaps precursor intensities between matched protein pairs.
This design is intended to reduce dependence on the variance asymmetry induced
by sample-label swapping in a dilution series.

The protein-swap truth list is generated after filtering to proteins with at
least two precursors. This gives a 2,244-protein universe with 224 Positives
organized as 112 matched protein pairs and 2,020 Negatives. The older 3,041-row
truth-list expectation included single-precursor proteins. I treat the current
2,244-row list as the cleaner benchmark definition because single-precursor
proteins cannot participate in the paired swap and are less reliable as
benchmark Negatives.

The same diagnostic is shown below for the protein-swap `good_data` report.
Here the data are read directly from
`CSF_Spectronaut_protein_swap/good_data/Report.tsv`: precursor-level
`FG.Quantity` values are log2-transformed, deduplicated at
`(run, protein, precursor)`, and averaged to one protein/run value before fold
changes and within-condition SDs are computed. No MSstats RDA object is used
for this protein-swap diagnostic. The density plot shows condition-specific
SDs, one value for each protein in each condition. The mean-variance plot below
uses a pooled within-condition SD only because that display requires one y-value
per protein.

The important point is directional. In a real treatment-control comparison,
regulated proteins can be more variable because they are the proteins that
respond to treatment, and that response can add biological heterogeneity on top
of technical variation. Therefore it is not suspicious if the Positive density
is broader or shifted to higher within-condition SD than the Negative density in the
protein-swap benchmark. What is suspicious in the authors' sample-swap
benchmark is the opposite construction: the null/Negative class is made more
variable by mixing Neat and 1:2 samples inside each condition.

```{r protein-swap-effect-variance, fig.cap="Explicit protein-swap benchmark, good_data. Intensities are direct report-derived protein/run summaries: log2 FG.Quantity values averaged from precursor to protein level within each run. Positives are matched protein-pair members; Negatives remain near the null. Left: per-protein effect size. Right: condition-specific within-condition SD (one density per protein per condition) — computed via the shared helper `protein_swap_sd_density_from_report()` in `R/figures.R`, the same one used by `review.qmd`'s `protein-swap-sd` figure, so the two vignettes show the byte-identical SD density. NOTE: the broader Positive SD distribution visible here is **predominantly a swap-pair selection effect** (n_precursors median 3 vs 9 for Negatives, exact-match constraint in the pair-builder), not an effect of the rank-aligned exchange itself — see `swap_pairs_before_after.qmd` for the empirical pre vs post G2-only check.", fig.width=10, fig.height=4}
protein_swap_stats <- variance_stats_for_review(
  q("CSF_Spectronaut_protein_swap/good_data/Report.tsv"),
  q("CSF_Spectronaut_protein_swap/good_data/annotation.csv"),
  truth_sets$protein_swap
)
protein_swap_plots <- plot_effect_variance(
  protein_swap_stats,
  # Empty condition_sd: the variance panel comes from the shared helper below;
  # plot_effect_variance ignores its variance side when fed an empty data.table.
  data.table::data.table(Protein = character(), Label = character(),
                         sd_log = numeric()),
  "Protein-swap good_data"
)
protein_swap_variance <- protein_swap_sd_density_from_report(
  report_path     = q("CSF_Spectronaut_protein_swap/good_data/Report.tsv"),
  annotation_path = q("CSF_Spectronaut_protein_swap/good_data/annotation.csv"),
  swap_list_path  = q("CSF_Spectronaut_protein_swap/CSF_protein_swap_list.csv"),
  title_prefix    = "Protein-swap good_data"
)
protein_swap_plots$effect + protein_swap_variance
```

```{r protein-swap-mean-variance, fig.cap="Explicit protein-swap benchmark, good_data. Each point is one protein summarized from the direct report-derived protein/run matrix. The y-axis is the pooled within-condition SD, used here only to place one variance summary per protein. Unlike the authors' sample-swap benchmark, the null/Negative class is not forced into a higher-variance dilution mixture.", fig.width=8, fig.height=5}
protein_swap_mv <- protein_swap_stats[is.finite(protein_swap_stats$sd_pool), ]
ggplot2::ggplot(
  protein_swap_mv,
  ggplot2::aes(x = mean_intensity, y = sd_pool, colour = Label)
) +
  ggplot2::geom_point(alpha = 0.4, size = 0.7) +
  ggplot2::geom_smooth(se = FALSE, method = "loess") +
  ggplot2::scale_colour_manual(values = c(Positive = "#2ca02c", Negative = "#7f7f7f")) +
  ggplot2::labs(
    title = "Mean-variance relationship in the protein-swap benchmark",
    subtitle = "Direct report-derived protein/run summaries from good_data",
    x = "mean log2 intensity",
    y = "pooled within-condition SD",
    colour = NULL
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

## Protein-swap Dataset Composition

The protein-swap dataset names refer to run subsets, not to different
protein-truth definitions. The same conservative 2,244-protein truth universe
is used throughout. What changes between `all_data`, `good_data`, `small`, and
`small_good_data` is which CSF injections enter the modelling step.

In the protein-swap benchmark, the Positive and Negative classes are defined by
the protein-pair swap, not by the sample dilution. The table below describes
the original run composition of each subset. `Condition1` and `Condition2` are
the two model groups; `Good` and `Bad` are the sample-quality labels inherited
from the original CSF annotation.

- `all_data` contains the full available CSF design: the Good Neat and 1:2
  runs plus the higher-dilution Bad runs.
- `good_data` contains only Good Neat runs, so any differential signal comes
  from the protein swap rather than from the dilution design.
- `small` is a smaller mixed subset containing Good Neat/1:2 runs and a few Bad
  higher-dilution runs.
- `small_good_data` is the small Neat-only subset, four runs in each model
  group.

```{r protein-swap-dataset-composition}
protein_swap_dataset_composition() |>
  dplyr::mutate(
    Dataset = as.character(Dataset),
    Dilution = as.character(Dilution)
  ) |>
  dplyr::rename(
    `Sample label` = Label,
    `Group 1 / Condition1` = Condition1,
    `Group 2 / Condition2` = Condition2
  ) |>
  knitr::kable(
    caption = "Run composition of the CSF_Spectronaut_protein_swap dataset subsets."
  )
```

## Protein-swap all_data, log2

```{r protein-swap-all-data-log2}
show_cell(
  "CSF_Spectronaut_protein_swap",
  "all_data",
  "log2"
) |>
  knitr::kable(caption = "Protein-swap benchmark: all_data/log2.")
```

## Protein-swap all_data, median

```{r protein-swap-all-data-median}
show_cell(
  "CSF_Spectronaut_protein_swap",
  "all_data",
  "median"
) |>
  knitr::kable(caption = "Protein-swap benchmark: all_data/median.")
```

## Protein-swap small, log2

```{r protein-swap-small-log2}
show_cell(
  "CSF_Spectronaut_protein_swap",
  "small",
  "log2"
) |>
  knitr::kable(caption = "Protein-swap benchmark: small/log2.")
```

## Protein-swap small, median

```{r protein-swap-small-median}
show_cell(
  "CSF_Spectronaut_protein_swap",
  "small",
  "median"
) |>
  knitr::kable(caption = "Protein-swap benchmark: small/median.")
```

## Interpretation Of The Protein-Swap Results

The available protein-swap results support a narrower interpretation of the
benchmark. In this
construction, the null/Negative proteins are not made artificially more
variable by mixing dilution levels inside a condition. Once that sample-swap
variance asymmetry is removed, methods that use variance shrinkage or p-value
moderation are no longer systematically disadvantaged. They are often
comparable to, and in several cells better than, the MSstats+ row.

For `all_data/log2`, MSstats+ has high raw-p sensitivity, but it is not the
only strong method: MaxLFQ+limma, msqrob2, and prolfqua are close at the raw
p-value threshold, and msqrob2 has the highest FDR sensitivity. For
`all_data/median`, the moderated methods are even stronger: MaxLFQ+limma,
msqrob2, and limpa have higher FDR TPR than MSstats+, while retaining high PPV.
This is the opposite of the story suggested by the authors' CSF sample-swap
benchmark, where moderation can be disadvantaged by the design.

The smaller `small` subset should be read more carefully. With nominal `log2`
normalization, all methods lose sensitivity because the sample size is small,
but the `msqrob2` row is not a clean log2-only comparison: the current msqrob2
pipeline applies additional median centering internally. I therefore do not use
that row as evidence for a pure log2 advantage. With explicit `median`
normalization, MaxLFQ+limma, msqrob2, prolfqua, and limpa all exceed MSstats+
in raw-p TPR, and the moderated methods also dominate the non-zero FDR
detections. This supports the central point of the reanalysis: the CSF Table 1
advantage alone does not establish a general property of the modelling problem.
It depends strongly on a benchmark construction that couples the ground-truth
label to different variance regimes.

# Points To Raise With The Authors

The following points are intended as inputs for the article and accompanying
materials in `manuscript_materials_second_submission`, not as private notes on
this reanalysis. Items 1-5 are core asks tied to the variance-coupling concern
and should be addressed for revision. Items 6-9 are constructive suggestions
for the next version.

## Core asks (variance-coupling concern)

1. **Protein universe definition.** The manuscript-style CSF benchmark uses an
   1,820-protein truth list. The rebuilt sample-swap and protein-swap workflows
   use a stricter 2,244-protein universe after the `n_precursors >= 2` filter.
   The manuscript should state the filtering rules for each benchmark and
   justify why each universe is appropriate for comparing methods.
2. **Thresholding convention.** The conclusions are sensitive to whether the
   table is read at raw `p < 0.05` or adjusted `p < 0.05`. The manuscript should
   make the thresholding convention explicit wherever benchmark tables are
   interpreted.
3. **Dependence on swap construction.** The sample-swap design creates
   label-linked variance asymmetry in CSF. I view this as a firm benchmark-design
   concern for interpreting Table 1, not merely as a minor implementation
   detail. The authors should clarify whether MSstats+ is expected to improve
   specifically under this sample-swap variance asymmetry, or more generally
   across independent synthetic and biological benchmark constructions.
4. **Separate benchmark creation from tool benchmarking.** The manuscript and
   archive should separate the code that creates a benchmark dataset from the
   code that benchmarks tools on that dataset. A benchmark should be written to
   disk as an explicit report plus annotation/truth files, and all modelling
   tools should then consume that fixed dataset. This makes the benchmark
   auditable and prevents method-specific analysis code from depending on the
   benchmark-generation code path. This is the structure used in the rebuilt
   sample-swap and protein-swap analyses here.
5. **Normalization and estimand.** The K562/CSF dilution benchmarks are also
   difficult settings for standard normalization comparisons. TIC, median,
   quantile, and VSN-style normalization can be appropriate in ordinary
   proteomics workflows, but in this benchmark they may change the estimand
   because the raw Neat versus 1:2 comparison is deliberately a total-material
   difference and the post-swap data are dominated by the manufactured Negative
   class. The manuscript should clarify whether comparator methods are being
   evaluated under their normal best-practice workflows or under constraints
   imposed by the benchmark construction.

## Additional suggestions for the next version

6. **Comparator workflows and benchmark venue.** Because the manuscript makes a
   comparative method-performance claim, the authors should consider engaging
   with developers or maintainers of comparator workflows such as `msqrob2`,
   `prolfqua`, and `limpa`, and with community benchmarking efforts such as
   ProteoBench. This would help distinguish MSstats+-specific gains from
   choices about benchmark construction, normalization, and method-specific
   workflow configuration.
7. **Runtime reporting.** The manuscript should report timing information for
   each tool and benchmark cell. In this reanalysis the observed runtimes differ
   substantially between tools and between datasets; without timing information,
   the reader cannot judge the practical cost of the reported performance.
8. **Memory reporting.** The manuscript should also report memory usage or peak
   memory ranges for each tool. Runtime alone is not sufficient for practical
   reproducibility, especially for methods that fit many protein-level models or
   require large intermediate objects.
9. **Include prolfqua.** The benchmark should include `prolfqua` as an
   additional contemporary modelling workflow. It is directly relevant to the
   comparison because it provides an independent implementation path for
   protein-level differential analysis and helps separate MSstats-specific
   behaviour from more general properties of moderated and non-moderated
   workflows.
