---
title: "Protein-swap pairs — before vs after the swap"
subtitle: "Five sample pairs from the CSF Spectronaut protein-swap benchmark"
author: "Witold Wolski"
date: today
format:
  html:
    self-contained: true
    toc: true
    code-fold: true
    fig-width: 12
    fig-height: 9
---

This page visualises five protein-swap pairs from the CSF Spectronaut
protein-swap benchmark before and after `src/swap_spectronaut_report.py`
was applied to the Spectronaut report. Each pair is the HIGH-abundance
member and its matched LOW-abundance partner (see
[`TODO/TODO_SWAP_DESIGN.md`](../TODO/TODO_SWAP_DESIGN.md)). Each panel
shows MSstats protein-level summaries (log2 intensities, one dot per
run), grouped by the G1 / G2 split that the swap script assigned within
each dilution. In the **Before** column intensities reflect the original
data; in the **After** column the HIGH partner's intensities in G2 runs
have been exchanged rank-for-rank with the LOW partner's intensities,
producing the synthetic differential-abundance signal at a target
|log2FC| ≈ 1.6.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(stringr)
})
theme_set(theme_bw(base_size = 12))
```

```{r load-inputs}
base_dir <- ".."
# Paths to the two MSstats protein-level summaries and the swap metadata.
rda_before <- file.path(base_dir,
  "CSF_Spectronaut/all_data/log2/swap/MSstats/MSstats_summarized.rda")
rda_after  <- file.path(base_dir,
  "CSF_Spectronaut_protein_swap/all_data/log2/swap/MSstats/MSstats_summarized.rda")
gt_path    <- file.path(base_dir,
  "CSF_Spectronaut_protein_swap/Report_swap_ground_truth.tsv")
grp_path   <- file.path(base_dir,
  "CSF_Spectronaut_protein_swap/Report_swap_group_annotation.csv")

stopifnot(file.exists(rda_before), file.exists(rda_after),
          file.exists(gt_path),    file.exists(grp_path))

# Pull ProteinLevelData out of each summary into its own local variable.
# The two folders' run_msstats.R scripts save the summary under different
# variable names — pre-swap is `base_msstats_summarized`, post-swap is
# `summarized`. Auto-detect either by scanning the freshly-loaded env.
load_pld <- function(rda) {
  e <- new.env()
  load(rda, envir = e)
  candidates <- intersect(c("base_msstats_summarized", "summarized"), ls(e))
  if (length(candidates) == 0L) {
    stop("No MSstats summary object found in ", rda)
  }
  obj <- e[[candidates[[1L]]]]
  as.data.table(obj$ProteinLevelData)
}
pld_before <- load_pld(rda_before)
pld_after  <- load_pld(rda_after)

gt    <- fread(gt_path,  sep = "\t")
grp   <- fread(grp_path)

# How many pairs the script produced overall (informational).
n_pairs_total <- nrow(gt)
```

The ground-truth file contains **`r n_pairs_total`** pairs. This page
plots the first five.

```{r pick-pairs}
n_show <- 5L
pairs5 <- gt[order(pair_id)][seq_len(min(.N, n_show))]
pairs5_long <- rbind(
  pairs5[, .(pair_id, PG = PG_high, role = "HIGH", partner = PG_low,
             expected_log2fc, n_precursors)],
  pairs5[, .(pair_id, PG = PG_low,  role = "LOW",  partner = PG_high,
             expected_log2fc, n_precursors)]
)
knitr::kable(
  pairs5[, .(pair_id, PG_high, PG_low, n_precursors,
             expected_log2fc = round(expected_log2fc, 3))],
  caption = "Pairs shown on this page (target |log2FC| ≈ 1.6 by design).")
```

```{r build-plot-data}
# Subset both protein-level tables to the 10 proteins of interest and
# attach the swap G1/G2 group to each run via originalRUN -> R.FileName.
make_panel <- function(pld, phase_label) {
  pld <- pld[Protein %in% pairs5_long$PG]
  pld <- merge(
    pld,
    grp[, .(R.FileName, group)],
    by.x = "originalRUN", by.y = "R.FileName",
    all.x = TRUE
  )
  pld <- merge(
    pld,
    pairs5_long[, .(PG, role, pair_id, partner)],
    by.x = "Protein", by.y = "PG",
    all.x = TRUE
  )
  pld[, phase := phase_label]
  pld[]
}
plot_dt <- rbind(
  make_panel(pld_before, "Before swap"),
  make_panel(pld_after,  "After swap")
)
plot_dt[, phase := factor(phase, levels = c("Before swap", "After swap"))]
plot_dt[, role  := factor(role,  levels = c("HIGH", "LOW"))]
plot_dt[, group := factor(group, levels = c("G1", "G2"))]
# Build the panel title (pair_id + the two protein IDs).
plot_dt <- merge(
  plot_dt,
  pairs5[, .(pair_id,
             facet = sprintf("pair %d (%s ↔ %s)", pair_id, PG_high, PG_low))],
  by = "pair_id"
)
plot_dt[, facet := factor(facet, levels = unique(facet[order(pair_id)]))]
```

```{r plot-pairs, fig.cap="Each panel = one swap pair. Columns split before-swap (left) and after-swap (right). Within a panel, x = the run's G1/G2 assignment, y = MSstats protein-level log2 intensity (one dot per run). Lines connect per-role group means."}
plot_means <- plot_dt[
  , .(LogIntensities = mean(LogIntensities, na.rm = TRUE)),
  by = .(facet, phase, group, role)
]

ggplot(plot_dt, aes(x = group, y = LogIntensities, colour = role)) +
  geom_jitter(width = 0.18, height = 0, alpha = 0.45, size = 1.8) +
  geom_point(data = plot_means, size = 3.6) +
  geom_line(data = plot_means, aes(group = role), linewidth = 1.0) +
  scale_colour_manual(values = c(HIGH = "#1f77b4", LOW = "#d62728")) +
  facet_grid(facet ~ phase, scales = "free_y") +
  labs(x = "Swap group (G1 / G2, per-dilution random split)",
       y = "log2 protein intensity (MSstats summary)",
       colour = NULL) +
  theme(legend.position = "top",
        strip.text.y    = element_text(size = 9, face = "bold"),
        strip.text.x    = element_text(size = 11, face = "bold"))
```

## How to read these panels

- **Before swap (left column).** Both pair members are real proteins
  with their real intensities. The G1 / G2 split is purely random within
  each dilution, so the HIGH and LOW lines should look roughly flat
  between groups (no systematic G1 vs G2 difference yet). The two
  proteins remain visibly different from each other: the HIGH partner
  sits well above the LOW partner.
- **After swap (right column).** In the G2 runs the HIGH partner's
  precursor / fragment intensities have been replaced rank-for-rank by
  the LOW partner's, and vice versa. So in G2 the **HIGH line should
  drop** to where the LOW line used to be, and the **LOW line should
  rise** to where the HIGH line used to be — the two lines cross. The
  G2-vs-G1 gap is the synthetic differential-abundance signal the
  benchmark recovers; by design its magnitude is ≈ 1.6 in log2 (table
  above).

If a panel does not show this crossing pattern in the right column, the
swap failed for that pair (e.g., dropped peptides removed too much
signal, or MSstats imputation absorbed it). All five pairs shown here
are early in the `pair_id` ordering, which is also the candidate-quality
ordering (smaller peptide-count mismatch + FC closer to the window
centre), so they should be among the cleanest.

## G2 only — precursor-level before vs after

The plot above summarises the swap at the protein level. To see what the
script actually did to the data, here is the precursor-level view
restricted to the **G2** runs (the half of each dilution whose
intensities were rewritten). Each dot is one precursor × one run,
coloured by which pair member it belongs to (HIGH blue, LOW red).
Reading left to right within a row you should see, after the swap, the
blue and red point clouds switch vertical position: the HIGH partner's
precursors drop down to the LOW partner's intensity band, and vice
versa.

Precursor intensities are aggregated by averaging MSstats-input
`Intensity` over fragments within each `(Run, Protein, PeptideSequence,
PrecursorCharge)` quadruple, then log2-transformed. Inputs come from the
two pipelines' `MSstats_input.csv` files; the same 5 pairs as above are
shown.

```{r g2-precursor-load}
mi_before_path <- file.path(base_dir,
  "CSF_Spectronaut/all_data/log2/swap/MSstats/MSstats_input.csv")
mi_after_path  <- file.path(base_dir,
  "CSF_Spectronaut_protein_swap/all_data/log2/swap/MSstats/MSstats_input.csv")
stopifnot(file.exists(mi_before_path), file.exists(mi_after_path))

target_proteins <- unique(pairs5_long$PG)
g2_runs <- grp[group == "G2", R.FileName]

# Only the columns we need; filter to the 10 proteins immediately after read.
keep_cols <- c("ProteinName", "PeptideSequence", "PrecursorCharge",
               "Run", "Intensity")
load_precursor <- function(path) {
  raw <- fread(path, select = keep_cols)
  raw <- raw[ProteinName %in% target_proteins & Run %in% g2_runs]
  raw[, .(intensity = mean(Intensity, na.rm = TRUE)),
       by = .(ProteinName, PeptideSequence, PrecursorCharge, Run)]
}
prec_before <- load_precursor(mi_before_path)
prec_after  <- load_precursor(mi_after_path)
prec_before[, phase := "Before swap"]
prec_after[,  phase := "After swap"]
prec_long <- rbind(prec_before, prec_after)
prec_long[, log2_intensity := log2(intensity)]
prec_long[, precursor_id := paste(PeptideSequence, PrecursorCharge, sep = "_")]

# Attach pair_id / role and the panel title used above.
prec_long <- merge(
  prec_long,
  pairs5_long[, .(PG, role, pair_id)],
  by.x = "ProteinName", by.y = "PG",
  all.x = TRUE
)
prec_long <- merge(
  prec_long,
  pairs5[, .(pair_id,
             facet = sprintf("pair %d (%s ↔ %s)", pair_id, PG_high, PG_low))],
  by = "pair_id"
)
prec_long[, phase := factor(phase, levels = c("Before swap", "After swap"))]
prec_long[, role  := factor(role,  levels = c("HIGH", "LOW"))]
prec_long[, facet := factor(facet, levels = unique(facet[order(pair_id)]))]

# Map each G2 run to a compact x-axis order using the Seq number in R.FileName
# (works for the CSF naming convention). Falls back to alphabetical order if
# no Seq token is present.
prec_long[, seq_int := suppressWarnings(as.integer(str_extract(Run, "(?<=Seq)\\d+")))]
prec_long[is.na(seq_int), seq_int := match(Run, sort(unique(Run)))]
prec_long[, sample := factor(Run, levels = unique(Run[order(seq_int)]))]
```

```{r g2-precursor-plot, fig.height=11, fig.cap="G2 runs only. Each dot is one precursor of one of the 10 pair members, in one G2 run. Columns: before-swap intensity (left), after-swap intensity (right). Colours: HIGH partner (blue), LOW partner (red). X-axis lists G2 R.FileNames ordered by injection sequence; labels suppressed for legibility."}
ggplot(prec_long,
       aes(x = sample, y = log2_intensity, colour = role)) +
  geom_point(position = position_jitter(width = 0.25, height = 0),
             alpha = 0.6, size = 1.6) +
  scale_colour_manual(values = c(HIGH = "#1f77b4", LOW = "#d62728")) +
  facet_grid(facet ~ phase, scales = "free_y") +
  labs(x = "G2 sample (R.FileName, ordered by injection sequence)",
       y = "log2 precursor intensity (mean over fragments)",
       colour = NULL) +
  theme(legend.position = "top",
        axis.text.x  = element_blank(),
        axis.ticks.x = element_blank(),
        strip.text.y = element_text(size = 9, face = "bold"),
        strip.text.x = element_text(size = 11, face = "bold"),
        panel.grid.minor.x = element_blank())
```

### Reading the G2-only panels

- **Before swap (left).** Each G2 sample shows the precursors of both
  pair members at their original intensities. The two clouds are
  vertically separated by ≈ 1.6 log2 (the target |log2FC|): blue (HIGH)
  on top, red (LOW) on the bottom.
- **After swap (right).** In every G2 sample, the script has replaced
  the HIGH partner's precursor intensities with the LOW partner's
  rank-aligned values, and vice versa. The two clouds swap vertical
  position — blue moves down, red moves up. Per-precursor scatter
  within a sample reflects the original LOW (resp. HIGH) precursor
  spread, since intensities were copied wholesale.

If you see a blue dot still sitting high on the right (or a red dot
still sitting low) in a given sample, that precursor's partner was
absent from that run and the swap had nothing to substitute in.

## Verification: each swapped precursor carries its partner's old value

What does it mean for the swap to "preserve intensities and only change
labels"? The right way to phrase it is: for each precursor in a G2 run,
the script writes that row's intensity columns to the **rank-aligned
partner precursor's pre-swap intensity** in the same run, so the
mapping is

`FG.Quantity[Run R, partner-of-P, post-swap] == FG.Quantity[Run R, P, pre-swap]`

whenever the partner's pre-swap value is non-null. (When the partner's
pre-swap value is NaN the script can't substitute it, so the original
stays — see the NaN edge case below.)

The plot below pulls `FG.Quantity` directly from the two Spectronaut
TSVs (pre- and post-swap), reconstructs the rank-aligned partner mapping
the same way [`src/swap_spectronaut_report.py`](../src/swap_spectronaut_report.py)
does (rank precursors within each protein by mean intensity over the
non-blank reference set, then pair by rank inside each protein pair),
and for every (G2 run × precursor) plots **the precursor's
post-swap value against its rank-aligned partner's pre-swap value**.
Points should sit exactly on `y = x`. Any deviation is either the
NaN-partner edge case (the original was kept) or a genuine bug.

```{r g2-verify-load, cache=TRUE}
# Read FG.Quantity from both Spectronaut TSVs. Filter immediately to the
# 10 proteins of interest to keep memory bounded. Cached because the
# input TSVs are multi-GB and rendering would otherwise be slow.
report_before_path <- file.path(base_dir, "CSF_Spectronaut/Report.tsv")
report_after_path  <- file.path(base_dir, "CSF_Spectronaut_protein_swap/Report.tsv")
report_cols <- c("R.FileName", "R.Condition",
                 "PG.ProteinGroups", "EG.PrecursorId", "FG.Quantity")
report_before <- fread(report_before_path, select = report_cols)
report_after  <- fread(report_after_path,  select = report_cols)
report_before <- unique(report_before[PG.ProteinGroups %in% target_proteins])
report_after  <- unique(report_after[ PG.ProteinGroups %in% target_proteins])
```

```{r g2-verify-build, fig.cap="Verification of the swap mechanism. For each precursor in each G2 run, x = its rank-aligned partner's pre-swap FG.Quantity, y = its own post-swap FG.Quantity. Cases where the partner's pre-swap value was NaN are not plotted (no x to plot against) and are counted separately in the subtitle.", fig.height=8, fig.width=11}
# Reference-set ranks (mirror the swap script: rank by mean FG.Quantity over
# non-blank R.Condition, descending, within each protein).
ref_means <- report_before[R.Condition != "blank" & !is.na(FG.Quantity),
                           .(prec_mean = mean(FG.Quantity)),
                           by = .(PG.ProteinGroups, EG.PrecursorId)]
ref_means[, ref_rank := rank(-prec_mean, ties.method = "first"),
           by = PG.ProteinGroups]

# Per-pair, per-rank: pair member precursors.
hi_ranks <- merge(
  pairs5[, .(pair_id, PG.ProteinGroups = PG_high)],
  ref_means, by = "PG.ProteinGroups", all.x = TRUE)[
    , .(pair_id, hi_prec = EG.PrecursorId, ref_rank)]
lo_ranks <- merge(
  pairs5[, .(pair_id, PG.ProteinGroups = PG_low)],
  ref_means, by = "PG.ProteinGroups", all.x = TRUE)[
    , .(pair_id, lo_prec = EG.PrecursorId, ref_rank)]
rank_pairs <- merge(hi_ranks, lo_ranks, by = c("pair_id", "ref_rank"))

# Per (Run, Precursor) pre- and post-swap FG.Quantity.
pre <- report_before[, .(R.FileName, EG.PrecursorId, fg_before = FG.Quantity)]
post <- report_after[,  .(R.FileName, EG.PrecursorId, fg_after  = FG.Quantity)]

# Materialise the two directional comparisons:
#   HIGH partner: self.after vs partner-LOW.before
#   LOW partner: self.after vs partner-HIGH.before
build_side <- function(side) {
  self_prec    <- if (side == "HIGH") "hi_prec" else "lo_prec"
  partner_prec <- if (side == "HIGH") "lo_prec" else "hi_prec"
  rp <- copy(rank_pairs)
  setnames(rp, c(self_prec, partner_prec), c("self_prec", "partner_prec"))
  g2_full <- CJ(R.FileName = g2_runs, rp_idx = seq_len(nrow(rp)))
  g2_full[, `:=`(pair_id = rp$pair_id[rp_idx],
                 ref_rank = rp$ref_rank[rp_idx],
                 self_prec = rp$self_prec[rp_idx],
                 partner_prec = rp$partner_prec[rp_idx])]
  g2_full[, rp_idx := NULL]
  g2_full <- merge(g2_full, pre,
                   by.x = c("R.FileName", "partner_prec"),
                   by.y = c("R.FileName", "EG.PrecursorId"), all.x = TRUE)
  setnames(g2_full, "fg_before", "partner_before")
  g2_full <- merge(g2_full, post,
                   by.x = c("R.FileName", "self_prec"),
                   by.y = c("R.FileName", "EG.PrecursorId"), all.x = TRUE)
  setnames(g2_full, "fg_after", "self_after")
  # Also the self pre-swap value, to identify the NaN-keep-own case.
  g2_full <- merge(g2_full, pre,
                   by.x = c("R.FileName", "self_prec"),
                   by.y = c("R.FileName", "EG.PrecursorId"), all.x = TRUE)
  setnames(g2_full, "fg_before", "self_before")
  g2_full[, side := side]
  g2_full[]
}
verify_dt <- rbind(build_side("HIGH"), build_side("LOW"))
verify_dt <- merge(verify_dt,
                   pairs5[, .(pair_id, facet = sprintf(
                     "pair %d (%s ↔ %s)", pair_id, PG_high, PG_low))],
                   by = "pair_id")
verify_dt[, facet := factor(facet, levels = unique(facet[order(pair_id)]))]
verify_dt[, side  := factor(side,  levels = c("HIGH", "LOW"))]

# Classify each row.
verify_dt[, status := fcase(
  is.na(self_after) & is.na(partner_before),
    "both NaN",
  is.na(partner_before) & !is.na(self_after) &
    !is.na(self_before) & abs(self_after - self_before) < 1e-9,
    "partner NaN — self.after = self.before (kept own)",
  is.na(partner_before),
    "partner NaN — other",
  !is.na(self_after) & !is.na(partner_before) &
    abs(self_after - partner_before) < 1e-6,
    "swap OK: self.after == partner.before",
  default = "MISMATCH"
)]
status_levels <- c(
  "swap OK: self.after == partner.before",
  "partner NaN — self.after = self.before (kept own)",
  "partner NaN — other",
  "both NaN",
  "MISMATCH"
)
verify_dt[, status := factor(status, levels = status_levels)]

# Tabulate and plot.
status_tbl <- verify_dt[, .N, by = status][order(match(status, status_levels))]
status_subtitle <- paste(sprintf("%s: %d", status_tbl$status, status_tbl$N),
                         collapse = " · ")
knitr::kable(status_tbl, col.names = c("Outcome per (G2 run × pair × side)", "n"),
             caption = "Per-row classification across all 5 pairs × 20 G2 runs × HIGH/LOW.")

plot_dt <- verify_dt[!is.na(partner_before) & !is.na(self_after)]
ggplot(plot_dt, aes(x = log2(partner_before), y = log2(self_after))) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey55") +
  geom_point(aes(colour = status), alpha = 0.55, size = 1.4) +
  facet_wrap(~ facet, ncol = 2, scales = "free") +
  scale_colour_manual(values = c(
    "swap OK: self.after == partner.before" = "#2ca02c",
    "MISMATCH" = "#d62728")) +
  labs(x = "log2 partner.before (rank-aligned partner's pre-swap FG.Quantity)",
       y = "log2 self.after (this precursor's post-swap FG.Quantity)",
       colour = NULL,
       subtitle = status_subtitle) +
  theme(legend.position = "top",
        strip.text = element_text(face = "bold"))
```

### What the panels confirm

- **Green dots on `y = x`** are the swap operating as documented: in
  this G2 run, the precursor's post-swap intensity equals its
  rank-aligned partner's pre-swap intensity. Every dot landing exactly
  on the line is a confirmation that no intensity value was created,
  destroyed, or rescaled — only relabelled.
- **Any red `MISMATCH` dot** would indicate a value that is not
  explained by the documented swap rule. There shouldn't be any.
- **NaN-partner cases** (counted in the table above, not plotted)
  happen when the rank-aligned partner has no measurement in that
  run. The script keeps the original value, so the precursor's
  post-swap value equals its pre-swap value rather than the partner's
  (which is NaN). This is the same "drop scope" issue captured in
  `TODO/TODO_SWAP_DESIGN.md` under "Known limitations" — it does not
  alter what the swap *did do* with the rows that have a partner.

## G2-only SD distribution: before vs after the swap

The swap only writes to G2 runs, so the cleanest test is to compute SD
**within the 7 G2 runs of `good_data`**: SD of mean log2 protein
intensity across the seven G2 samples, one value per protein. We do
this on the original (pre-swap) Report.tsv restricted to those seven
runs, and on the swap-script's `good_data/Report.tsv` (post-swap).

What the swap should do at this view:

- Negative proteins are never written to → their G2 SD distribution is
  **invariant**.
- Each Positive protein's G2 values are replaced rank-for-rank with
  its **partner Positive's** pre-swap G2 values. Aggregated across the
  Positive class, the multiset `{post_swap_G2_SD(p) : p ∈ Positives}`
  is therefore a **permutation** of `{pre_swap_G2_SD(p) : p ∈
  Positives}` — same numbers, different protein labels. So the
  Positive density should also be **invariant** under the swap.

The plot below tests both predictions empirically.

```{r sd-g2-load, cache=TRUE}
summarize_report <- function(report_path, run_set) {
  rep <- fread(report_path,
               select = c("R.FileName", "PG.ProteinGroups",
                          "EG.PrecursorId", "FG.Quantity"))
  rep <- rep[R.FileName %in% run_set &
             nzchar(PG.ProteinGroups) &
             is.finite(FG.Quantity) & FG.Quantity > 0]
  rep <- unique(rep, by = c("R.FileName", "PG.ProteinGroups",
                            "EG.PrecursorId"))
  rep[, .(LogIntensity = mean(log2(FG.Quantity), na.rm = TRUE)),
        by = .(R.FileName, Protein = PG.ProteinGroups)]
}

good_data_ann <- fread(file.path(base_dir,
  "CSF_Spectronaut_protein_swap/good_data/annotation.csv"))
swap_list <- fread(file.path(base_dir,
  "CSF_Spectronaut_protein_swap/CSF_protein_swap_list.csv"))
g2_run_set <- good_data_ann[Condition == "Condition2", R.FileName]
stopifnot(length(g2_run_set) >= 2)

pre_g2  <- summarize_report(file.path(base_dir,
  "CSF_Spectronaut/Report.tsv"), g2_run_set)
post_g2 <- summarize_report(file.path(base_dir,
  "CSF_Spectronaut_protein_swap/good_data/Report.tsv"), g2_run_set)
pre_g2[,  phase := "Before swap"]
post_g2[, phase := "After swap"]

# Per-protein SD across the 7 G2 runs, then join the swap-truth label.
sd_g2 <- function(pld) {
  m <- merge(pld, swap_list, by = "Protein")
  m[, .(sd_log = sd(LogIntensity, na.rm = TRUE), n = .N),
      by = .(Protein, Label, phase)
  ][is.finite(sd_log) & n >= 2]
}
sd_g2_all <- rbind(sd_g2(pre_g2), sd_g2(post_g2))
sd_g2_all[, phase := factor(phase, levels = c("Before swap", "After swap"))]
sd_g2_all[, Label := factor(Label, levels = c("Negative", "Positive"))]
```

```{r sd-g2-plot, fig.height=4.5, fig.width=11, fig.cap="G2-only within-group SD on the protein-swap good_data subset (the 7 G2 runs). Each density is one value per protein. **Left:** computed on the original Report.tsv. **Right:** computed on the swap-script's good_data/Report.tsv. Each density's median is marked by a dashed line."}
medians <- sd_g2_all[, .(med = median(sd_log)), by = .(phase, Label)]
ggplot(sd_g2_all, aes(x = sd_log, fill = Label, colour = Label)) +
  geom_density(alpha = 0.35, linewidth = 0.6) +
  geom_vline(data = medians, aes(xintercept = med, colour = Label),
             linetype = "dashed", linewidth = 0.5) +
  facet_wrap(~ phase, scales = "free_y", ncol = 2) +
  scale_fill_manual(values   = c(Negative = "#7f7f7f", Positive = "#2ca02c")) +
  scale_colour_manual(values = c(Negative = "#7f7f7f", Positive = "#2ca02c")) +
  labs(x = "SD of mean log2 protein intensity across the 7 G2 runs",
       y = "density", fill = NULL, colour = NULL) +
  theme(legend.position = "top",
        strip.text = element_text(face = "bold"))
```

```{r sd-g2-table}
sd_g2_summary <- sd_g2_all[
  ,
  .(n_proteins = uniqueN(Protein),
    median_sd  = round(median(sd_log), 3),
    iqr_sd     = round(IQR(sd_log), 3)),
  by = .(phase, Label)
]
knitr::kable(
  sd_g2_summary,
  caption = "G2-only SD summary (median + IQR) on good_data, by phase and label.",
  col.names = c("Phase", "Label", "n proteins (≥2 obs)", "median SD", "IQR")
)
```

### Why is the Positive baseline already higher than Negative?

The Negative densities **and** the Positive densities each match across
the swap, as expected. What stands out is that the **Positive baseline
is well above the Negative baseline already at pre-swap**, in a slice
of data the swap has never touched. The skepticism in the user's note
on this is correct — it cannot be an "abundance extremes" argument,
since the pair-builder draws candidates from the full abundance range.
The real driver is the `n_precursors` matching constraint:

```{r prot-class-bias}
ann_full <- good_data_ann$R.FileName
prot_chars <- summarize_report(file.path(base_dir,
  "CSF_Spectronaut/Report.tsv"), ann_full)
# n_precursors per protein observed in the good_data subset, plus mean log2
prot_chars2 <- merge(
  fread(file.path(base_dir, "CSF_Spectronaut/Report.tsv"),
        select = c("R.FileName","PG.ProteinGroups","EG.PrecursorId","FG.Quantity")
       )[R.FileName %in% ann_full & nzchar(PG.ProteinGroups) &
         is.finite(FG.Quantity) & FG.Quantity > 0,
         .(n_precursors = uniqueN(EG.PrecursorId)),
         by = .(Protein = PG.ProteinGroups)],
  prot_chars[, .(mean_log2 = mean(LogIntensity, na.rm = TRUE)),
              by = Protein],
  by = "Protein"
)
prot_chars2 <- merge(prot_chars2, swap_list, by = "Protein")
class_summary <- prot_chars2[, .(
  n_proteins         = .N,
  median_n_precursors = median(n_precursors),
  median_log2_abundance = round(median(mean_log2), 2)
), by = Label]
knitr::kable(class_summary,
             caption = "Class-level structure on good_data (15 neat runs). The pair-builder requires exact n_precursors match between HIGH and LOW partners; the greedy matcher consumes pairs in quality order, so the densest low-precursor bins fill up first and the Positive class is enriched 3× for low-precursor proteins. Low n_precursors directly inflates within-condition SD of the protein-level mean.")
```

The Positive median `n_precursors` is **3** versus **9** for the
Negative class. With only ~3 precursors averaged into a protein-level
intensity per run, the run-to-run noise of the mean is much larger
than with ~9, which is the dominant driver of the within-G2 SD shift
between classes. The ~1.25-log2 abundance gap between the medians is a
secondary correlation — a side effect of the same matching dynamics,
not a primary driver. Importantly, **the pair-builder does not filter
by abundance**, so the Positive abundance distribution overlaps the
Negative distribution heavily; the n_precursors constraint is what
shapes the selection.

### Implications for reading the review figures

- The "Positives have higher within-condition SD" pattern seen in
  `review.qmd`'s `protein-swap-sd` figure and
  `review_supplement.qmd`'s `protein-swap-effect-variance` figure is
  **mostly a precursor-count selection effect, not a swap-mechanism
  effect**. The same shift is visible in the pre-swap data.
- A clean follow-up to confirm: pick a Negative subsample matched to
  the Positive class on `n_precursors` and rerun the SD density on
  that matched Negative set. If the Negative-matched curve aligns with
  the Positive curve, the selection-bias hypothesis is confirmed.
- The swap itself does what it should: each Positive's G2 intensities
  are exchanged with a partner's, so the Positive-class SD multiset is
  permuted within the class but not changed in shape — exactly what
  the G2-only plot shows.
