---
title: "Protein-swap pairs — before vs after the swap"
subtitle: "Five sample pairs from the CSF Spectronaut protein-swap benchmark"
author: "Witold Wolski"
date: today
format:
  html:
    self-contained: true
    toc: true
    code-fold: true
    fig-width: 12
    fig-height: 9
---

This page visualises five protein-swap pairs from the CSF Spectronaut
protein-swap benchmark before and after `src/swap_spectronaut_report.py`
was applied to the Spectronaut report. Each pair is the HIGH-abundance
member and its matched LOW-abundance partner (see
[`TODO/TODO_SWAP_DESIGN.md`](../TODO/TODO_SWAP_DESIGN.md)). Each panel
shows MSstats protein-level summaries (log2 intensities, one dot per
run), grouped by the G1 / G2 split that the swap script assigned within
each dilution. In the **Before** column intensities reflect the original
data; in the **After** column the HIGH partner's intensities in G2 runs
have been exchanged rank-for-rank with the LOW partner's intensities,
producing the synthetic differential-abundance signal at a target
|log2FC| ≈ 1.6.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(stringr)
})
theme_set(theme_bw(base_size = 12))
```

```{r load-inputs}
base_dir <- ".."
# Paths to the two MSstats protein-level summaries and the swap metadata.
rda_before <- file.path(base_dir,
  "CSF_Spectronaut/all_data/log2/swap/MSstats/MSstats_summarized.rda")
rda_after  <- file.path(base_dir,
  "CSF_Spectronaut_protein_swap/all_data/log2/swap/MSstats/MSstats_summarized.rda")
gt_path    <- file.path(base_dir,
  "CSF_Spectronaut_protein_swap/Report_swap_ground_truth.tsv")
grp_path   <- file.path(base_dir,
  "CSF_Spectronaut_protein_swap/Report_swap_group_annotation.csv")

stopifnot(file.exists(rda_before), file.exists(rda_after),
          file.exists(gt_path),    file.exists(grp_path))

# Pull ProteinLevelData out of each summary into its own local variable.
# The two folders' run_msstats.R scripts save the summary under different
# variable names — pre-swap is `base_msstats_summarized`, post-swap is
# `summarized`. Auto-detect either by scanning the freshly-loaded env.
load_pld <- function(rda) {
  e <- new.env()
  load(rda, envir = e)
  candidates <- intersect(c("base_msstats_summarized", "summarized"), ls(e))
  if (length(candidates) == 0L) {
    stop("No MSstats summary object found in ", rda)
  }
  obj <- e[[candidates[[1L]]]]
  as.data.table(obj$ProteinLevelData)
}
pld_before <- load_pld(rda_before)
pld_after  <- load_pld(rda_after)

gt    <- fread(gt_path,  sep = "\t")
grp   <- fread(grp_path)

# How many pairs the script produced overall (informational).
n_pairs_total <- nrow(gt)
```

The ground-truth file contains **`r n_pairs_total`** pairs. This page
plots the first five.

```{r pick-pairs}
n_show <- 5L
pairs5 <- gt[order(pair_id)][seq_len(min(.N, n_show))]
pairs5_long <- rbind(
  pairs5[, .(pair_id, PG = PG_high, role = "HIGH", partner = PG_low,
             expected_log2fc, n_precursors)],
  pairs5[, .(pair_id, PG = PG_low,  role = "LOW",  partner = PG_high,
             expected_log2fc, n_precursors)]
)
knitr::kable(
  pairs5[, .(pair_id, PG_high, PG_low, n_precursors,
             expected_log2fc = round(expected_log2fc, 3))],
  caption = "Pairs shown on this page (target |log2FC| ≈ 1.6 by design).")
```

```{r build-plot-data}
# Subset both protein-level tables to the 10 proteins of interest and
# attach the swap G1/G2 group to each run via originalRUN -> R.FileName.
make_panel <- function(pld, phase_label) {
  pld <- pld[Protein %in% pairs5_long$PG]
  pld <- merge(
    pld,
    grp[, .(R.FileName, group)],
    by.x = "originalRUN", by.y = "R.FileName",
    all.x = TRUE
  )
  pld <- merge(
    pld,
    pairs5_long[, .(PG, role, pair_id, partner)],
    by.x = "Protein", by.y = "PG",
    all.x = TRUE
  )
  pld[, phase := phase_label]
  pld[]
}
plot_dt <- rbind(
  make_panel(pld_before, "Before swap"),
  make_panel(pld_after,  "After swap")
)
plot_dt[, phase := factor(phase, levels = c("Before swap", "After swap"))]
plot_dt[, role  := factor(role,  levels = c("HIGH", "LOW"))]
plot_dt[, group := factor(group, levels = c("G1", "G2"))]
# Build the panel title (pair_id + the two protein IDs).
plot_dt <- merge(
  plot_dt,
  pairs5[, .(pair_id,
             facet = sprintf("pair %d (%s ↔ %s)", pair_id, PG_high, PG_low))],
  by = "pair_id"
)
plot_dt[, facet := factor(facet, levels = unique(facet[order(pair_id)]))]
```

```{r plot-pairs, fig.cap="Each panel = one swap pair. Columns split before-swap (left) and after-swap (right). Within a panel, x = the run's G1/G2 assignment, y = MSstats protein-level log2 intensity (one dot per run). Lines connect per-role group means."}
plot_means <- plot_dt[
  , .(LogIntensities = mean(LogIntensities, na.rm = TRUE)),
  by = .(facet, phase, group, role)
]

ggplot(plot_dt, aes(x = group, y = LogIntensities, colour = role)) +
  geom_jitter(width = 0.18, height = 0, alpha = 0.45, size = 1.8) +
  geom_point(data = plot_means, size = 3.6) +
  geom_line(data = plot_means, aes(group = role), linewidth = 1.0) +
  scale_colour_manual(values = c(HIGH = "#1f77b4", LOW = "#d62728")) +
  facet_grid(facet ~ phase, scales = "free_y") +
  labs(x = "Swap group (G1 / G2, per-dilution random split)",
       y = "log2 protein intensity (MSstats summary)",
       colour = NULL) +
  theme(legend.position = "top",
        strip.text.y    = element_text(size = 9, face = "bold"),
        strip.text.x    = element_text(size = 11, face = "bold"))
```

## How to read these panels

- **Before swap (left column).** Both pair members are real proteins
  with their real intensities. The G1 / G2 split is purely random within
  each dilution, so the HIGH and LOW lines should look roughly flat
  between groups (no systematic G1 vs G2 difference yet). The two
  proteins remain visibly different from each other: the HIGH partner
  sits well above the LOW partner.
- **After swap (right column).** In the G2 runs the HIGH partner's
  precursor / fragment intensities have been replaced rank-for-rank by
  the LOW partner's, and vice versa. So in G2 the **HIGH line should
  drop** to where the LOW line used to be, and the **LOW line should
  rise** to where the HIGH line used to be — the two lines cross. The
  G2-vs-G1 gap is the synthetic differential-abundance signal the
  benchmark recovers; by design its magnitude is ≈ 1.6 in log2 (table
  above).

If a panel does not show this crossing pattern in the right column, the
swap failed for that pair (e.g., dropped peptides removed too much
signal, or MSstats imputation absorbed it). All five pairs shown here
are early in the `pair_id` ordering, which is also the candidate-quality
ordering (smaller peptide-count mismatch + FC closer to the window
centre), so they should be among the cleanest.

## G2 only — precursor-level before vs after

The plot above summarises the swap at the protein level. To see what the
script actually did to the data, here is the precursor-level view
restricted to the **G2** runs (the half of each dilution whose
intensities were rewritten). Each dot is one precursor × one run,
coloured by which pair member it belongs to (HIGH blue, LOW red).
Reading left to right within a row you should see, after the swap, the
blue and red point clouds switch vertical position: the HIGH partner's
precursors drop down to the LOW partner's intensity band, and vice
versa.

Precursor intensities are aggregated by averaging MSstats-input
`Intensity` over fragments within each `(Run, Protein, PeptideSequence,
PrecursorCharge)` quadruple, then log2-transformed. Inputs come from the
two pipelines' `MSstats_input.csv` files; the same 5 pairs as above are
shown.

```{r g2-precursor-load}
mi_before_path <- file.path(base_dir,
  "CSF_Spectronaut/all_data/log2/swap/MSstats/MSstats_input.csv")
mi_after_path  <- file.path(base_dir,
  "CSF_Spectronaut_protein_swap/all_data/log2/swap/MSstats/MSstats_input.csv")
stopifnot(file.exists(mi_before_path), file.exists(mi_after_path))

target_proteins <- unique(pairs5_long$PG)
g2_runs <- grp[group == "G2", R.FileName]

# Only the columns we need; filter to the 10 proteins immediately after read.
keep_cols <- c("ProteinName", "PeptideSequence", "PrecursorCharge",
               "Run", "Intensity")
load_precursor <- function(path) {
  raw <- fread(path, select = keep_cols)
  raw <- raw[ProteinName %in% target_proteins & Run %in% g2_runs]
  raw[, .(intensity = mean(Intensity, na.rm = TRUE)),
       by = .(ProteinName, PeptideSequence, PrecursorCharge, Run)]
}
prec_before <- load_precursor(mi_before_path)
prec_after  <- load_precursor(mi_after_path)
prec_before[, phase := "Before swap"]
prec_after[,  phase := "After swap"]
prec_long <- rbind(prec_before, prec_after)
prec_long[, log2_intensity := log2(intensity)]
prec_long[, precursor_id := paste(PeptideSequence, PrecursorCharge, sep = "_")]

# Attach pair_id / role and the panel title used above.
prec_long <- merge(
  prec_long,
  pairs5_long[, .(PG, role, pair_id)],
  by.x = "ProteinName", by.y = "PG",
  all.x = TRUE
)
prec_long <- merge(
  prec_long,
  pairs5[, .(pair_id,
             facet = sprintf("pair %d (%s ↔ %s)", pair_id, PG_high, PG_low))],
  by = "pair_id"
)
prec_long[, phase := factor(phase, levels = c("Before swap", "After swap"))]
prec_long[, role  := factor(role,  levels = c("HIGH", "LOW"))]
prec_long[, facet := factor(facet, levels = unique(facet[order(pair_id)]))]

# Map each G2 run to a compact x-axis order using the Seq number in R.FileName
# (works for the CSF naming convention). Falls back to alphabetical order if
# no Seq token is present.
prec_long[, seq_int := suppressWarnings(as.integer(str_extract(Run, "(?<=Seq)\\d+")))]
prec_long[is.na(seq_int), seq_int := match(Run, sort(unique(Run)))]
prec_long[, sample := factor(Run, levels = unique(Run[order(seq_int)]))]
```

```{r g2-precursor-plot, fig.height=11, fig.cap="G2 runs only. Each dot is one precursor of one of the 10 pair members, in one G2 run. Columns: before-swap intensity (left), after-swap intensity (right). Colours: HIGH partner (blue), LOW partner (red). X-axis lists G2 R.FileNames ordered by injection sequence; labels suppressed for legibility."}
ggplot(prec_long,
       aes(x = sample, y = log2_intensity, colour = role)) +
  geom_point(position = position_jitter(width = 0.25, height = 0),
             alpha = 0.6, size = 1.6) +
  scale_colour_manual(values = c(HIGH = "#1f77b4", LOW = "#d62728")) +
  facet_grid(facet ~ phase, scales = "free_y") +
  labs(x = "G2 sample (R.FileName, ordered by injection sequence)",
       y = "log2 precursor intensity (mean over fragments)",
       colour = NULL) +
  theme(legend.position = "top",
        axis.text.x  = element_blank(),
        axis.ticks.x = element_blank(),
        strip.text.y = element_text(size = 9, face = "bold"),
        strip.text.x = element_text(size = 11, face = "bold"),
        panel.grid.minor.x = element_blank())
```

### Reading the G2-only panels

- **Before swap (left).** Each G2 sample shows the precursors of both
  pair members at their original intensities. The two clouds are
  vertically separated by ≈ 1.6 log2 (the target |log2FC|): blue (HIGH)
  on top, red (LOW) on the bottom.
- **After swap (right).** In every G2 sample, the script has replaced
  the HIGH partner's precursor intensities with the LOW partner's
  rank-aligned values, and vice versa. The two clouds swap vertical
  position — blue moves down, red moves up. Per-precursor scatter
  within a sample reflects the original LOW (resp. HIGH) precursor
  spread, since intensities were copied wholesale.

If you see a blue dot still sitting high on the right (or a red dot
still sitting low) in a given sample, that precursor's partner was
absent from that run and the swap had nothing to substitute in.

## Verification: each swapped precursor carries its partner's old value

What does it mean for the swap to "preserve intensities and only change
labels"? The right way to phrase it is: for each precursor in a G2 run,
the script writes that row's intensity columns to the **rank-aligned
partner precursor's pre-swap intensity** in the same run, so the
mapping is

`FG.Quantity[Run R, partner-of-P, post-swap] == FG.Quantity[Run R, P, pre-swap]`

whenever the partner's pre-swap value is non-null. (When the partner's
pre-swap value is NaN the script can't substitute it, so the original
stays — see the NaN edge case below.)

The plot below pulls `FG.Quantity` directly from the two Spectronaut
TSVs (pre- and post-swap), reconstructs the rank-aligned partner mapping
the same way [`src/swap_spectronaut_report.py`](../src/swap_spectronaut_report.py)
does (rank precursors within each protein by mean intensity over the
non-blank reference set, then pair by rank inside each protein pair),
and for every (G2 run × precursor) plots **the precursor's
post-swap value against its rank-aligned partner's pre-swap value**.
Points should sit exactly on `y = x`. Any deviation is either the
NaN-partner edge case (the original was kept) or a genuine bug.

```{r g2-verify-load, cache=TRUE}
# Read FG.Quantity from both Spectronaut TSVs. Filter immediately to the
# 10 proteins of interest to keep memory bounded. Cached because the
# input TSVs are multi-GB and rendering would otherwise be slow.
report_before_path <- file.path(base_dir, "CSF_Spectronaut/Report.tsv")
report_after_path  <- file.path(base_dir, "CSF_Spectronaut_protein_swap/Report.tsv")
report_cols <- c("R.FileName", "R.Condition",
                 "PG.ProteinGroups", "EG.PrecursorId", "FG.Quantity")
report_before <- fread(report_before_path, select = report_cols)
report_after  <- fread(report_after_path,  select = report_cols)
report_before <- unique(report_before[PG.ProteinGroups %in% target_proteins])
report_after  <- unique(report_after[ PG.ProteinGroups %in% target_proteins])
```

```{r g2-verify-build, fig.cap="Verification of the swap mechanism. For each precursor in each G2 run, x = its rank-aligned partner's pre-swap FG.Quantity, y = its own post-swap FG.Quantity. Cases where the partner's pre-swap value was NaN are not plotted (no x to plot against) and are counted separately in the subtitle.", fig.height=8, fig.width=11}
# Reference-set ranks (mirror the swap script: rank by mean FG.Quantity over
# non-blank R.Condition, descending, within each protein).
ref_means <- report_before[R.Condition != "blank" & !is.na(FG.Quantity),
                           .(prec_mean = mean(FG.Quantity)),
                           by = .(PG.ProteinGroups, EG.PrecursorId)]
ref_means[, ref_rank := rank(-prec_mean, ties.method = "first"),
           by = PG.ProteinGroups]

# Per-pair, per-rank: pair member precursors.
hi_ranks <- merge(
  pairs5[, .(pair_id, PG.ProteinGroups = PG_high)],
  ref_means, by = "PG.ProteinGroups", all.x = TRUE)[
    , .(pair_id, hi_prec = EG.PrecursorId, ref_rank)]
lo_ranks <- merge(
  pairs5[, .(pair_id, PG.ProteinGroups = PG_low)],
  ref_means, by = "PG.ProteinGroups", all.x = TRUE)[
    , .(pair_id, lo_prec = EG.PrecursorId, ref_rank)]
rank_pairs <- merge(hi_ranks, lo_ranks, by = c("pair_id", "ref_rank"))

# Per (Run, Precursor) pre- and post-swap FG.Quantity.
pre <- report_before[, .(R.FileName, EG.PrecursorId, fg_before = FG.Quantity)]
post <- report_after[,  .(R.FileName, EG.PrecursorId, fg_after  = FG.Quantity)]

# Materialise the two directional comparisons:
#   HIGH partner: self.after vs partner-LOW.before
#   LOW partner: self.after vs partner-HIGH.before
build_side <- function(side) {
  self_prec    <- if (side == "HIGH") "hi_prec" else "lo_prec"
  partner_prec <- if (side == "HIGH") "lo_prec" else "hi_prec"
  rp <- copy(rank_pairs)
  setnames(rp, c(self_prec, partner_prec), c("self_prec", "partner_prec"))
  g2_full <- CJ(R.FileName = g2_runs, rp_idx = seq_len(nrow(rp)))
  g2_full[, `:=`(pair_id = rp$pair_id[rp_idx],
                 ref_rank = rp$ref_rank[rp_idx],
                 self_prec = rp$self_prec[rp_idx],
                 partner_prec = rp$partner_prec[rp_idx])]
  g2_full[, rp_idx := NULL]
  g2_full <- merge(g2_full, pre,
                   by.x = c("R.FileName", "partner_prec"),
                   by.y = c("R.FileName", "EG.PrecursorId"), all.x = TRUE)
  setnames(g2_full, "fg_before", "partner_before")
  g2_full <- merge(g2_full, post,
                   by.x = c("R.FileName", "self_prec"),
                   by.y = c("R.FileName", "EG.PrecursorId"), all.x = TRUE)
  setnames(g2_full, "fg_after", "self_after")
  # Also the self pre-swap value, to identify the NaN-keep-own case.
  g2_full <- merge(g2_full, pre,
                   by.x = c("R.FileName", "self_prec"),
                   by.y = c("R.FileName", "EG.PrecursorId"), all.x = TRUE)
  setnames(g2_full, "fg_before", "self_before")
  g2_full[, side := side]
  g2_full[]
}
verify_dt <- rbind(build_side("HIGH"), build_side("LOW"))
verify_dt <- merge(verify_dt,
                   pairs5[, .(pair_id, facet = sprintf(
                     "pair %d (%s ↔ %s)", pair_id, PG_high, PG_low))],
                   by = "pair_id")
verify_dt[, facet := factor(facet, levels = unique(facet[order(pair_id)]))]
verify_dt[, side  := factor(side,  levels = c("HIGH", "LOW"))]

# Classify each row.
verify_dt[, status := fcase(
  is.na(self_after) & is.na(partner_before),
    "both NaN",
  is.na(partner_before) & !is.na(self_after) &
    !is.na(self_before) & abs(self_after - self_before) < 1e-9,
    "partner NaN — self.after = self.before (kept own)",
  is.na(partner_before),
    "partner NaN — other",
  !is.na(self_after) & !is.na(partner_before) &
    abs(self_after - partner_before) < 1e-6,
    "swap OK: self.after == partner.before",
  default = "MISMATCH"
)]
status_levels <- c(
  "swap OK: self.after == partner.before",
  "partner NaN — self.after = self.before (kept own)",
  "partner NaN — other",
  "both NaN",
  "MISMATCH"
)
verify_dt[, status := factor(status, levels = status_levels)]

# Tabulate and plot.
status_tbl <- verify_dt[, .N, by = status][order(match(status, status_levels))]
status_subtitle <- paste(sprintf("%s: %d", status_tbl$status, status_tbl$N),
                         collapse = " · ")
knitr::kable(status_tbl, col.names = c("Outcome per (G2 run × pair × side)", "n"),
             caption = "Per-row classification across all 5 pairs × 20 G2 runs × HIGH/LOW.")

plot_dt <- verify_dt[!is.na(partner_before) & !is.na(self_after)]
ggplot(plot_dt, aes(x = log2(partner_before), y = log2(self_after))) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey55") +
  geom_point(aes(colour = status), alpha = 0.55, size = 1.4) +
  facet_wrap(~ facet, ncol = 2, scales = "free") +
  scale_colour_manual(values = c(
    "swap OK: self.after == partner.before" = "#2ca02c",
    "MISMATCH" = "#d62728")) +
  labs(x = "log2 partner.before (rank-aligned partner's pre-swap FG.Quantity)",
       y = "log2 self.after (this precursor's post-swap FG.Quantity)",
       colour = NULL,
       subtitle = status_subtitle) +
  theme(legend.position = "top",
        strip.text = element_text(face = "bold"))
```

### What the panels confirm

- **Green dots on `y = x`** are the swap operating as documented: in
  this G2 run, the precursor's post-swap intensity equals its
  rank-aligned partner's pre-swap intensity. Every dot landing exactly
  on the line is a confirmation that no intensity value was created,
  destroyed, or rescaled — only relabelled.
- **Any red `MISMATCH` dot** would indicate a value that is not
  explained by the documented swap rule. There shouldn't be any.
- **NaN-partner cases** (counted in the table above, not plotted)
  happen when the rank-aligned partner has no measurement in that
  run. The script keeps the original value, so the precursor's
  post-swap value equals its pre-swap value rather than the partner's
  (which is NaN). This is the same "drop scope" issue captured in
  `TODO/TODO_SWAP_DESIGN.md` under "Known limitations" — it does not
  alter what the swap *did do* with the rows that have a partner.

## Within-condition SD before vs after the swap

A natural sanity check: if the only thing the swap does is rewrite
intensities for the Positive class in G2 runs, then **before** the swap
there should be no systematic Positive/Negative within-condition SD
difference — at that point the labels are arbitrary relative to a
random G1/G2 partition. Whether this expectation holds is, in itself,
informative about the **swap-pair selection bias**: the swap candidates
are chosen from proteins with matched precursor counts and a fold-change
window across the abundance distribution, which can enrich the Positive
class for proteins with intrinsically different variability than the
average.

Below, two density plots over the protein-swap `good_data` subset
(15 neat runs: 8 in Condition1 ≡ G1, 7 in Condition2 ≡ G2). Each
protein contributes one SD per condition (per-condition SD, not pooled),
matching the convention in [`vignettes/review_supplement.qmd`'s
`protein-swap-effect-variance`
chunk](review_supplement.html) so the **right panel here is directly
comparable** to that figure. The **left** panel computes the same
quantity on the *original* (pre-swap) report restricted to the same
15 runs and the same Positive / Negative label assignment from
[`CSF_protein_swap_list.csv`](../CSF_Spectronaut_protein_swap/CSF_protein_swap_list.csv).

```{r sd-pre-post-load, cache=TRUE}
# Same protein/run summarization as review_supplement.qmd:
# log2(FG.Quantity), dedupe at (run, protein, precursor), then mean
# across precursors per (run, protein).
summarize_report <- function(report_path, run_set) {
  rep <- fread(report_path,
               select = c("R.FileName", "PG.ProteinGroups",
                          "EG.PrecursorId", "FG.Quantity"))
  rep <- rep[R.FileName %in% run_set &
             nzchar(PG.ProteinGroups) &
             is.finite(FG.Quantity) & FG.Quantity > 0]
  rep <- unique(rep, by = c("R.FileName", "PG.ProteinGroups",
                            "EG.PrecursorId"))
  rep[, .(LogIntensity = mean(log2(FG.Quantity), na.rm = TRUE)),
        by = .(R.FileName, Protein = PG.ProteinGroups)]
}

good_data_ann <- fread(file.path(base_dir,
  "CSF_Spectronaut_protein_swap/good_data/annotation.csv"))
swap_list <- fread(file.path(base_dir,
  "CSF_Spectronaut_protein_swap/CSF_protein_swap_list.csv"))
run_set <- good_data_ann$R.FileName
ann_g <- good_data_ann[, .(R.FileName, Condition)]

# Pre-swap = original Spectronaut Report restricted to the same 15
# good_data runs; G1/G2 condition comes from good_data_ann so the
# partition matches exactly.
pre_pld  <- summarize_report(file.path(base_dir,
  "CSF_Spectronaut/Report.tsv"), run_set)
# Post-swap = swap-script's good_data Report (15 runs, same set).
post_pld <- summarize_report(file.path(base_dir,
  "CSF_Spectronaut_protein_swap/good_data/Report.tsv"), run_set)
pre_pld[,  phase := "Before swap"]
post_pld[, phase := "After swap"]

# Join with condition + truth label, compute per-(protein, condition) SD.
per_cond_sd <- function(pld) {
  m <- merge(pld, ann_g, by = "R.FileName")
  m <- merge(m, swap_list, by = "Protein")
  m[, .(sd_log = sd(LogIntensity, na.rm = TRUE), n = .N),
      by = .(Protein, Label, Condition, phase)
  ][is.finite(sd_log) & n >= 2]
}
sd_pre  <- per_cond_sd(pre_pld)
sd_post <- per_cond_sd(post_pld)
sd_all  <- rbind(sd_pre, sd_post)
sd_all[, phase := factor(phase, levels = c("Before swap", "After swap"))]
sd_all[, Label := factor(Label, levels = c("Negative", "Positive"))]
```

```{r sd-pre-post-plot, fig.height=4.5, fig.width=11, fig.cap="Within-condition SD density on the protein-swap good_data subset (15 neat runs, 8 Condition1 + 7 Condition2). One value per protein per condition. **Left:** original Report.tsv values (pre-swap) using the same G1/G2 partition — Positive and Negative densities should overlap because the partition is random within neat runs and no swap has been applied yet. **Right:** swap-script's good_data Report (post-swap) — the Positive density shifts right because rank-aligned precursor exchanges between high- and low-abundance partners inflate the within-condition spread for the Positive class only."}
medians <- sd_all[, .(med = median(sd_log)), by = .(phase, Label)]
ggplot(sd_all, aes(x = sd_log, fill = Label, colour = Label)) +
  geom_density(alpha = 0.35, linewidth = 0.6) +
  geom_vline(data = medians, aes(xintercept = med, colour = Label),
             linetype = "dashed", linewidth = 0.5) +
  facet_wrap(~ phase, scales = "free_y", ncol = 2) +
  scale_fill_manual(values   = c(Negative = "#7f7f7f", Positive = "#2ca02c")) +
  scale_colour_manual(values = c(Negative = "#7f7f7f", Positive = "#2ca02c")) +
  labs(x = "within-condition SD of mean log2 protein intensity",
       y = "density", fill = NULL, colour = NULL) +
  theme(legend.position = "top",
        strip.text = element_text(face = "bold"))
```

```{r sd-pre-post-table}
sd_summary <- sd_all[
  ,
  .(n_proteins = uniqueN(Protein),
    n_obs      = .N,
    median_sd  = round(median(sd_log), 3),
    iqr_sd     = round(IQR(sd_log), 3)),
  by = .(phase, Label)
]
knitr::kable(
  sd_summary,
  caption = "Per-condition SD summary (median + IQR) on good_data, by phase and label.",
  col.names = c("Phase", "Label", "n proteins (≥2 obs)",
                "n (protein × condition) rows", "median SD", "IQR")
)
```

### What the panels actually show (and what that means)

What the data above reveals is *not* the clean "Negatives = Positives
before swap, Positives broader after swap" story I expected. The
medians in the table give the punchline:

- **Negatives are stable across the swap.** Median per-condition SD on
  the protein-swap `good_data` subset is **0.314** before and
  **0.314** after. Negatives are never written to, so this is just a
  cross-check.
- **Positives are broader than Negatives already *before* the swap.**
  Median per-condition SD ≈ **0.597**, roughly double the Negative
  value. After the swap it's **0.587** — essentially unchanged.

So the broader Positive SD distribution that shows up in
`review.qmd`'s `protein-swap-sd` figure (MSstats-summarized) and
`review_supplement.qmd`'s `protein-swap-effect-variance` figure (raw
report) is **predominantly a swap-pair selection bias**, not a
manufactured-by-the-swap effect. The pair-builder in
[`src/swap_spectronaut_report.py`](../src/swap_spectronaut_report.py)
keeps two structural constraints — exact `n_precursors` match and
`log2(prot_mean_hi) - log2(prot_mean_lo) ∈ [1.4, 1.8]` — that, after
greedy bipartite matching, end up picking the Positive class from a
slice of the abundance distribution that already has higher
within-condition variability than the bulk.

The swap mechanism itself **rearranges** SD values inside the Positive
class (each Positive's G2-condition SD becomes its rank-aligned
partner's pre-swap G2-condition SD) rather than inflating them. That
explains why the Positive median barely moves.

This caveat does not affect what the swap does to the **mean** signal —
the manufactured ~1.6 log2 effect that drives the benchmark is real
and lives in the mean, not the SD. But it does mean that any
interpretation of the existing review figures as "the rank-aligned
exchange inflates Positive variance" should be read as
"the Positives are *drawn from* a more variable subset, and the
exchange preserves that variability when relabelling".

A practical follow-up would be to construct an additional matched
Negative class with the same `n_precursors` distribution and a similar
abundance range as the Positives, to confirm that the Positive-class
SD distribution is reproduced by selection alone.
