---
title: "Protein-swap pairs — before vs after the swap"
subtitle: "Five sample pairs from the CSF Spectronaut protein-swap benchmark"
author: "Witold Wolski"
date: today
format:
  html:
    self-contained: true
    toc: true
    code-fold: true
    fig-width: 12
    fig-height: 9
---

This page visualises five protein-swap pairs from the CSF Spectronaut
protein-swap benchmark before and after `src/swap_spectronaut_report.py`
was applied to the Spectronaut report. Each pair is the HIGH-abundance
member and its matched LOW-abundance partner (see
[`TODO/TODO_SWAP_DESIGN.md`](../TODO/TODO_SWAP_DESIGN.md)). Each panel
shows MSstats protein-level summaries (log2 intensities, one dot per
run), grouped by the G1 / G2 split that the swap script assigned within
each dilution. In the **Before** column intensities reflect the original
data; in the **After** column the HIGH partner's intensities in G2 runs
have been exchanged rank-for-rank with the LOW partner's intensities,
producing the synthetic differential-abundance signal at a target
|log2FC| ≈ 1.6.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(stringr)
})
theme_set(theme_bw(base_size = 12))
```

```{r load-inputs}
base_dir <- ".."
# Paths to the two MSstats protein-level summaries and the swap metadata.
rda_before <- file.path(base_dir,
  "CSF_Spectronaut/all_data/log2/swap/MSstats/MSstats_summarized.rda")
rda_after  <- file.path(base_dir,
  "CSF_Spectronaut_protein_swap/all_data/log2/swap/MSstats/MSstats_summarized.rda")
gt_path    <- file.path(base_dir,
  "CSF_Spectronaut_protein_swap/Report_swap_ground_truth.tsv")
grp_path   <- file.path(base_dir,
  "CSF_Spectronaut_protein_swap/Report_swap_group_annotation.csv")

stopifnot(file.exists(rda_before), file.exists(rda_after),
          file.exists(gt_path),    file.exists(grp_path))

# Pull ProteinLevelData out of each summary into its own local variable.
# The two folders' run_msstats.R scripts save the summary under different
# variable names — pre-swap is `base_msstats_summarized`, post-swap is
# `summarized`. Auto-detect either by scanning the freshly-loaded env.
load_pld <- function(rda) {
  e <- new.env()
  load(rda, envir = e)
  candidates <- intersect(c("base_msstats_summarized", "summarized"), ls(e))
  if (length(candidates) == 0L) {
    stop("No MSstats summary object found in ", rda)
  }
  obj <- e[[candidates[[1L]]]]
  as.data.table(obj$ProteinLevelData)
}
pld_before <- load_pld(rda_before)
pld_after  <- load_pld(rda_after)

gt    <- fread(gt_path,  sep = "\t")
grp   <- fread(grp_path)

# How many pairs the script produced overall (informational).
n_pairs_total <- nrow(gt)
```

The ground-truth file contains **`r n_pairs_total`** pairs. This page
plots the first five.

```{r pick-pairs}
n_show <- 5L
pairs5 <- gt[order(pair_id)][seq_len(min(.N, n_show))]
pairs5_long <- rbind(
  pairs5[, .(pair_id, PG = PG_high, role = "HIGH", partner = PG_low,
             expected_log2fc, n_precursors)],
  pairs5[, .(pair_id, PG = PG_low,  role = "LOW",  partner = PG_high,
             expected_log2fc, n_precursors)]
)
knitr::kable(
  pairs5[, .(pair_id, PG_high, PG_low, n_precursors,
             expected_log2fc = round(expected_log2fc, 3))],
  caption = "Pairs shown on this page (target |log2FC| ≈ 1.6 by design).")
```

```{r build-plot-data}
# Subset both protein-level tables to the 10 proteins of interest and
# attach the swap G1/G2 group to each run via originalRUN -> R.FileName.
make_panel <- function(pld, phase_label) {
  pld <- pld[Protein %in% pairs5_long$PG]
  pld <- merge(
    pld,
    grp[, .(R.FileName, group)],
    by.x = "originalRUN", by.y = "R.FileName",
    all.x = TRUE
  )
  pld <- merge(
    pld,
    pairs5_long[, .(PG, role, pair_id, partner)],
    by.x = "Protein", by.y = "PG",
    all.x = TRUE
  )
  pld[, phase := phase_label]
  pld[]
}
plot_dt <- rbind(
  make_panel(pld_before, "Before swap"),
  make_panel(pld_after,  "After swap")
)
plot_dt[, phase := factor(phase, levels = c("Before swap", "After swap"))]
plot_dt[, role  := factor(role,  levels = c("HIGH", "LOW"))]
plot_dt[, group := factor(group, levels = c("G1", "G2"))]
# Build the panel title (pair_id + the two protein IDs).
plot_dt <- merge(
  plot_dt,
  pairs5[, .(pair_id,
             facet = sprintf("pair %d (%s ↔ %s)", pair_id, PG_high, PG_low))],
  by = "pair_id"
)
plot_dt[, facet := factor(facet, levels = unique(facet[order(pair_id)]))]
```

```{r plot-pairs, fig.cap="Each panel = one swap pair. Columns split before-swap (left) and after-swap (right). Within a panel, x = the run's G1/G2 assignment, y = MSstats protein-level log2 intensity (one dot per run). Lines connect per-role group means."}
plot_means <- plot_dt[
  , .(LogIntensities = mean(LogIntensities, na.rm = TRUE)),
  by = .(facet, phase, group, role)
]

ggplot(plot_dt, aes(x = group, y = LogIntensities, colour = role)) +
  geom_jitter(width = 0.18, height = 0, alpha = 0.45, size = 1.8) +
  geom_point(data = plot_means, size = 3.6) +
  geom_line(data = plot_means, aes(group = role), linewidth = 1.0) +
  scale_colour_manual(values = c(HIGH = "#1f77b4", LOW = "#d62728")) +
  facet_grid(facet ~ phase, scales = "free_y") +
  labs(x = "Swap group (G1 / G2, per-dilution random split)",
       y = "log2 protein intensity (MSstats summary)",
       colour = NULL) +
  theme(legend.position = "top",
        strip.text.y    = element_text(size = 9, face = "bold"),
        strip.text.x    = element_text(size = 11, face = "bold"))
```

## How to read these panels

- **Before swap (left column).** Both pair members are real proteins
  with their real intensities. The G1 / G2 split is purely random within
  each dilution, so the HIGH and LOW lines should look roughly flat
  between groups (no systematic G1 vs G2 difference yet). The two
  proteins remain visibly different from each other: the HIGH partner
  sits well above the LOW partner.
- **After swap (right column).** In the G2 runs the HIGH partner's
  precursor / fragment intensities have been replaced rank-for-rank by
  the LOW partner's, and vice versa. So in G2 the **HIGH line should
  drop** to where the LOW line used to be, and the **LOW line should
  rise** to where the HIGH line used to be — the two lines cross. The
  G2-vs-G1 gap is the synthetic differential-abundance signal the
  benchmark recovers; by design its magnitude is ≈ 1.6 in log2 (table
  above).

If a panel does not show this crossing pattern in the right column, the
swap failed for that pair (e.g., dropped peptides removed too much
signal, or MSstats imputation absorbed it). All five pairs shown here
are early in the `pair_id` ordering, which is also the candidate-quality
ordering (smaller peptide-count mismatch + FC closer to the window
centre), so they should be among the cleanest.

## G2 only — precursor-level before vs after

The plot above summarises the swap at the protein level. To see what the
script actually did to the data, here is the precursor-level view
restricted to the **G2** runs (the half of each dilution whose
intensities were rewritten). Each dot is one precursor × one run,
coloured by which pair member it belongs to (HIGH blue, LOW red).
Reading left to right within a row you should see, after the swap, the
blue and red point clouds switch vertical position: the HIGH partner's
precursors drop down to the LOW partner's intensity band, and vice
versa.

Precursor intensities are aggregated by averaging MSstats-input
`Intensity` over fragments within each `(Run, Protein, PeptideSequence,
PrecursorCharge)` quadruple, then log2-transformed. Inputs come from the
two pipelines' `MSstats_input.csv` files; the same 5 pairs as above are
shown.

```{r g2-precursor-load}
mi_before_path <- file.path(base_dir,
  "CSF_Spectronaut/all_data/log2/swap/MSstats/MSstats_input.csv")
mi_after_path  <- file.path(base_dir,
  "CSF_Spectronaut_protein_swap/all_data/log2/swap/MSstats/MSstats_input.csv")
stopifnot(file.exists(mi_before_path), file.exists(mi_after_path))

target_proteins <- unique(pairs5_long$PG)
g2_runs <- grp[group == "G2", R.FileName]

# Only the columns we need; filter to the 10 proteins immediately after read.
keep_cols <- c("ProteinName", "PeptideSequence", "PrecursorCharge",
               "Run", "Intensity")
load_precursor <- function(path) {
  raw <- fread(path, select = keep_cols)
  raw <- raw[ProteinName %in% target_proteins & Run %in% g2_runs]
  raw[, .(intensity = mean(Intensity, na.rm = TRUE)),
       by = .(ProteinName, PeptideSequence, PrecursorCharge, Run)]
}
prec_before <- load_precursor(mi_before_path)
prec_after  <- load_precursor(mi_after_path)
prec_before[, phase := "Before swap"]
prec_after[,  phase := "After swap"]
prec_long <- rbind(prec_before, prec_after)
prec_long[, log2_intensity := log2(intensity)]
prec_long[, precursor_id := paste(PeptideSequence, PrecursorCharge, sep = "_")]

# Attach pair_id / role and the panel title used above.
prec_long <- merge(
  prec_long,
  pairs5_long[, .(PG, role, pair_id)],
  by.x = "ProteinName", by.y = "PG",
  all.x = TRUE
)
prec_long <- merge(
  prec_long,
  pairs5[, .(pair_id,
             facet = sprintf("pair %d (%s ↔ %s)", pair_id, PG_high, PG_low))],
  by = "pair_id"
)
prec_long[, phase := factor(phase, levels = c("Before swap", "After swap"))]
prec_long[, role  := factor(role,  levels = c("HIGH", "LOW"))]
prec_long[, facet := factor(facet, levels = unique(facet[order(pair_id)]))]

# Map each G2 run to a compact x-axis order using the Seq number in R.FileName
# (works for the CSF naming convention). Falls back to alphabetical order if
# no Seq token is present.
prec_long[, seq_int := suppressWarnings(as.integer(str_extract(Run, "(?<=Seq)\\d+")))]
prec_long[is.na(seq_int), seq_int := match(Run, sort(unique(Run)))]
prec_long[, sample := factor(Run, levels = unique(Run[order(seq_int)]))]
```

```{r g2-precursor-plot, fig.height=11, fig.cap="G2 runs only. Each dot is one precursor of one of the 10 pair members, in one G2 run. Columns: before-swap intensity (left), after-swap intensity (right). Colours: HIGH partner (blue), LOW partner (red). X-axis lists G2 R.FileNames ordered by injection sequence; labels suppressed for legibility."}
ggplot(prec_long,
       aes(x = sample, y = log2_intensity, colour = role)) +
  geom_point(position = position_jitter(width = 0.25, height = 0),
             alpha = 0.6, size = 1.6) +
  scale_colour_manual(values = c(HIGH = "#1f77b4", LOW = "#d62728")) +
  facet_grid(facet ~ phase, scales = "free_y") +
  labs(x = "G2 sample (R.FileName, ordered by injection sequence)",
       y = "log2 precursor intensity (mean over fragments)",
       colour = NULL) +
  theme(legend.position = "top",
        axis.text.x  = element_blank(),
        axis.ticks.x = element_blank(),
        strip.text.y = element_text(size = 9, face = "bold"),
        strip.text.x = element_text(size = 11, face = "bold"),
        panel.grid.minor.x = element_blank())
```

### Reading the G2-only panels

- **Before swap (left).** Each G2 sample shows the precursors of both
  pair members at their original intensities. The two clouds are
  vertically separated by ≈ 1.6 log2 (the target |log2FC|): blue (HIGH)
  on top, red (LOW) on the bottom.
- **After swap (right).** In every G2 sample, the script has replaced
  the HIGH partner's precursor intensities with the LOW partner's
  rank-aligned values, and vice versa. The two clouds swap vertical
  position — blue moves down, red moves up. Per-precursor scatter
  within a sample reflects the original LOW (resp. HIGH) precursor
  spread, since intensities were copied wholesale.

If you see a blue dot still sitting high on the right (or a red dot
still sitting low) in a given sample, that precursor's partner was
absent from that run and the swap had nothing to substitute in.
