---
title: "Protein-swap pairs — before vs after the swap"
subtitle: "Five sample pairs from the CSF Spectronaut protein-swap benchmark"
author: "Witold Wolski"
date: today
format:
  html:
    self-contained: true
    toc: true
    code-fold: true
    fig-width: 12
    fig-height: 9
---

This page visualises five protein-swap pairs from the CSF Spectronaut
protein-swap benchmark before and after `src/swap_spectronaut_report.py`
was applied to the Spectronaut report. Each pair is the HIGH-abundance
member and its matched LOW-abundance partner (see
[`TODO/TODO_SWAP_DESIGN.md`](../TODO/TODO_SWAP_DESIGN.md)). Each panel
shows MSstats protein-level summaries (log2 intensities, one dot per
run), grouped by the G1 / G2 split that the swap script assigned within
each dilution. In the **Before** column intensities reflect the original
data; in the **After** column the HIGH partner's intensities in G2 runs
have been exchanged rank-for-rank with the LOW partner's intensities,
producing the synthetic differential-abundance signal at a target
|log2FC| ≈ 1.6.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(stringr)
})
theme_set(theme_bw(base_size = 12))
```

```{r load-inputs}
base_dir <- ".."
# Paths to the two MSstats protein-level summaries and the swap metadata.
rda_before <- file.path(base_dir,
  "CSF_Spectronaut/all_data/log2/swap/MSstats/MSstats_summarized.rda")
rda_after  <- file.path(base_dir,
  "CSF_Spectronaut_protein_swap/all_data/log2/swap/MSstats/MSstats_summarized.rda")
gt_path    <- file.path(base_dir,
  "CSF_Spectronaut_protein_swap/Report_swap_ground_truth.tsv")
grp_path   <- file.path(base_dir,
  "CSF_Spectronaut_protein_swap/Report_swap_group_annotation.csv")

stopifnot(file.exists(rda_before), file.exists(rda_after),
          file.exists(gt_path),    file.exists(grp_path))

# Pull ProteinLevelData out of each summary into its own local variable.
# The two folders' run_msstats.R scripts save the summary under different
# variable names — pre-swap is `base_msstats_summarized`, post-swap is
# `summarized`. Auto-detect either by scanning the freshly-loaded env.
load_pld <- function(rda) {
  e <- new.env()
  load(rda, envir = e)
  candidates <- intersect(c("base_msstats_summarized", "summarized"), ls(e))
  if (length(candidates) == 0L) {
    stop("No MSstats summary object found in ", rda)
  }
  obj <- e[[candidates[[1L]]]]
  as.data.table(obj$ProteinLevelData)
}
pld_before <- load_pld(rda_before)
pld_after  <- load_pld(rda_after)

gt    <- fread(gt_path,  sep = "\t")
grp   <- fread(grp_path)

# How many pairs the script produced overall (informational).
n_pairs_total <- nrow(gt)
```

The ground-truth file contains **`r n_pairs_total`** pairs. This page
plots the first five.

```{r pick-pairs}
n_show <- 5L
pairs5 <- gt[order(pair_id)][seq_len(min(.N, n_show))]
pairs5_long <- rbind(
  pairs5[, .(pair_id, PG = PG_high, role = "HIGH", partner = PG_low,
             expected_log2fc, n_precursors)],
  pairs5[, .(pair_id, PG = PG_low,  role = "LOW",  partner = PG_high,
             expected_log2fc, n_precursors)]
)
knitr::kable(
  pairs5[, .(pair_id, PG_high, PG_low, n_precursors,
             expected_log2fc = round(expected_log2fc, 3))],
  caption = "Pairs shown on this page (target |log2FC| ≈ 1.6 by design).")
```

```{r build-plot-data}
# Subset both protein-level tables to the 10 proteins of interest and
# attach the swap G1/G2 group to each run via originalRUN -> R.FileName.
make_panel <- function(pld, phase_label) {
  pld <- pld[Protein %in% pairs5_long$PG]
  pld <- merge(
    pld,
    grp[, .(R.FileName, group)],
    by.x = "originalRUN", by.y = "R.FileName",
    all.x = TRUE
  )
  pld <- merge(
    pld,
    pairs5_long[, .(PG, role, pair_id, partner)],
    by.x = "Protein", by.y = "PG",
    all.x = TRUE
  )
  pld[, phase := phase_label]
  pld[]
}
plot_dt <- rbind(
  make_panel(pld_before, "Before swap"),
  make_panel(pld_after,  "After swap")
)
plot_dt[, phase := factor(phase, levels = c("Before swap", "After swap"))]
plot_dt[, role  := factor(role,  levels = c("HIGH", "LOW"))]
plot_dt[, group := factor(group, levels = c("G1", "G2"))]
# Build the panel title (pair_id + the two protein IDs).
plot_dt <- merge(
  plot_dt,
  pairs5[, .(pair_id,
             facet = sprintf("pair %d (%s ↔ %s)", pair_id, PG_high, PG_low))],
  by = "pair_id"
)
plot_dt[, facet := factor(facet, levels = unique(facet[order(pair_id)]))]
```

```{r plot-pairs, fig.cap="Each panel = one swap pair. Columns split before-swap (left) and after-swap (right). Within a panel, x = the run's G1/G2 assignment, y = MSstats protein-level log2 intensity (one dot per run). Lines connect per-role group means."}
plot_means <- plot_dt[
  , .(LogIntensities = mean(LogIntensities, na.rm = TRUE)),
  by = .(facet, phase, group, role)
]

ggplot(plot_dt, aes(x = group, y = LogIntensities, colour = role)) +
  geom_jitter(width = 0.18, height = 0, alpha = 0.45, size = 1.8) +
  geom_point(data = plot_means, size = 3.6) +
  geom_line(data = plot_means, aes(group = role), linewidth = 1.0) +
  scale_colour_manual(values = c(HIGH = "#1f77b4", LOW = "#d62728")) +
  facet_grid(facet ~ phase, scales = "free_y") +
  labs(x = "Swap group (G1 / G2, per-dilution random split)",
       y = "log2 protein intensity (MSstats summary)",
       colour = NULL) +
  theme(legend.position = "top",
        strip.text.y    = element_text(size = 9, face = "bold"),
        strip.text.x    = element_text(size = 11, face = "bold"))
```

## How to read these panels

- **Before swap (left column).** Both pair members are real proteins
  with their real intensities. The G1 / G2 split is purely random within
  each dilution, so the HIGH and LOW lines should look roughly flat
  between groups (no systematic G1 vs G2 difference yet). The two
  proteins remain visibly different from each other: the HIGH partner
  sits well above the LOW partner.
- **After swap (right column).** In the G2 runs the HIGH partner's
  precursor / fragment intensities have been replaced rank-for-rank by
  the LOW partner's, and vice versa. So in G2 the **HIGH line should
  drop** to where the LOW line used to be, and the **LOW line should
  rise** to where the HIGH line used to be — the two lines cross. The
  G2-vs-G1 gap is the synthetic differential-abundance signal the
  benchmark recovers; by design its magnitude is ≈ 1.6 in log2 (table
  above).

If a panel does not show this crossing pattern in the right column, the
swap failed for that pair (e.g., dropped peptides removed too much
signal, or MSstats imputation absorbed it). All five pairs shown here
are early in the `pair_id` ordering, which is also the candidate-quality
ordering (smaller peptide-count mismatch + FC closer to the window
centre), so they should be among the cleanest.
