---
title: "Mix of Proteome benchmark — design visualization"
author: "Generated to accompany Mix_of_Proteome_design.md"
format:
  html:
    toc: true
    code-fold: true
    fig-width: 9
    fig-height: 5
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

This notebook visualises the three-organism Mix_of_Proteome benchmark
(E. coli + Human + Yeast at known proportions). Read alongside
[`Mix_of_Proteome_design.md`](Mix_of_Proteome_design.md). Unlike the CSF /
K562 dilution-series benchmarks, this dataset has a built-in null class
(Human, 50 % in every condition) and a built-in positive class (E. coli
and Yeast, varied) — no swap is needed.

```{r setup}
library(data.table)
library(dplyr)
library(ggplot2)
library(stringr)
library(tidyr)
library(patchwork)

theme_set(theme_minimal(base_size = 11))

annotation = fread("Mix_of_Proteome_annotation.csv")
idmap      = fread("idmapping.tsv")

# Load (or recompute) summarized protein-level data
sum_path = "MSstats/msstats_summarized.rda"
if (!file.exists(sum_path)) {
  library(MSstats)
  msstats_input = fread("MSstats/msstats_input.csv") |> as.data.frame()
  msstats_input$AnomalyScores = NA
  msstats_summarized = dataProcess(msstats_input,
                                   normalization  = FALSE,
                                   featureSubset  = "topN",
                                   n_top_feature  = 100,
                                   MBimpute       = TRUE,
                                   summaryMethod  = "TMP",
                                   numberOfCores  = 4)
  save(msstats_summarized, file = sum_path)
} else {
  load(sum_path)
}

prot = as.data.table(msstats_summarized$ProteinLevelData)
prot[, Protein := as.character(Protein)]
prot[, GROUP   := as.character(GROUP)]

# Attach Organism (3-class label: Human / E. coli / Yeast / NA)
org_lookup = idmap[, .(Protein = Entry,
                        Organism = fcase(
                          grepl("Homo sapiens",      Organism), "Human",
                          grepl("Escherichia coli",  Organism), "E. coli",
                          grepl("Saccharomyces|yeast", Organism, ignore.case = TRUE), "Yeast",
                          default = NA_character_))]
prot = merge(prot, org_lookup, by = "Protein", all.x = TRUE)
```

# 1. Acquisition schedule

Each row is one MS run. `Order` is injection order; `Condition` codes the
mixing ratio (E.coli / Human / Yeast in %). Human is held at 50 % in every
run; only E. coli (5 → 45 %) and Yeast (45 → 5 %) vary.

```{r schedule, fig.height=4, fig.width=12}
ann_plot = copy(annotation)
ann_plot[, Ecoli := as.integer(str_extract(Condition, "(?<=E)[0-9]+"))]
ann_plot[, Yeast := as.integer(str_extract(Condition, "(?<=Y)[0-9]+"))]

cond_levels = ann_plot[order(Ecoli), unique(Condition)]
ann_plot[, Condition := factor(Condition, levels = cond_levels)]

ggplot(ann_plot, aes(x = Order, y = Condition, fill = Ecoli)) +
  geom_point(shape = 21, size = 6, colour = "white", stroke = 0.6) +
  geom_text(aes(label = BioReplicate), size = 2.6, colour = "grey20") +
  scale_fill_gradient(low = "#fee5d9", high = "#a50f15",
                      name = "E. coli %", limits = c(0, 50)) +
  scale_x_continuous(breaks = 1:18, expand = expansion(add = 0.5)) +
  labs(title = "Run schedule (Order × Condition)",
       subtitle = "Number inside each marker = BioReplicate. Conditions ordered by E. coli %.",
       x = "Injection order", y = NULL) +
  theme(panel.grid.major.y = element_line(colour = "grey90"),
        panel.grid.minor.x = element_blank(),
        axis.text.y = element_text(family = "mono"),
        legend.position = "right")
```

The 18 runs are not blocked by condition — they are interleaved in
injection order, which spreads any temporal drift across all six
conditions.

# 2. Per-protein abundance heatmap by organism

Rows are proteins (sampled to 250 per organism for legibility), columns are
runs sorted by condition (E5 → E45). Each row is z-scored. Top facet =
Human (expected flat across columns), middle = E. coli (expected
left-to-right gradient), bottom = Yeast (expected right-to-left gradient).

```{r heatmap, fig.height=11, fig.width=12}
run_order = ann_plot[order(Condition, Order), .(originalRUN = R.FileName, Condition)]

prot_h = prot[!is.na(Organism) & originalRUN %in% run_order$originalRUN]

# z-score per protein
prot_h[, z := scale(LogIntensities)[, 1], by = Protein]
prot_h[, z := pmax(pmin(z, 2), -2)]

# Sample 250 proteins per organism
set.seed(42)
keep_proteins = prot_h[, .(P = unique(Protein)), by = Organism][
  , .(P = sample(P, min(.N, 250))), by = Organism]
prot_h = prot_h[Protein %in% keep_proteins$P]

# Cluster rows within each organism
make_row_order = function(dat) {
  m   = dcast(dat, Protein ~ originalRUN, value.var = "z")
  pids = m$Protein
  mat = as.matrix(m[, -1, with = FALSE]); mat[is.na(mat)] = 0
  pids[if (nrow(mat) > 1) hclust(dist(mat))$order else 1]
}
ord_by_org = prot_h[, .(P = list(make_row_order(.SD))), by = Organism]
prot_h[, Protein := factor(Protein,
                            levels = unlist(ord_by_org$P))]
prot_h[, originalRUN := factor(originalRUN, levels = run_order$originalRUN)]
prot_h[, Organism := factor(Organism, levels = c("Human", "E. coli", "Yeast"))]

col_strip = run_order[, .(originalRUN = factor(originalRUN, levels = run_order$originalRUN), Condition)]
p_strip = ggplot(col_strip, aes(x = originalRUN, y = 1, fill = Condition)) +
  geom_tile() +
  scale_fill_brewer(palette = "Spectral", name = "Condition") +
  scale_x_discrete(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0)) +
  theme_void() +
  theme(legend.position = "top", plot.margin = margin(0, 0, 0, 0))

p_heat = ggplot(prot_h, aes(x = originalRUN, y = Protein, fill = z)) +
  geom_tile() +
  facet_grid(Organism ~ ., scales = "free_y", space = "free_y", switch = "y") +
  scale_fill_gradient2(low = "#2ca02c", mid = "grey90", high = "#d62728",
                       midpoint = 0, limits = c(-2, 2),
                       name = "row z-score") +
  scale_x_discrete(expand = c(0, 0)) +
  labs(title = "Per-protein intensity by organism",
       subtitle = "Human ~ flat. E. coli increases left→right (E5→E45). Yeast decreases left→right.",
       x = NULL, y = NULL) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        strip.text.y.left = element_text(angle = 0, face = "bold"),
        legend.position = "right",
        panel.spacing.y = unit(0.4, "lines"))

p_strip / p_heat + plot_layout(heights = c(0.04, 1), guides = "collect") &
  theme(legend.position = "right")
```

**What to see:**

- **Human (top).** No left/right gradient — rows are scrambled red/green
  noise. This is the null class.
- **E. coli (middle).** Clear gradient: green on the left (E5, E10) → red
  on the right (E40, E45).
- **Yeast (bottom).** Mirror of E. coli: red on the left, green on the
  right.

# 3. Per-protein effect size per pairwise comparison

For each of the 15 condition pairs, plot the density of
`mean_log2(c1) − mean_log2(c2)` split by organism. Vertical dashed lines
mark the expected log2FC for each organism in that contrast.

```{r effect_density, fig.height=10, fig.width=12}
# Pairwise contrasts (matches the 15 in Mixture_of_proteomes_processing.R)
cond_pct = unique(ann_plot[, .(Condition,
                                 E = Ecoli, H = 50, Y = Yeast)])
pairs_dt = CJ(c1 = cond_pct$Condition, c2 = cond_pct$Condition)[
  as.character(c1) < as.character(c2)]

pairs_dt = merge(pairs_dt, cond_pct[, .(Condition, E1 = E, Y1 = Y)],
                 by.x = "c1", by.y = "Condition")
pairs_dt = merge(pairs_dt, cond_pct[, .(Condition, E2 = E, Y2 = Y)],
                 by.x = "c2", by.y = "Condition")
pairs_dt[, `:=`(exp_Human = 0,
                 `exp_E. coli` = log2(E1 / E2),
                 exp_Yeast    = log2(Y1 / Y2))]
pairs_dt[, Contrast := paste(c1, "vs", c2)]

per_cond = prot[!is.na(GROUP), .(mean_log = mean(LogIntensities, na.rm = TRUE)),
                 by = .(Protein, GROUP)]
per_cond_w = dcast(per_cond, Protein ~ GROUP, value.var = "mean_log")

eff_long = pairs_dt[, {
  c1n = as.character(c1); c2n = as.character(c2)
  diff_vec = per_cond_w[[c1n]] - per_cond_w[[c2n]]
  data.table(Protein = per_cond_w$Protein, diff = diff_vec)
}, by = .(Contrast, c1, c2,
           exp_Human, `exp_E. coli`, exp_Yeast)]
eff_long = merge(eff_long, org_lookup, by = "Protein")
eff_long = eff_long[!is.na(Organism) & is.finite(diff)]

# Long format for expected lines
exp_long = melt(unique(pairs_dt[, .(Contrast,
                                      Human   = exp_Human,
                                      `E. coli` = `exp_E. coli`,
                                      Yeast   = exp_Yeast)]),
                id.vars = "Contrast",
                variable.name = "Organism", value.name = "expected")

ggplot(eff_long, aes(x = diff, fill = Organism)) +
  geom_density(alpha = 0.45, colour = NA) +
  geom_vline(data = exp_long,
             aes(xintercept = expected, colour = Organism),
             linetype = 2, linewidth = 0.5) +
  facet_wrap(~ Contrast, ncol = 3, scales = "free_y") +
  scale_fill_manual(values = c("Human" = "#7f7f7f",
                                "E. coli" = "#1f77b4",
                                "Yeast" = "#2ca02c")) +
  scale_colour_manual(values = c("Human" = "#7f7f7f",
                                  "E. coli" = "#1f77b4",
                                  "Yeast" = "#2ca02c"),
                      guide = "none") +
  coord_cartesian(xlim = c(-4.5, 4.5)) +
  labs(title = "Per-protein log2 fold-change by contrast and organism",
       subtitle = "Dashed lines = expected log2FC from the mixing design.",
       x = "log2(c1) − log2(c2)", y = "density", fill = NULL) +
  theme(legend.position = "top",
        strip.text = element_text(size = 8))
```

Human density is centred on 0 in every panel (it is the null class). E. coli
and Yeast densities are centred on the expected value from the mixing
design, with the symmetric inversion between organisms visible across
panels (where E. coli is shifted left, Yeast is shifted right).

# 4. Per-protein within-condition spread by organism

Pooled within-condition SD per protein, by organism. Unlike CSF/K562, no
operation has been applied to one organism but not the others — so the
distributions should largely overlap.

```{r sd_density, fig.height=4}
per_run_stats = prot[!is.na(GROUP) & !is.na(Organism),
                      .(mean_log = mean(LogIntensities, na.rm = TRUE),
                        sd_log   = sd(LogIntensities,   na.rm = TRUE),
                        n        = .N),
                      by = .(Protein, GROUP, Organism)]

# Pool over conditions: mean of within-condition SDs per protein, weighted by df
pooled = per_run_stats[, .(sd_pool = sqrt(sum((n - 1) * sd_log^2, na.rm = TRUE) /
                                            max(sum(n - 1, na.rm = TRUE), 1)),
                           mean_intensity = mean(mean_log, na.rm = TRUE)),
                        by = .(Protein, Organism)]
pooled[, Organism := factor(Organism, levels = c("Human", "E. coli", "Yeast"))]

ggplot(pooled[is.finite(sd_pool) & sd_pool > 0],
        aes(x = sd_pool, fill = Organism)) +
  geom_density(alpha = 0.45, colour = NA) +
  scale_fill_manual(values = c("Human" = "#7f7f7f",
                                "E. coli" = "#1f77b4",
                                "Yeast" = "#2ca02c")) +
  coord_cartesian(xlim = c(0, 1.5)) +
  labs(title = "Within-condition pooled SD per protein, by organism",
       x = "sd(log2 intensity) inside each condition (pooled)",
       y = "density", fill = NULL) +
  theme(legend.position = "top")
```

The three densities sit on top of each other — the within-condition spread
is governed by the measurement noise floor, not by which organism a peptide
came from. This is why empirical-Bayes variance moderation is *not*
penalised on this benchmark (contrast with CSF/K562 §5).

# 5. Mean–variance trend by organism

```{r mean_var_trend, fig.height=4}
ggplot(pooled[is.finite(sd_pool) & sd_pool > 0],
        aes(x = mean_intensity, y = sd_pool, colour = Organism)) +
  geom_point(alpha = 0.35, size = 0.7) +
  geom_smooth(se = FALSE, method = "loess") +
  scale_colour_manual(values = c("Human" = "#7f7f7f",
                                  "E. coli" = "#1f77b4",
                                  "Yeast" = "#2ca02c")) +
  coord_cartesian(ylim = c(0, 1.5)) +
  labs(title = "Mean–variance relationship by organism",
       subtitle = "Trends overlap — no class-asymmetric variance bias.",
       x = "mean log2 intensity", y = "pooled SD", colour = NULL) +
  theme(legend.position = "top")
```

The three loess curves trace approximately the same downward trend (high
SD at low intensity from drop-out / LOQ, low SD at high intensity), with
no class-wide vertical separation. Empirical-Bayes moderation sees a
single, well-defined mean-variance trend and stabilises variance estimates
without biasing one organism relative to another.

# 6. Take-aways

1. **Ground truth is built into the design**: Human = null (50 % in every
   condition), E. coli + Yeast = positives with known per-comparison
   log2FC. No swap, no manufactured null.
2. **Effect sizes range from 0.17 to 3.17 log2 units** across the 15
   pairwise comparisons. Adjacent contrasts (e.g. E10 vs E20) are hard;
   extreme contrasts (E5 vs E45) are easy.
3. **Within-condition variance is approximately equal across organisms** —
   the structural disadvantage that variance moderation suffers on
   CSF/K562 does not apply here. Differences between methods on this
   dataset reflect the methods, not the design.
